### 2019-10-15 prisc-2020/acmart-sigplanproc/acmart-sigplanproc-template.tex
prisc 2020

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan]{acmart}\settopmatter{}

%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
\acmYear{2018}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from traditional SIGPLAN
%% proceedings format to PACMPL format must update the
%% '\documentclass' and topmatter commands above; see
%% 'acmart-pacmpl-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
%\usepackage[showframe]{geometry}% http://ctan.org/pkg/geometry
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{graphicx} % need this package
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage{dsfont}
\usepackage{float}
\usepackage{IEEEtrantools}

\newcommand{\sep}{-\kern-.6em\raisebox{-.659ex}{*}\ }
\newcommand{\bupd}[1]{=\kern-.6em\{#1\}\ \kern-.9em =\ \kern-.8em\raisebox{-.659ex}{*}\ }

\newcommand{\bigsep}{\mathop{\scalebox{2.5}{\raisebox{-0.4ex}{$*$}}}}%

\newcommand{\ra}[1] {
  \begin{tikzpicture}
        \node[draw,dashed] {#1};
    \end{tikzpicture}}

\newcommand{\inv}[1] {
  \begin{tikzpicture}
        \node[draw] {#1};
    \end{tikzpicture}}

\newcommand{\interp}[2]{(#1)(#2)}
\newcommand\sepimp{\mathrel{-\mkern-6mu*}}

\begin{document}

%% Title information
\title[Short Title]{Full Title}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Subtitle}                     %% \subtitle is optional
\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{First1 Last1}
\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}                    %% \country is recommended
}
\email{first1.last1@inst1.edu}          %% \email is recommended

%% Author with two affiliations and emails.
\author{First2 Last2}
\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2a}
  \department{Department2a}             %% \department is recommended
  \institution{Institution2a}           %% \institution is required
  \streetaddress{Street2a Address2a}
  \city{City2a}
  \state{State2a}
  \postcode{Post-Code2a}
  \country{Country2a}                   %% \country is recommended
}
\email{first2.last2@inst2a.com}         %% \email is recommended
\affiliation{
  \position{Position2b}
  \department{Department2b}             %% \department is recommended
  \institution{Institution2b}           %% \institution is required
  \streetaddress{Street3b Address2b}
  \city{City2b}
  \state{State2b}
  \postcode{Post-Code2b}
  \country{Country2b}                   %% \country is recommended
}
\email{first2.last2@inst2b.org}         %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Text of abstract \ldots.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

Text of paper \ldots

\section{Iris Primer}
??

\section{Iris Embedding of a Capability Machine}
The setting: we want to show full abstraction from a high level language, to a low level capability machine. Some basic properties of a high level language: local state encapsulation of well bracketed control flow. These are not natively enforced by a capability machine. When calling an adversary, any guarantees will depend on the calling convention the compiler implements. For LSE, the calling convention should ensure that no passed capability give authority over local state. For WBCR, the situation is more subtle. Skorstensgaard et. al. give detailed explanations of multiple control flow attacks that would happen with a naive calling conventions.

With these goals in mind, it is important not to \textit{accidentally} formalize a language where certain control flow abstractions are enforced. We also want to be as architecture independent as possible. This we are interested in allowing: arbitrary jumps, and (almost) arbitrary capability manipulations (within the given range of authority).

Iris is a higher-order concurrent separation logic framework. Its support of higher-order ghost state makes Iris a great candidate for complex reasoning about \textit{state}. This is exactly what we need for reasoning about a machine, where computation \textit{is} state manipulation.

Machine instructions operate through memory. Typically a machine will have a special register; the program counter, which contains a pointer to an address in memory. That address in turn will contain an integer, which can then be decoded to various machine dependent instructions, such as Load, Store, Jump, etc. Once an instruction has been executed, the program counter is incremented and will now point to the next instruction in memory. A jump now simply consist of a store into the program counter itself.

\section{Logical Relation}
It is well know that defining and reasoning about logical relations for languages with higher-order store, due to the so called World-circularity problem \cite{amal}.

\subsection{Definition of the Logical Relation}
The logical relation is a unary relation that relates a word to  \textit{capability safety} (fig. \ref{fig:relation}). The definition is the Iris implementation of the logical relation defined by Skorstensgaard et. al. \cite{skorstensgaard}. The value relation $\mathcal{V}$ is an Iris relation of type $World \to Word \to iProp$ (we will later discuss what the World type is). We then define the register relation $\mathcal{R}$ as the validity of each word that the register state maps to.

The expression relation should capture what it means for programs to be capability safe. However, as previously discussed, programs in a machine lie in memory, pointed to by the program counter. Strictly speaking, there are no "expressions" in such a machine language. Instead, the expression relation relates Words to the notion of program capability safety, where the word in question will be the program counter state.

\subsection{The Iris Implementation of the Context}
In the expression relation, the \textsf{context} contains the ghost resources necessary to run a program pointed to by the program counter. When the program has finished running, ownership of these resources is given back. To start out, the running program gets full ownership of all machine registers. A ghost register will be represented by the full ownership of the map from RegName to Word. We write 
$$ r \mapsto_r w$$
to indicate that registers r currently contains the word w. The context will contain such a statement for each register in the register map.

A program may also need the ghost resources for the pieces of memory it has authority over.

\begin{figure*}[ht]
	\begin{minipage}{\textwidth}
	\begin{align*}
		\mathcal{E}\interp{W}{pc} \triangleq&~\forall r, \mathcal{R}(W)(r) ~*~\textsf{context}(W)(r[\textsf{\small PC}:=pc])\\
		&\sep~\textsc{WP}~\textsf{Seq (Instr Executable)}~\{ v, v = HaltedV \implies \exists W' r', \textsf{context}(W')(r')\}
	\end{align*}
	\end{minipage}
	\label{fig:expr}
	\caption{Logical Relation for Expressions}
\end{figure*}

\begin{figure*}[ht]
	\begin{minipage}{\textwidth}
	\begin{align*}
		\mathcal{R}\interp{W}{r} \triangleq&~\forall (reg : \text{RegName} \backslash \textsf{PC}), \mathcal{V}\interp{W}{r(reg)}
	\end{align*}
	\end{minipage}
	\label{fig:expr}
	\caption{Logical Relation for Register States}
\end{figure*}

\begin{figure*}[ht]
	\begin{minipage}{\textwidth}
	\begin{align*}
 		\mathcal{V}\interp{W}{z} \triangleq&~\exists z' \in \mathds{Z}. z = z' \\
 		\mathcal{V}\interp{W}{((\text{\tiny{O}},g),b,e,a)} \triangleq&~\top
 \\
 		\left.
		\begin{IEEEeqnarraybox}[\IEEEeqnarraystrutmode \IEEEeqnarraystrutsizeadd{2pt}{2pt}][c]{rCl} 
			\mathcal{V}\interp{W}{((\text{\tiny{RO}},g),b,e,a)} \\
			\mathcal{V}\interp{W}{((\text{\tiny{RW}},g),b,e,a)} \\
			\mathcal{V}\interp{W}{((\text{\tiny{RWL}},g),b,e,a)}
  \end{IEEEeqnarraybox}
  \, \right\}  &\triangleq  \left.\exists~p', p \sqsubseteq p' * \textsf{read\_write\_cond}(\textsf{p',b,e})
  \right. &\textit{where p is {\tiny{RO}}, {\tiny{RW}}, {\tiny{RWL}} resp.}
  \\
   		\left.
		\begin{IEEEeqnarraybox}[\IEEEeqnarraystrutmode \IEEEeqnarraystrutsizeadd{2pt}{2pt}][c]{rCl} 
			\mathcal{V}\interp{W}{((\text{\tiny{RX}},g),b,e,a)} \\
			\mathcal{V}\interp{W}{((\text{\tiny{RWX}},g),b,e,a)} \\
			\mathcal{V}\interp{W}{((\text{\tiny{RWLX}},g),b,e,a)}
  \end{IEEEeqnarraybox}
  \, \right\}  &\triangleq  \left.
	\begin{IEEEeqnarraybox}[
    \IEEEeqnarraystrutmode][c]{rCl}
    \exists~p', p \sqsubseteq p' &*& \textsf{read\_write\_cond}(\textsf{p',b,e})\\
  &*& ~\square~\textsf{exec\_cond(W)(\textsf{p,g,b,e})}
  \end{IEEEeqnarraybox}  
  \right. &\textit{where p is {\tiny{RX}}, {\tiny{RWX}}, {\tiny{RWLX}} resp.}
  \\
 		\mathcal{V}\interp{W}{((\text{\tiny{E}},g),b,e,a)} \triangleq&~\square~\textsf{enter\_cond}(W)(\textsf{g,b,e,a})
	\end{align*}
	\end{minipage}
	\label{fig:relation}
	\caption{Logical Relation for Words}
\end{figure*}

\begin{figure*}[ht]
	\begin{minipage}{\textwidth}
	\begin{align*}
 		\textsf{read\_write\_cond}(p,b,e) \triangleq&~\bigsep_{a \in [b,e]} \text{rel}(a,p,\mathcal{V})\\
 		\textsf{exec\_cond(W)(p,g,b,e)} \triangleq&~\forall a \in [b~e], W' \sqsubseteq^g W.~\rhd~\mathcal{E}\interp{W'}{((p,g),b,e,a)}\\
 		\textsf{enter\_cond(W)(g,b,e,a)} \triangleq&~\forall W' \sqsubseteq^g W.~\rhd~\mathcal{E}\interp{W}{((\textsf{\tiny{RX}},g),b,e,a)}
	\end{align*}
	\end{minipage}
	\\[2em]
	where $(\sigma',\rho_{pub}',\rho_{priv}') \sqsubseteq^{\tiny Global} (\sigma,\rho_{pub},\rho_{priv}) = (\sigma',\rho_{priv}') \sqsubseteq (\sigma,\rho_{priv})$\\ and 
	$(\sigma',\rho_{pub}',\rho_{priv}') \sqsubseteq^{\tiny Local} (\sigma,\rho_{pub},\rho_{priv}) = (\sigma',\rho_{pub}') \sqsubseteq (\sigma,\rho_{pub}))$

	\label{fig:condition}
	\caption{Capability Conditions}
\end{figure*}

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography
%\bibliography{bibfile}

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

\end{document}

### 2019-10-15 prisc-2020/acmart-sigplanproc/acmart-sigplanproc-template.tex
prisc 2020 draft

@@
 Its support of higher-order ghost state makes Iris a great candidate for complex reasoning about \textit{state}.
-This is exactly what we need for reasoning about a machine, where computation \textit{is} state manipulation.
+This is exactly what we need for reasoning about a machine, where computation \textit{is nothing but} state manipulation.
 
@@
 \subsection{Definition of the Logical Relation}
-The logical relation is a unary relation that relates a word to  \textit{capability safety} (fig.
-\ref{fig:relation}).
-The definition is the Iris implementation of the logical relation defined by Skorstensgaard et.
+The logical relation is a unary relation that relates a word to  \textit{capability safety}.
+Figure \ref{fig:relation}) describes the Iris implementation of the logical relation defined by Skorstensgaard et.
 al.
 \cite{skorstensgaard}.
-The value relation $\mathcal{V}$ is an Iris relation of type $World \to Word \to iProp$ (we will later discuss what the World type is).
+The value relation $\mathcal{V}$ is an Iris relation of type $World \to Word \to iProp$ (we will later discuss what the World is).
 We then define the register relation $\mathcal{R}$ as the validity of each word that the register state maps to.
@@
 
-\subsection{The Iris Implementation of the Context}
+\subsection{The Context}
 In the expression relation, the \textsf{context} contains the ghost resources necessary to run a program pointed to by the program counter.
@@
 A program may also need the ghost resources for the pieces of memory it has authority over.
+
+\subsection{The World}
 

### 2019-10-15 prisc-2020/acmart-sigplanproc/acmart-sigplanproc-template.tex
progress on prisc abstract

@@
 \usepackage{tikz}
+\usetikzlibrary{decorations.pathreplacing}
+\usepackage{xcolor}
 \usepackage{graphicx} % need this package
@@
 \newcommand\sepimp{\mathrel{-\mkern-6mu*}}
+
+\usetikzlibrary{decorations.pathreplacing}
+
+\newcommand\MemoryLayout[1]{
+  \begin{tikzpicture}[scale=0.3]
+     \draw[thick](0,0)--++(0,1);
+     \foreach \pt/\col/\lab [remember=\pt as \tp (initially 0)] in {#1} {
+     \if\lab\relax\relax\else
+         \draw[thick,decorate, decoration={brace,amplitude=4mm}]
+            (-\tp,-0.2)--node[below=4mm]{\lab} (-\pt+2,-0.2);
+            \draw[fill=\col] (-\tp,0) rectangle (-\pt+2,1);
+       \fi
+       \foreach \a in {\tp,...,\pt} {
+          \draw(-\a,0) rectangle ++(-1,1);
+       }
+       \draw[thick](-\pt,0)--++(0,1);
+
+     }
+     \node at (-4.5,0.6) {\small B};
+  \end{tikzpicture}
+}
 
@@
 
-Text of paper \ldots
+Part of showing full abstraction consists of showing that the expressive power of a compiled program, is exactly the expressive power as the source program.
+This becomes a challenge when the expressive power of the target language is bigger than the expressive power of the source language.
 
-\section{Iris Primer}
-??
+Take for instance the compilation from a high level functional language, to a low level machine language.
+Certain high level abstractions are taken for granted by the high level language, such as local state encapsulation and well bracketed control flow.
+At the lowest level, a machine may allow arbitrary jumps to instructions in memory, and arbitrarily manipulate pointers to access any local state otherwise hidden by the top level abstraction.
+It must therefore be up to the compiler to enforce such abstractions, if we want to show secure compilation.
 
-\section{Iris Embedding of a Capability Machine}
-The setting: we want to show full abstraction from a high level language, to a low level capability machine.
-Some basic properties of a high level language: local state encapsulation of well bracketed control flow.
-These are not natively enforced by a capability machine.
-When calling an adversary, any guarantees will depend on the calling convention the compiler implements.
-For LSE, the calling convention should ensure that no passed capability give authority over local state.
-For WBCR, the situation is more subtle.
-Skorstensgaard et.
+Capability Machines allow for fine grained control over the \textit{authority} of memory.
+At the machine level, pointers are replaced by capabilities, to which is attached a range of authority and a permission.
+The language then executes instructions by dynamically checking that the instruction uses a capability within its range of authority.
+Using a capability machine language as the target of a secure compiler allows us to take advantage these dynamic checks to enforce the high level abstractions mentioned above.
+
+In \cite{skorstengaardESOP18} and \cite{SkorstengaardPOPL19}, Skorstensgaard et.
 al.
-give detailed explanations of multiple control flow attacks that would happen with a naive calling conventions.
-
-With these goals in mind, it is important not to \textit{accidentally} formalize a language where certain control flow abstractions are enforced.
-We also want to be as architecture independent as possible.
-This we are interested in allowing: arbitrary jumps, and (almost) arbitrary capability manipulations (within the given range of authority).
+present two different calling conventions that enforce well bracketed control flow, and methods for defining and reasoning about capability machines.
+In each case, they define a step indexed Kripke Logical Relation to define a notion of capability safety.
+However, such logical relations are notoriously difficult to reason about \cite{Ahmed2004SemanticsOT}.
+A formalization of these definitions into a proof assistant could facilitate such reasoning, and enable easier exploration of extensions and variations to the language.
 
@@
 Its support of higher-order ghost state makes Iris a great candidate for complex reasoning about \textit{state}.
-This is exactly what we need for reasoning about a machine, where computation \textit{is nothing but} state manipulation.
+This is exactly what we need for reasoning about a machine, where computation consist of state manipulation.
+Furthermore, the model of Iris is itself step-indexed, and comes equipped with a later modality: $\triangleright$.
+The later modality enables elegant Iris definitions of step-indexed logical relations \cite{Krebbers:2017} \cite{Timany:2017}.
+
+The work in process formalization will make the following contributions: 
+\begin{itemize}
+	\item A formalization of a capability machine program logic in Iris, in which programs lie in memory, and execution is completely statefull. 
+	\item An Iris formalization of a Logical Relation to reason about programs in a capability machine language, which can be used to reason about local state while distinguishing between well-bracketed, and non well-bracketed calls.
+\end{itemize}
+
+\section{Iris Embedding of a Capability Machine}
+The first step of formalizing the logical relation, is to embed the capability machine language into Iris.
+We want a formalization which can be used to show full abstraction from a high level language, to a low level capability machine.
+We are interested in the case where the target language does not natively enforce certain basic abstractions of a high level language: local state encapsulation of well bracketed control flow.
+With this goal in mind, it is important not to \textit{accidentally} formalize a language where certain control flow abstractions are still enforced, such as implementing the language using frames and labels.
+Rather, we are interested in reasoning about a language with arbitrary jumps, and (almost) arbitrary capability manipulations (within its given range of authority).
 
@@
 A jump now simply consist of a store into the program counter itself.
+Strictly speaking, such a language does have expressions, rather it executes given the configurations: the state of registers and the state of memory.
 
 \section{Logical Relation}
-It is well know that defining and reasoning about logical relations for languages with higher-order store, due to the so called World-circularity problem \cite{amal}.
-
-\subsection{Definition of the Logical Relation}
-The logical relation is a unary relation that relates a word to  \textit{capability safety}.
-Figure \ref{fig:relation}) describes the Iris implementation of the logical relation defined by Skorstensgaard et.
-al.
-\cite{skorstensgaard}.
-The value relation $\mathcal{V}$ is an Iris relation of type $World \to Word \to iProp$ (we will later discuss what the World is).
-We then define the register relation $\mathcal{R}$ as the validity of each word that the register state maps to.
-
-The expression relation should capture what it means for programs to be capability safe.
-However, as previously discussed, programs in a machine lie in memory, pointed to by the program counter.
-Strictly speaking, there are no "expressions" in such a machine language.
-Instead, the expression relation relates Words to the notion of program capability safety, where the word in question will be the program counter state.
-
-\subsection{The Context}
-In the expression relation, the \textsf{context} contains the ghost resources necessary to run a program pointed to by the program counter.
-When the program has finished running, ownership of these resources is given back.
-To start out, the running program gets full ownership of all machine registers.
-A ghost register will be represented by the full ownership of the map from RegName to Word.
-We write
-$$ r \mapsto_r w$$
-to indicate that registers r currently contains the word w. The context will contain such a statement for each register in the register map.
-
-A program may also need the ghost resources for the pieces of memory it has authority over.
-
-\subsection{The World}
-
 \begin{figure*}[ht]
@@
 		\mathcal{E}\interp{W}{pc} \triangleq&~\forall r, \mathcal{R}(W)(r) ~*~\textsf{context}(W)(r[\textsf{\small PC}:=pc])\\
-		&\sep~\textsc{WP}~\textsf{Seq (Instr Executable)}~\{ v, v = HaltedV \implies \exists W' r', \textsf{context}(W')(r')\}
+		&\sep~\textsc{WP}~\textsf{Seq (Instr Executable)}~\{ v, v = HaltedV \implies \exists W' r', W' \sqsubseteq_{priv} W * \textsf{context}(W')(r')\}
 	\end{align*}
@@
 	\\[2em]
-	where $(\sigma',\rho_{pub}',\rho_{priv}') \sqsubseteq^{\tiny Global} (\sigma,\rho_{pub},\rho_{priv}) = (\sigma',\rho_{priv}') \sqsubseteq (\sigma,\rho_{priv})$\\ and 
-	$(\sigma',\rho_{pub}',\rho_{priv}') \sqsubseteq^{\tiny Local} (\sigma,\rho_{pub},\rho_{priv}) = (\sigma',\rho_{pub}') \sqsubseteq (\sigma,\rho_{pub}))$
+	where $W' \sqsubseteq^{\tiny Global} W \iff W' \sqsubseteq_{priv} W$\\ and 
+	$W' \sqsubseteq^{\tiny Local} W \iff W' \sqsubseteq_{pub} W$
 
@@
 \end{figure*}
+
+\subsection{Definition of the Logical Relation}
+The logical relation is a unary relation that relates a word to  \textit{capability safety}.
+Figure \ref{fig:relation}) describes the Iris implementation of the logical relation defined by Skorstensgaard et.
+al.
+\cite{skorstengaardESOP18}, in which the authors define a capability machine with Local capabilities, which imposes certain restrictions on where to store such capabilities.
+The value relation $\mathcal{V}$ is an Iris relation of type $World \to Word \to iProp$, where the World is a collection of state transition systems used to reason about local state, and a Word can be an integer of a capability.
+We then define the register relation $\mathcal{R}$ as the validity of each word that the register state maps to.
+
+The expression relation should capture what it means for programs to be capability safe.
+However, as previously discussed, programs in a machine lie in memory, pointed to by the program counter.
+Strictly speaking, there are no "expressions" in such a machine language.
+Instead, the expression relation relates Words to the notion of program capability safety, where the word in question will be the program counter state.
+
+\subsection{The Context}
+In the expression relation, the \textsf{context} contains the ghost resources necessary to run a program pointed to by the program counter.
+When the program has finished running, ownership of these resources is given back.
+To start out, the running program gets full ownership of all machine registers.
+A ghost register will be represented by the full ownership of the map from RegName to Word.
+We write
+$$ r \mapsto_r w$$
+to indicate that registers r currently contains the word w. The context will contain such a statement for each register in the register map.
+
+A program may also need the ghost resources for the pieces of memory it has authority over.
+However, we want to respect the principle of least privilege.
+A program should only be able to gain ownership over the ghost resources for memory pointed to by a capability.
+Say a register contains a capability A.
+Consider the following memory layout.
+
+\begin{center}
+\MemoryLayout{
+    12/blue!30!/A,
+    21/blue!30!/B\relax
+  }
+\end{center}
+
+In that case, the program may have authority over the region given by A, as well as the region given by B.
+However if no other capabilities are given, the program should not be given access to the ghost resources denoted by the white regions.
+
+One way to enforce this, is to define the value relation of a capability as an invariant containing the resources for its range of authority, which at the same time guarantees that all the words that these words point to are also valid.
+However, as the name indicates, an invariant can never change.
+This is not a problem when the only variant is the current state of a location.
+In this case however, the value relation is indexed over a particular \textit{World}.
+This World will capture the behaviour of local state.
+The validity of an executable capability will consider any possible future world.
+This becomes crucial when showing safety of programs that depend on well bracketed control flow.
+
+\subsection{The World}
 
@@
 
-%% Bibliography
-%\bibliography{bibfile}
+% Bibliography
+\bibliography{reading_list}
 

### 2019-10-16 prisc-2020/paper.tex
paper added

--- 2019-10-15 prisc-2020/acmart-sigplanproc/acmart-sigplanproc-template.tex
@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %% For double-blind review submission, w/ CCS and ACM Reference
@@
 %% Title information
-\title[Short Title]{Full Title}         %% [Short Title] is optional;
+\title[Short Title]{Mechanized Reasoning about a Capability Machine}
+%% [Short Title] is optional;
                                         %% when present, will be used in
                                         %% header instead of Full Title.
-\titlenote{with title note}             %% \titlenote is optional;
+%\titlenote{with title note}             %% \titlenote is optional;
                                         %% can be repeated if necessary;
                                         %% contents suppressed with 'anonymous'
-\subtitle{Subtitle}                     %% \subtitle is optional
-\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
+%\subtitle{Subtitle}                     %% \subtitle is optional
+%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                         %% can be repeated if necessary;
@@
 %% Author with single affiliation.
-\author{First1 Last1}
+\author{A{\"i}na Linn Georges}
 \authornote{with author1 note}          %% \authornote is optional;
@@
 \affiliation{
-  \position{Position1}
-  \department{Department1}              %% \department is recommended
-  \institution{Institution1}            %% \institution is required
-  \streetaddress{Street1 Address1}
-  \city{City1}
-  \state{State1}
-  \postcode{Post-Code1}
-  \country{Country1}                    %% \country is recommended
+  %\position{Position1}
+  %\department{Department1}              %% \department is recommended
+  \institution{Aarhus University}            %% \institution is required
+  %\streetaddress{Street1 Address1}
+  %\city{City1}
+  %\state{State1}
+  %\postcode{Post-Code1}
+  %\country{Denmark}                    %% \country is recommended
 }
-\email{first1.last1@inst1.edu}          %% \email is recommended
+\email{ageorges@cs.au.dk}          %% \email is recommended
 
-%% Author with two affiliations and emails.
-\author{First2 Last2}
+\author{Lars Birkedal}
 \authornote{with author2 note}          %% \authornote is optional;
@@
 \affiliation{
-  \position{Position2a}
-  \department{Department2a}             %% \department is recommended
-  \institution{Institution2a}           %% \institution is required
-  \streetaddress{Street2a Address2a}
-  \city{City2a}
-  \state{State2a}
-  \postcode{Post-Code2a}
-  \country{Country2a}                   %% \country is recommended
+  %\position{Position2a}
+  %\department{Department2a}             %% \department is recommended
+  \institution{Aarhus University}           %% \institution is required
+  %\streetaddress{Street2a Address2a}
+  %\city{City2a}
+  %\state{State2a}
+  %\postcode{Post-Code2a}
+  %\country{Denmark}                   %% \country is recommended
 }
-\email{first2.last2@inst2a.com}         %% \email is recommended
-\affiliation{
-  \position{Position2b}
-  \department{Department2b}             %% \department is recommended
-  \institution{Institution2b}           %% \institution is required
-  \streetaddress{Street3b Address2b}
-  \city{City2b}
-  \state{State2b}
-  \postcode{Post-Code2b}
-  \country{Country2b}                   %% \country is recommended
-}
-\email{first2.last2@inst2b.org}         %% \email is recommended
+\email{birkedal@cs.au.dk}         %% \email is recommended
 
@@
   %% extraction tools.
+
+  This research was supported in part by the ModuRes Sapere Aude
+  Advanced Grant from The Danish Council for Independent Research
+  for the Natural Sciences (FNU)
+  and by XXX  
   This material is based upon work supported by the

### 2019-10-16 prisc-2020/paper.tex
acks

@@
 
-  This research was supported in part by the ModuRes Sapere Aude
-  Advanced Grant from The Danish Council for Independent Research
-  for the Natural Sciences (FNU)
-  and by XXX  
-  This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
+  This research was supported in part by the ModuRes Sapere Aude Advanced Grant from The Danish Council for Independent Research for the Natural Sciences (FNU) and by a Villum Investigator grant (no.
+  25804), Center for Basic Research in Program Verification (CPV), from the VILLUM Foundation.
+
+  % This material is based upon work supported by the
+  % \grantsponsor{GS100000001}{National Science
+  %   Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
+  % No.~\grantnum{GS100000001}{nnnnnnn} and Grant
+  % No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
+  % conclusions or recommendations expressed in this material are those
+  % of the author and do not necessarily reflect the views of the
+  % National Science Foundation.
 \end{acks}

### 2019-10-16 prisc-2020/paper.tex
abstract draft

@@
 \begin{abstract}
-Text of abstract \ldots.
+  Capability Machines are promising targets for secure compilers since capabilities can be used to enforce abstractions that are usually expected for high-level languages, such as control-flow integrity (CFI) and local state encapsulation (LSE).
+  We present the first formalization of a capability machine that supports mechanized reasoning about deep semantic properties, including CFI and LSE.
+  Our formalization is done in the Coq implementation of Iris, a state-of-the-art concurrent higher-order separation logic, and includes a formalization of the logical relation defined by Skorstensgaard et.
+al.
+\cite{skorstengaardESOP18}, which can used to prove CFI and LSE.
 \end{abstract}

### 2019-10-16 prisc-2020/paper.tex
updated introduction and section 2

@@
 
-Part of showing full abstraction consists of showing that the expressive power of a compiled program, is exactly the expressive power as the source program.
-This becomes a challenge when the expressive power of the target language is bigger than the expressive power of the source language.
-
-Take for instance the compilation from a high level functional language, to a low level machine language.
-Certain high level abstractions are taken for granted by the high level language, such as local state encapsulation and well bracketed control flow.
-At the lowest level, a machine may allow arbitrary jumps to instructions in memory, and arbitrarily manipulate pointers to access any local state otherwise hidden by the top level abstraction.
-It must therefore be up to the compiler to enforce such abstractions, if we want to show secure compilation.
-
+Showing full abstraction consists of showing that the expressive power of a compiled programs is exactly that of the source program.
+Most often, the source language will enforce certain high level abstractions, not enforced in the target language.
 Capability Machines allow for fine grained control over the \textit{authority} of memory.
@@
 The language then executes instructions by dynamically checking that the instruction uses a capability within its range of authority.
-Using a capability machine language as the target of a secure compiler allows us to take advantage these dynamic checks to enforce the high level abstractions mentioned above.
+Using a capability machine language as the target of a secure compiler allows us to take advantage these dynamic checks to enforce such high level abstractions.
 
@@
 In each case, they define a step indexed Kripke Logical Relation to define a notion of capability safety.
-However, such logical relations are notoriously difficult to reason about \cite{Ahmed2004SemanticsOT}.
-A formalization of these definitions into a proof assistant could facilitate such reasoning, and enable easier exploration of extensions and variations to the language.
+
+However, such logical relations are notoriously difficult to construct, due to the so-called World circularity problem \cite{Ahmed2004SemanticsOT}.
+This challenge has been addressed by giving a logical account of the relation, thus raising the level of abstraction
+\cite{Krogh-Jespersen2017} \cite{dreyer_neis_birkedal_2012}.
+Such logical accounts have typically been for logical relations of typed high level languages.
+We show how to give a logical account of an untyped low level language.
 
@@
 Furthermore, the model of Iris is itself step-indexed, and comes equipped with a later modality: $\triangleright$.
-The later modality enables elegant Iris definitions of step-indexed logical relations \cite{Krebbers:2017} \cite{Timany:2017}.
+The later modality, and other logical features of Iris, is enough to define these step-indexed logical relations\cite{Timany:2017} \cite{Jung:2017} \cite{timany2019}.
+Additionally, we can take advantage of the Iris proof mode \cite{Krebbers:2017} for an interactive proof style in Coq.
 
-The work in process formalization will make the following contributions: 
+We have formalized the unary logical relation defined in \cite{skorstengaardESOP18} in Iris. The formalization makes the following contributions: 
 \begin{itemize}
@@
 
-\section{Iris Embedding of a Capability Machine}
-The first step of formalizing the logical relation, is to embed the capability machine language into Iris.
-We want a formalization which can be used to show full abstraction from a high level language, to a low level capability machine.
-We are interested in the case where the target language does not natively enforce certain basic abstractions of a high level language: local state encapsulation of well bracketed control flow.
-With this goal in mind, it is important not to \textit{accidentally} formalize a language where certain control flow abstractions are still enforced, such as implementing the language using frames and labels.
-Rather, we are interested in reasoning about a language with arbitrary jumps, and (almost) arbitrary capability manipulations (within its given range of authority).
+\section{The Program Logic of a Capability Machine}
+In order to give a logical account of the logical relation, we must first define a program logic for the capability machine language.
+Iris is a separation logic framework, which means we can embed any language into Iris by describing its small step semantics.
+However, the framework was designed with high level languages in mind.
+Embedding a language requires a definition of the expressions, the values, the state and the small step semantics.
 
-Machine instructions operate through memory.
+A high level language will typically have expressions, that may reduce to an expression or a value.
+On the other hand, machine instructions operate through memory.
 Typically a machine will have a special register; the program counter, which contains a pointer to an address in memory.
@@
 Once an instruction has been executed, the program counter is incremented and will now point to the next instruction in memory.
-A jump now simply consist of a store into the program counter itself.
-Strictly speaking, such a language does have expressions, rather it executes given the configurations: the state of registers and the state of memory.
+
+Strictly speaking, a machine does not have expressions.
+Rather the small step semantics are defined over machine configurations, and we want the program logic to illustrate that.
+Like in \cite{Jensen:2013}, we define a notion of points-to for registers.
+Since a capability machine replaces pointers with capabilities, we will also replace the conventional points-to for memory with a points-to assertion with a permission attached to it:
+$$ r \mapsto_r w \hspace{1cm} a \mapsto_a[p] w $$
+
+Iris defines the program logic rules as weakest precondition rules. Similar to \cite{Jensen:2013}, the program logic rules for each instruction will be of the following form: 
+\begin{align*}
+\small
+	\textit{run time conditions } \wedge &\textit{ decode}(w) = instr \Rightarrow  \\
+	\{\{\{ \textsf{PC} \mapsto_r ((p,g),b,e,a) &* a \mapsto_a[p] w * ... \}\}\} \\ 
+	\textsf{Instr Executable }\\
+	\{\{\{\textsf{PC} \mapsto_r ((p,g),b,e,a&+1) * a \mapsto_a[p] w * ... \}\}\}
+\end{align*} 
+Where the run time conditions correspond to the dynamic checks done by the capability machine, and \textit{Instr Executable} is a flag indicating how the configuration should be interpreted.
+Unlike \cite{Jensen:2013}, the decoding function is in our case assumed.
+We now have the necessary building blocks for implementing the logical relation.
 

### 2019-10-16 prisc-2020/paper.tex
Fix some typos and compilation issues

@@
 
+\newcommand\world{\emph{World}}
+
 \begin{document}
@@
 \author{A{\"i}na Linn Georges}
-\authornote{with author1 note}          %% \authornote is optional;
+%\authornote{with author1 note}          %% \authornote is optional;
                                         %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
 \affiliation{
@@
 \author{Lars Birkedal}
-\authornote{with author2 note}          %% \authornote is optional;
+%\authornote{with author2 note}          %% \authornote is optional;
                                         %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
 \affiliation{
@@
 \begin{abstract}
-  Capability Machines are promising targets for secure compilers since capabilities can be used to enforce abstractions that are usually expected for high-level languages, such as control-flow integrity (CFI) and local state encapsulation (LSE).
+  Capability machines are promising targets for secure compilers since capabilities can be used to enforce abstractions that are usually expected for high-level languages, such as control-flow integrity (CFI) and local state encapsulation (LSE).
   We present the first formalization of a capability machine that supports mechanized reasoning about deep semantic properties, including CFI and LSE.
@@
 Most often, the source language will enforce certain high level abstractions, not enforced in the target language.
-Capability Machines allow for fine grained control over the \textit{authority} of memory.
+Capability machines allow for fine grained control over the \textit{authority} of memory.
 At the machine level, pointers are replaced by capabilities, to which is attached a range of authority and a permission.
@@
 present two different calling conventions that enforce well bracketed control flow, and methods for defining and reasoning about capability machines.
-In each case, they define a step indexed Kripke Logical Relation to define a notion of capability safety.
+In each case, they define a step-indexed Kripke logical relation to define a notion of capability safety.
 
@@
 \begin{itemize}
-	\item A formalization of a capability machine program logic in Iris, in which programs lie in memory, and execution is completely statefull. 
+	\item A formalization of a capability machine program logic in Iris, in which programs lie in memory, and execution is completely stateful. 
 	\item An Iris formalization of a Logical Relation to reason about programs in a capability machine language, which can be used to reason about local state while distinguishing between well-bracketed, and non well-bracketed calls.
@@
 	\end{minipage}
+	\caption{Logical Relation for Expressions}
 	\label{fig:expr}
-	\caption{Logical Relation for Expressions}
 \end{figure*}
@@
 	\end{minipage}
-	\label{fig:expr}
 	\caption{Logical Relation for Register States}
+	\label{fig:regs}
 \end{figure*}
@@
 	\end{minipage}
+	\caption{Logical Relation for Words}
 	\label{fig:relation}
-	\caption{Logical Relation for Words}
 \end{figure*}
@@
 
+	\caption{Capability Conditions}
 	\label{fig:condition}
-	\caption{Capability Conditions}
 \end{figure*}
@@
 The logical relation is a unary relation that relates a word to  \textit{capability safety}.
-Figure \ref{fig:relation}) describes the Iris implementation of the logical relation defined by Skorstensgaard et.
+Figure \ref{fig:relation} describes the Iris implementation of the logical relation defined by Skorstensgaard et.
 al.
 \cite{skorstengaardESOP18}, in which the authors define a capability machine with Local capabilities, which imposes certain restrictions on where to store such capabilities.
-The value relation $\mathcal{V}$ is an Iris relation of type $World \to Word \to iProp$, where the World is a collection of state transition systems used to reason about local state, and a Word can be an integer of a capability.
+The value relation $\mathcal{V}$ is an Iris relation of type $\world \to \emph{Word} \to \emph{iProp}$, where the World is a collection of state transition systems used to reason about local state, and a Word can be an integer of a capability.
 We then define the register relation $\mathcal{R}$ as the validity of each word that the register state maps to.

### 2019-10-16 prisc-2020/paper.tex
Some comments

@@
 \usepackage{IEEEtrantools}
+\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
+\presetkeys%
+    {todonotes}%
+    {inline,backgroundcolor=orange}{}
 
@@
 
-\newcommand\world{\emph{World}}
+\def\world{\emph{World}}
+\def\word{\emph{Word}}
 
@@
 
+\todo{Alix: Add a sentence to say that we artificially add expressions to fit into the framework, but the operational semantics is equivalent to the standard one}
 Strictly speaking, a machine does not have expressions.
@@
 
+\todo{Alix: We don't really see the weakest precondition in this example, shouldn't we say Hoare triple instead?}
 Iris defines the program logic rules as weakest precondition rules. Similar to \cite{Jensen:2013}, the program logic rules for each instruction will be of the following form: 
@@
 \end{align*} 
-Where the run time conditions correspond to the dynamic checks done by the capability machine, and \textit{Instr Executable} is a flag indicating how the configuration should be interpreted.
+Where the run time conditions correspond to the dynamic checks done by the capability machine, and \textsf{Instr Executable} is a flag \todo{Alix: (an artificial expression)} indicating how the configuration should be interpreted.
 Unlike \cite{Jensen:2013}, the decoding function is in our case assumed.
@@
 al.
-\cite{skorstengaardESOP18}, in which the authors define a capability machine with Local capabilities, which imposes certain restrictions on where to store such capabilities.
-The value relation $\mathcal{V}$ is an Iris relation of type $\world \to \emph{Word} \to \emph{iProp}$, where the World is a collection of state transition systems used to reason about local state, and a Word can be an integer of a capability.
+\cite{skorstengaardESOP18}, in which the authors define a capability machine with so-called \emph{local} capabilities, which imposes certain restrictions on where to store such capabilities.
+The value relation $\mathcal{V}$ is an Iris relation of type $\world \to \word \to \emph{iProp}$, where \world{} is a collection of state transition systems used to reason about local states\todo{Alix: explain local as in local to a function call}, and a Word can be an integer of a capability.
 We then define the register relation $\mathcal{R}$ as the validity of each word that the register state maps to.
@@
 Strictly speaking, there are no "expressions" in such a machine language.
-Instead, the expression relation relates Words to the notion of program capability safety, where the word in question will be the program counter state.
+Instead, the expression relation relates words to the notion of program capability safety, where the word in question is pointed to by the program counter.
 
 \subsection{The Context}
+\todo{Everything ghost might be too much of a Iris'sm ?}
 In the expression relation, the \textsf{context} contains the ghost resources necessary to run a program pointed to by the program counter.
@@
 This is not a problem when the only variant is the current state of a location.
-In this case however, the value relation is indexed over a particular \textit{World}.
+In this case however, the value relation is indexed over a particular \world.
 This World will capture the behaviour of local state.

### 2019-10-16 prisc-2020/paper.tex
one more comment
changes to section 3

@@
 $$ r \mapsto_r w \hspace{1cm} a \mapsto_a[p] w $$
+\todo{Alix: give a quick introduction to permissions, and says for instance that a RO permission does not allow to update the ressource}
 
@@
 \subsection{Definition of the Logical Relation}
-The logical relation is a unary relation that relates a word to  \textit{capability safety}.
-Figure \ref{fig:relation} describes the Iris implementation of the logical relation defined by Skorstensgaard et.
-al.
-\cite{skorstengaardESOP18}, in which the authors define a capability machine with so-called \emph{local} capabilities, which imposes certain restrictions on where to store such capabilities.
-The value relation $\mathcal{V}$ is an Iris relation of type $\world \to \word \to \emph{iProp}$, where \world{} is a collection of state transition systems used to reason about local states\todo{Alix: explain local as in local to a function call}, and a Word can be an integer of a capability.
+The logical relation is a unary relation that relates a word to a notion of \textit{capability safety}.
+The value relation $\mathcal{V}$ is an Iris relation of type $World \to Word \to iProp$, where the World is a collection of state transition systems used to reason about local state\todo{Alix: explain local as in local to a function call}, and a Word can be an integer or a capability.
 We then define the register relation $\mathcal{R}$ as the validity of each word that the register state maps to.
@@
 However, as previously discussed, programs in a machine lie in memory, pointed to by the program counter.
-Strictly speaking, there are no "expressions" in such a machine language.
-Instead, the expression relation relates words to the notion of program capability safety, where the word in question is pointed to by the program counter.
+The expression relation relates Words to the notion of program capability safety, where the word in question will be the current state of the program.
 
@@
 When the program has finished running, ownership of these resources is given back.
+
 To start out, the running program gets full ownership of all machine registers.
 A ghost register will be represented by the full ownership of the map from RegName to Word.
-We write
-$$ r \mapsto_r w$$
-to indicate that registers r currently contains the word w. The context will contain such a statement for each register in the register map.
+The context will contain a points-to assertion for each register in the register map.
 
@@
 One way to enforce this, is to define the value relation of a capability as an invariant containing the resources for its range of authority, which at the same time guarantees that all the words that these words point to are also valid.
+The memory resource will then be given through the $\mathcal{R}$ relation.
+
 However, as the name indicates, an invariant can never change.
 This is not a problem when the only variant is the current state of a location.
-In this case however, the value relation is indexed over a particular \world.
-This World will capture the behaviour of local state.
+In this case however, the value relation is parametrized over a particular World.
+The World captures the behaviour of local state.
 The validity of an executable capability will consider any possible future world.
 This becomes crucial when showing safety of programs that depend on well bracketed control flow.
+We do not want the invariant to depend on any one World.
 
-\subsection{The World}
-
+Instead, we use Iris \textit{saved predicates}.
+The context contains the full authority over a map from locations to saved predicate names, asserting the resource for each location in the map, as well as the saved predicate applied to the current World and the word the location maps to.
+The map itself is abstracted by an existential quantifier.
+The Iris proposition rel($a,p,\mathcal{V}$) then states that the location a is included in that map, the points to relation has permission p, and the saved predicate is the value relation.
+\newpage
 %% Acknowledgments

### 2019-10-17 prisc-2020/paper.tex
fixing authors
notes added

@@
 
+\author{Alix Trieu}
+%\authornote{with author2 note}          %% \authornote is optional;
+                                        %% can be repeated if necessary
+%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\affiliation{
+  %\position{Position2a}
+  %\department{Department2a}             %% \department is recommended
+  \institution{Aarhus University}           %% \institution is required
+  %\streetaddress{Street2a Address2a}
+  %\city{City2a}
+  %\state{State2a}
+  %\postcode{Post-Code2a}
+  %\country{Denmark}                   %% \country is recommended
+}
+\email{alix.trieu@cs.au.dk}         %% \email is recommended
+
 \author{Lars Birkedal}
@@
 \section{The Program Logic of a Capability Machine}
+
+\begin{itemize}
+\item we define the operational semantics of the CM
+\item we prove basic hoare triples for each of the atomic instructions
+  of the CM (i.e., for load, store, etc.)
+\item non-trivial
+  \begin{itemize}
+  \item while Iris is a framework and supports many languages, it is
+    geared towards models of higher-level languages which abstract from
+    program counters and the fact that programs are stored in memory.
+    -- here our low-level machine model includes registers, in particular
+    the program counter register, and instructions are stored in memory.
+   \item hence we 
+  \end{itemize}
+\end{itemize}
+
 In order to give a logical account of the logical relation, we must first define a program logic for the capability machine language.

### 2019-10-17 prisc-2020/paper.tex

@@
     the program counter register, and instructions are stored in memory.
-   \item hence we 
+  \item to capture the semantics of the CM in the Iris framework, 
+    we introduce an abstract notion of instruction, whose
+    operational meaning is to execute the next instruction the program counter
+    points to, and abstract notions of values, failed and halted.
+  \item for each machine instruction, we can then prove a hoare triple
+    which expresses the pre- and post-conditions for executing the instruction
   \end{itemize}
@@
 \section{Logical Relation}
+
+\begin{itemize}
+\item LS: worlds contained all invariants, both on local state and ``standard islands'' for capabilities
+\item now, simplified, world only contains sts's for local state, whereas the conditions on capabilities are captured by the permission of the points-to connective
+\end{itemize}
+
 \begin{figure*}[ht]

### 2019-10-17 prisc-2020/paper.tex
intro pass

@@
 \begin{abstract}
-  Capability machines are promising targets for secure compilers since capabilities can be used to enforce abstractions that are usually expected for high-level languages, such as control-flow integrity (CFI) and local state encapsulation (LSE).
-  We present the first formalization of a capability machine that supports mechanized reasoning about deep semantic properties, including CFI and LSE.
+  Capability machines are promising targets for secure compilers since capabilities can be used to enforce abstractions that are usually expected for high-level languages, such as well-bracketed control-flow (WBCF) and local state encapsulation (LSE).
+  We present the first formalization of a capability machine that supports mechanized reasoning about deep semantic properties, including WBCF and LSE.
   Our formalization is done in the Coq implementation of Iris, a state-of-the-art concurrent higher-order separation logic, and includes a formalization of the logical relation defined by Skorstensgaard et.
 al.
-\cite{skorstengaardESOP18}, which can used to prove CFI and LSE.
+\cite{skorstengaardESOP18}, which can used to prove WBCF and LSE.
 \end{abstract}
@@
 
-Showing full abstraction consists of showing that the expressive power of a compiled programs is exactly that of the source program.
-Most often, the source language will enforce certain high level abstractions, not enforced in the target language.
-Capability machines allow for fine grained control over the \textit{authority} of memory.
+Capability machines allow for fine grained control over the authority of memory \cite{levy_capability-based_1984,watson_cheri:_2015}.
 At the machine level, pointers are replaced by capabilities, to which is attached a range of authority and a permission.
-The language then executes instructions by dynamically checking that the instruction uses a capability within its range of authority.
-Using a capability machine language as the target of a secure compiler allows us to take advantage these dynamic checks to enforce such high level abstractions.
+When the machine executes an instruction it dynamically checks that the instruction uses a capability within its range of authority.
+Capability machines are promising targets for secure compilers because these dynamic checks can be used to enforce abstractions that are usually expected for high-level languages, such as well-bracketed
+control-flow (WBCF) and local state encapsulation (LSE).
 
-In \cite{skorstengaardESOP18} and \cite{SkorstengaardPOPL19}, Skorstensgaard et.
+We present the first formalization of a capability machine that supports mechanized reasoning about deep semantic properties, including WBCF and LSE.
+
+Our formalization builds upon earlier work of Skorstensgaard et.
+al.
+\cite{skorstengaardESOP18} and \cite{SkorstengaardPOPL19}, who present two different capability machines and calling conventions that enforce well-bracketed control flow, and methods for defining and reasoning about capability machines.
+In each case, they define a logical relation to capture a semantic
+notion of capability safety and use it to prove WBCF and LSE.
+
+The logical relations in \cite{skorstengaardESOP18} and \cite{SkorstengaardPOPL19} are so-called step-indexed Kripke logical relations, which means that they are indexed over recursively-defined worlds, which contain descriptions of invariants of the memory of the machine.
+It is well-known that it is non-trivial to define and work with such step-indexed Kripke logical relations~\cite{ahmed-2002, birkedal2011step}.
+Therefore we do not formalize the logical relations of Skorstensgaard et.
+al.
+directly, but rather give a more abstract \emph{logical} definition of the logical relations in the Iris program logic, which comes with built-in support for abstract reasoning about recursion (qua the later modality and L{\"o}b induction) and invariants.
+Another reason for using Iris is that we can use the Coq implementation of Iris and
+the Iris proof mode \cite{Krebbers:2017} to mechanize our development.
+
+In summary, we present Iris formalizations of 
+\begin{itemize}
+  \item a program logic for reasoning about capability machine programs.
+  \item the logical relation from \cite{skorstengaardESOP18}, 
+     which captures capability safety and which can be used to reason about WBCF and LSE.
+\end{itemize}
+All of the technical development is mechanized in Coq using the Coq implementation of Iris.
+
+% Such a logical approach to defining logical relations has been used successfully before for logical relations for typed high-level languages (e.g., \cite{lslr, XXX}); here we use it for the first time for a low-level untyped machine language.
+
+% In \cite{skorstengaardESOP18} and \cite{SkorstengaardPOPL19}, Skorstensgaard et.
 al.
@@
 
-However, such logical relations are notoriously difficult to construct, due to the so-called World circularity problem \cite{Ahmed2004SemanticsOT}.
+% However, such logical relations are notoriously difficult to construct, due to the so-called World circularity problem \cite{Ahmed2004SemanticsOT}.
 This challenge has been addressed by giving a logical account of the relation, thus raising the level of abstraction
-\cite{Krogh-Jespersen2017} \cite{dreyer_neis_birkedal_2012}.
+% \cite{Krogh-Jespersen2017} \cite{dreyer_neis_birkedal_2012}.
 Such logical accounts have typically been for logical relations of typed high level languages.
@@
 
-Iris is a higher-order concurrent separation logic framework.
+% Iris is a higher-order concurrent separation logic framework.
 Its support of higher-order ghost state makes Iris a great candidate for complex reasoning about \textit{state}.
@@
 
-We have formalized the unary logical relation defined in \cite{skorstengaardESOP18} in Iris. The formalization makes the following contributions: 
-\begin{itemize}
-	\item A formalization of a capability machine program logic in Iris, in which programs lie in memory, and execution is completely stateful. 
-	\item An Iris formalization of a Logical Relation to reason about programs in a capability machine language, which can be used to reason about local state while distinguishing between well-bracketed, and non well-bracketed calls.
-\end{itemize}
+% We have formalized the unary logical relation defined in \cite{skorstengaardESOP18} in Iris. The formalization makes the following contributions: 
+% \begin{itemize}
+% 	\item A formalization of a capability machine program logic in Iris, in which programs lie in memory, and execution is completely stateful. 
+% 	\item An Iris formalization of a Logical Relation to reason about programs in a capability machine language, which can be used to reason about local state while distinguishing between well-bracketed, and non well-bracketed calls.
+% \end{itemize}
 

### 2019-10-17 prisc-2020/paper.tex
intro pass

@@
 al.
-directly, but rather give a more abstract \emph{logical} definition of the logical relations in the Iris program logic, which comes with built-in support for abstract reasoning about recursion (qua the later modality and L{\"o}b induction) and invariants.
+directly, but rather give a more abstract \emph{logical} definition of the logical relations in the Iris program logic framework \cite{iris,iris2,iris3, iris-jfp}, which comes with built-in support for abstract reasoning about recursion (qua the later modality and L{\"o}b induction) and invariants.
+Such a logical approach to defining logical relations has been used successfully before for logical relations for typed high-level languages (e.g., \cite{lslr-journal,iris:type-effect,rustbelt,runst-popl18}); here we use it for the first time for a low-level untyped machine language.
 Another reason for using Iris is that we can use the Coq implementation of Iris and
@@
 
+% Such a logical approach to defining logical relations has been used successfully before for logical relations for typed high-level languages (e.g., \cite{lslr, XXX}); here we use it for the first time for a low-level untyped machine language.
 % Such a logical approach to defining logical relations has been used successfully before for logical relations for typed high-level languages (e.g., \cite{lslr, XXX}); here we use it for the first time for a low-level untyped machine language.

### 2019-10-17 prisc-2020/paper.tex
section 3

--- 2019-10-17 prisc-2020/paper.tex
@@
 
-\subsection{Definition of the Logical Relation}
+\subsection{A Logical Account of the Value Logical Relation}
 The logical relation is a unary relation that relates a word to a notion of \textit{capability safety}.
-The value relation $\mathcal{V}$ is an Iris relation of type $World \to Word \to iProp$, where the World is a collection of state transition systems used to reason about local state\todo{Alix: explain local as in local to a function call}, and a Word can be an integer or a capability.
-We then define the register relation $\mathcal{R}$ as the validity of each word that the register state maps to.
+The value relation $\mathcal{V}$ is an Iris relation of type $World \to Word \to iProp$, where the World is a collection of state transition systems used to reason about local state.
 
-The expression relation should capture what it means for programs to be capability safe.
-However, as previously discussed, programs in a machine lie in memory, pointed to by the program counter.
-The expression relation relates Words to the notion of program capability safety, where the word in question will be the current state of the program.
+This notion of World is different from the logical relation given by \cite{skorstengaardESOP18}, where the World is a collection of invariants describing the behaviour of all memory.
+Skorstensgaard et.
+al.
+use the World to define validity of capabilities, by relating it to certain standard islands for capabilities.
+
+In the logical account, we use the capability points-to assertion to both describe what we may get from the read permission of a capability, and the restrictions imposed on us by the write permission of a capability.
+On one hand, a points-to assertion lets us execute an instruction using that location.
+On the other, the permission of the points-to assertion restricts which updates are allowed, and thus which instructions will fail.
+The permission of the points-to assertion suffices to describe the behaviour of the above mentioned standard islands.
+
+We need the value relation both to give us the points-to predicate, while asserting that what the predicate points to is valid. The challenge is now to define how the value relation can do that.
+
+\subsection{A Logical Account of the Expression Relation}
+
+Logical accounts of Logical Relations for high level languages with references have used Iris invariants to define the validity of a reference.
+In order to reason about local state, it suffices to define a future world relation only in the expression relation.
+This approach works in a high level language, where the calls are inherently well-bracketed.
+
+In our case however, jumping to such an executable capability may be non well-bracketed.
+Skorstensgaard et.
+al.
+distinguishes between well-bracketed and non well-bracketed calls by respectively using public- and private future world relations given the appropriate conditions.
+
+Iris does not make such a distinction in its implementation of invariants.
+Using invariants alone is therefore not sufficient for our purposes.
+Instead we explicitly define notions of public and private future worlds, and use these notions to distinguish between well-bracketed and non well-bracketed calls.
+
+A callback is an executable capability.
+This distinction must therefore be done by the value relation.
+However, if the value relation is indexed by a world, we again have difficulties using Iris invariants.
+Asserting $\exists w, \mathcal{V}(W)(w)$ in an invariant does not allow for $W$ to change.
+
+This particular challenge was solved by using
 

### 2019-10-17 prisc-2020/paper.tex
progress towards section 3, edits in section 2

@@
 Iris is a separation logic framework, which means we can embed any language into Iris by describing its small step semantics.
-However, the framework was designed with high level languages in mind.
-Embedding a language requires a definition of the expressions, the values, the state and the small step semantics.
 
-A high level language will typically have expressions, that may reduce to an expression or a value.
-On the other hand, machine instructions operate through memory.
+While Iris is a framework and supports many languages, it is geared towards models of higher-level languages.
+Embedding a language asks for a definition of expressions, values, and the state, which abstracts away from the fact that programs are stored in memory.
+
 Typically a machine will have a special register; the program counter, which contains a pointer to an address in memory.
@@
 
-\todo{Alix: Add a sentence to say that we artificially add expressions to fit into the framework, but the operational semantics is equivalent to the standard one}
-Strictly speaking, a machine does not have expressions.
-Rather the small step semantics are defined over machine configurations, and we want the program logic to illustrate that.
+To capture the semantics of the Capability Machine in the Iris framework, we introduce an abstract notion of an instruction, whose operational meaning is to execute the instruction the program counter points to, and an abstraction notion of values for halted and failed configurations.
+
+Next we use the Iris framework to prove Hoare Triples for each instruction.
 Like in \cite{Jensen:2013}, we define a notion of points-to for registers.
 Since a capability machine replaces pointers with capabilities, we will also replace the conventional points-to for memory with a points-to assertion with a permission attached to it:
-$$ r \mapsto_r w \hspace{1cm} a \mapsto_a[p] w $$
-\todo{Alix: give a quick introduction to permissions, and says for instance that a RO permission does not allow to update the ressource}
+$$ r \mapsto_r w \hspace{1cm} a \mapsto_a[p]~w $$
+The capability points-to predicate states that address $a$ points to word $w$ with permission $p$.
+The permission restricts the words an address can update to.
+For instance, in $a \mapsto_a[\textsc{\tiny RO}]~w$, the $\textsc{\tiny RO}$ permission does not allow for any updates to $a$.
 
 \todo{Alix: We don't really see the weakest precondition in this example, shouldn't we say Hoare triple instead?}
-Iris defines the program logic rules as weakest precondition rules. Similar to \cite{Jensen:2013}, the program logic rules for each instruction will be of the following form: 
+Similar to \cite{Jensen:2013}, the program logic rules for each instruction are of the following form: 
 \begin{align*}
@@
 \end{align*} 
-Where the run time conditions correspond to the dynamic checks done by the capability machine, and \textsf{Instr Executable} is a flag \todo{Alix: (an artificial expression)} indicating how the configuration should be interpreted.
+Where the run time conditions correspond to the dynamic checks done by the capability machine, and \textsf{Instr Executable} is the abstract expression \todo{Alix: (an artificial expression)} for executing the next instruction in memory.
 Unlike \cite{Jensen:2013}, the decoding function is in our case assumed.
@@
 \section{Logical Relation}
-
-\begin{itemize}
-\item LS: worlds contained all invariants, both on local state and ``standard islands'' for capabilities
-\item now, simplified, world only contains sts's for local state, whereas the conditions on capabilities are captured by the permission of the points-to connective
-\end{itemize}
 
@@
 
-\subsection{A Logical Account of the Value Logical Relation}
+\subsection{The Value Relation}
+\begin{itemize}
+	\item definition of the relation and World
+	\item LS definitions
+	\item now: standard islands are not used 
+\end{itemize}
+
 The logical relation is a unary relation that relates a word to a notion of \textit{capability safety}.
@@
 al.
-use the World to define validity of capabilities, by relating it to certain standard islands for capabilities.
+use the World to define validity of capabilities, by relating the invariant of the region in question to certain standard islands for capabilities.
 
 In the logical account, we use the capability points-to assertion to both describe what we may get from the read permission of a capability, and the restrictions imposed on us by the write permission of a capability.
-On one hand, a points-to assertion lets us execute an instruction using that location.
+On the one hand, a points-to assertion lets us execute an instruction using that location.
 On the other, the permission of the points-to assertion restricts which updates are allowed, and thus which instructions will fail.
@@
 
-We need the value relation both to give us the points-to predicate, while asserting that what the predicate points to is valid. The challenge is now to define how the value relation can do that.
+We want the value relation to give us the points-to predicate, while asserting that what the predicate points to is valid. The challenge is now to define how that should be done.
 
-\subsection{A Logical Account of the Expression Relation}
+\subsection{Reasoning about Local State}
+\begin{itemize}
+	\item How to reason about local state in high level languages
+	\item public/private distinction
+	\item TT style
+	\item difficulties w/ iris invariants 
+\end{itemize}
 
-Logical accounts of Logical Relations for high level languages with references have used Iris invariants to define the validity of a reference.
-In order to reason about local state, it suffices to define a future world relation only in the expression relation.
+Logical accounts of relations for high level languages with references have used Iris invariants to define the validity of a reference.
+In order to reason about local state, it suffices to add a future world relation only in the definition of the expression relation.
 This approach works in a high level language, where the calls are inherently well-bracketed.
 
-In our case however, jumping to such an executable capability may be non well-bracketed.
+In our case however, calls will not in all cases be well-bracketed.
 Skorstensgaard et.
@@
 
-\subsection{The Context}
-\todo{Everything ghost might be too much of a Iris'sm ?}
-In the expression relation, the \textsf{context} contains the ghost resources necessary to run a program pointed to by the program counter.
-When the program has finished running, ownership of these resources is given back.
-
-To start out, the running program gets full ownership of all machine registers.
-A ghost register will be represented by the full ownership of the map from RegName to Word.
-The context will contain a points-to assertion for each register in the register map.
-
-A program may also need the ghost resources for the pieces of memory it has authority over.
-However, we want to respect the principle of least privilege.
-A program should only be able to gain ownership over the ghost resources for memory pointed to by a capability.
-Say a register contains a capability A.
-Consider the following memory layout.
-
-\begin{center}
-\MemoryLayout{
-    12/blue!30!/A,
-    21/blue!30!/B\relax
-  }
-\end{center}
-
-In that case, the program may have authority over the region given by A, as well as the region given by B.
-However if no other capabilities are given, the program should not be given access to the ghost resources denoted by the white regions.
-
-One way to enforce this, is to define the value relation of a capability as an invariant containing the resources for its range of authority, which at the same time guarantees that all the words that these words point to are also valid.
-The memory resource will then be given through the $\mathcal{R}$ relation.
-
-However, as the name indicates, an invariant can never change.
-This is not a problem when the only variant is the current state of a location.
-In this case however, the value relation is parametrized over a particular World.
-The World captures the behaviour of local state.
-The validity of an executable capability will consider any possible future world.
-This becomes crucial when showing safety of programs that depend on well bracketed control flow.
-We do not want the invariant to depend on any one World.
-
-Instead, we use Iris \textit{saved predicates}.
-The context contains the full authority over a map from locations to saved predicate names, asserting the resource for each location in the map, as well as the saved predicate applied to the current World and the word the location maps to.
-The map itself is abstracted by an existential quantifier.
-The Iris proposition rel($a,p,\mathcal{V}$) then states that the location a is included in that map, the points to relation has permission p, and the saved predicate is the value relation.
-\newpage
 %% Acknowledgments

### 2019-10-17 prisc-2020/paper.tex
a bit on saved predicates

@@
 
-Our formalization builds upon earlier work of Skorstensgaard et.
-al.
-\cite{skorstengaardESOP18} and \cite{SkorstengaardPOPL19}, who present two different capability machines and calling conventions that enforce well-bracketed control flow, and methods for defining and reasoning about capability machines.
+Our formalization builds upon earlier work of \citeauthor{skorstengaardESOP18} \cite{skorstengaardESOP18} and \cite{SkorstengaardPOPL19}, who present two different capability machines and calling conventions that enforce well-bracketed control flow, and methods for defining and reasoning about capability machines.
 In each case, they define a logical relation to capture a semantic
@@
 It is well-known that it is non-trivial to define and work with such step-indexed Kripke logical relations~\cite{ahmed-2002, birkedal2011step}.
-Therefore we do not formalize the logical relations of Skorstensgaard et.
-al.
+Therefore we do not formalize the logical relations of \citeauthor{skorstengaardESOP18} 
 directly, but rather give a more abstract \emph{logical} definition of the logical relations in the Iris program logic framework \cite{iris,iris2,iris3, iris-jfp}, which comes with built-in support for abstract reasoning about recursion (qua the later modality and L{\"o}b induction) and invariants.
@@
 
-In our case however, calls will not in all cases be well-bracketed.
-Skorstensgaard et.
-al.
+In our case however, calls are not always well-bracketed.
+\citeauthor{skorstengaardESOP18}.
 distinguishes between well-bracketed and non well-bracketed calls by respectively using public- and private future world relations given the appropriate conditions.
@@
 Iris does not make such a distinction in its implementation of invariants.
-Using invariants alone is therefore not sufficient for our purposes.
-Instead we explicitly define notions of public and private future worlds, and use these notions to distinguish between well-bracketed and non well-bracketed calls.
+Using invariants alone, is therefore not sufficient for our purposes.
+Instead we explicitly define notions of public and private future world relations, and use these notions to distinguish between well-bracketed and non well-bracketed calls.
 
@@
 However, if the value relation is indexed by a world, we again have difficulties using Iris invariants.
-Asserting $\exists w, \mathcal{V}(W)(w)$ in an invariant does not allow for $W$ to change.
+Asserting $\exists w, \mathcal{V}(W)(w)$ as invariant does not allow for $W$ to change.
 
-This particular challenge was solved by using
+The difficulty is to support reasoning about local state, while distinguishing between public- and private future world relations in the value relation, rather than the expression relation.
+We implement this by using Iris' saved predicates, which allows us to save a predicate by associating it to a unique name.
+Using that unique name, we are then able to refer to that predicate somewhere else, and state it with the argument that we want.
+That way, if we save the value relation predicate and associate it to a location, we can later on state that predicate to the future world in question.
 

### 2019-10-17 prisc-2020/paper.tex
typos

@@
 
-Iris does not make such a distinction in its implementation of invariants.
-Using invariants alone, is therefore not sufficient for our purposes.
+Iris invariants do not make such a distinction.
+Using invariants alone is therefore not sufficient for our purposes.
 Instead we explicitly define notions of public and private future world relations, and use these notions to distinguish between well-bracketed and non well-bracketed calls.
@@
 Using that unique name, we are then able to refer to that predicate somewhere else, and state it with the argument that we want.
-That way, if we save the value relation predicate and associate it to a location, we can later on state that predicate to the future world in question.
+That way, if we save the value relation predicate and associate it to a location, we can later on state that predicate and apply it to the future world in question.
 

### 2019-10-17 prisc-2020/paper.tex
Small comment

@@
 
+\todo{Alix: I think we need a short discussion (not specifically here) about the fact that we don't really have a theorem that says something like "being in the logical relation implies LSE and WBCF" as that's not really well defined, but we can prove things about specific programs that rely on it}
+
 % Such a logical approach to defining logical relations has been used successfully before for logical relations for typed high-level languages (e.g., \cite{lslr, XXX}); here we use it for the first time for a low-level untyped machine language.
@@
 
-Next we use the Iris framework to prove Hoare Triples for each instruction.
+Next we use the Iris framework to prove Hoare triples for each instruction.
 Like in \cite{Jensen:2013}, we define a notion of points-to for registers.
@@
 
-\todo{Alix: We don't really see the weakest precondition in this example, shouldn't we say Hoare triple instead?}
 Similar to \cite{Jensen:2013}, the program logic rules for each instruction are of the following form: 
@@
 \end{align*} 
-Where the run time conditions correspond to the dynamic checks done by the capability machine, and \textsf{Instr Executable} is the abstract expression \todo{Alix: (an artificial expression)} for executing the next instruction in memory.
+Where the run time conditions correspond to the dynamic checks done by the capability machine, and \textsf{Instr Executable} is the abstract expression for executing the next instruction in memory.
 Unlike \cite{Jensen:2013}, the decoding function is in our case assumed.

### 2019-10-17 prisc-2020/paper.tex
types and commenting out a technical paragraph in section 3

@@
 
-\begin{itemize}
-\item we define the operational semantics of the CM
-\item we prove basic hoare triples for each of the atomic instructions
-  of the CM (i.e., for load, store, etc.)
-\item non-trivial
-  \begin{itemize}
-  \item while Iris is a framework and supports many languages, it is
-    geared towards models of higher-level languages which abstract from
-    program counters and the fact that programs are stored in memory.
-    -- here our low-level machine model includes registers, in particular
-    the program counter register, and instructions are stored in memory.
-  \item to capture the semantics of the CM in the Iris framework, 
-    we introduce an abstract notion of instruction, whose
-    operational meaning is to execute the next instruction the program counter
-    points to, and abstract notions of values, failed and halted.
-  \item for each machine instruction, we can then prove a hoare triple
-    which expresses the pre- and post-conditions for executing the instruction
-  \end{itemize}
-\end{itemize}
+%\begin{itemize}
+%\item we define the operational semantics of the CM
+%\item we prove basic hoare triples for each of the atomic instructions
+%  of the CM (i.e., for load, store, etc.)
+%\item non-trivial
+%  \begin{itemize}
+%  \item while Iris is a framework and supports many languages, it is
+%    geared towards models of higher-level languages which abstract from
+%    program counters and the fact that programs are stored in memory.
+%    -- here our low-level machine model includes registers, in particular
+%    the program counter register, and instructions are stored in memory.
+%  \item to capture the semantics of the CM in the Iris framework, 
+%    we introduce an abstract notion of instruction, whose
+%    operational meaning is to execute the next instruction the program counter
+%    points to, and abstract notions of values, failed and halted.
+%  \item for each machine instruction, we can then prove a hoare triple
+%    which expresses the pre- and post-conditions for executing the instruction
+%  \end{itemize}
+%\end{itemize}
 
@@
 
-To capture the semantics of the Capability Machine in the Iris framework, we introduce an abstract notion of an instruction, whose operational meaning is to execute the instruction the program counter points to, and an abstraction notion of values for halted and failed configurations.
+To capture the semantics of the Capability Machine in the Iris framework, we introduce an abstract notion of an instruction, whose operational meaning is to execute the instruction the program counter points to, and an abstract notion of values for halted and failed configurations.
 
@@
 \subsection{The Value Relation}
-\begin{itemize}
-	\item definition of the relation and World
-	\item LS definitions
-	\item now: standard islands are not used 
-\end{itemize}
+%\begin{itemize}
+%	\item definition of the relation and World
+%	\item LS definitions
+%	\item now: standard islands are not used 
+%\end{itemize}
 
@@
 \subsection{Reasoning about Local State}
-\begin{itemize}
-	\item How to reason about local state in high level languages
-	\item public/private distinction
-	\item TT style
-	\item difficulties w/ iris invariants 
-\end{itemize}
+%\begin{itemize}
+%	\item How to reason about local state in high level languages
+%	\item public/private distinction
+%	\item TT style
+%	\item difficulties w/ iris invariants 
+%\end{itemize}
 
@@
 In our case however, calls are not always well-bracketed.
-\citeauthor{skorstengaardESOP18}.
-distinguishes between well-bracketed and non well-bracketed calls by respectively using public- and private future world relations given the appropriate conditions.
+\citeauthor{skorstengaardESOP18} distinguishes between well-bracketed and non well-bracketed calls by respectively using public- and private future world relations at the site of the callback.
 
-Iris invariants do not make such a distinction.
+%Iris invariants do not make such a distinction.
 Using invariants alone is therefore not sufficient for our purposes.
@@
 This distinction must therefore be done by the value relation.
-However, if the value relation is indexed by a world, we again have difficulties using Iris invariants.
-Asserting $\exists w, \mathcal{V}(W)(w)$ as invariant does not allow for $W$ to change.
+However, if the value relation is parametrized by a world, we have difficulties using Iris invariants.
+Asserting $\mathcal{V}(W)(w)$ as invariant does not allow for $W$ to change.
 
-The difficulty is to support reasoning about local state, while distinguishing between public- and private future world relations in the value relation, rather than the expression relation.
+The difficulty lies in supporting reasoning about local state, while distinguishing between public- and private future world relations in the value relation, rather than the expression relation.
 We implement this by using Iris' saved predicates, which allows us to save a predicate by associating it to a unique name.

### 2019-10-17 prisc-2020/paper.tex
editing pass

@@
   \item the logical relation from \cite{skorstengaardESOP18}, 
-     which captures capability safety and which can be used to reason about WBCF and LSE.
+    which captures capability safety and which can be used to reason about examples
+    that rely on WBCF and LSE.
 \end{itemize}
@@
 \todo{Alix: I think we need a short discussion (not specifically here) about the fact that we don't really have a theorem that says something like "being in the logical relation implies LSE and WBCF" as that's not really well defined, but we can prove things about specific programs that rely on it}
+\todo{Lars: I just tried to reduce the claim above}
 
@@
 
-\section{The Program Logic of a Capability Machine}
+\section{A Program Logic for a Capability Machine}
 
-%\begin{itemize}
-%\item we define the operational semantics of the CM
-%\item we prove basic hoare triples for each of the atomic instructions
+% \begin{itemize}
+% \item we define the operational semantics of the CM
+% \item we prove basic hoare triples for each of the atomic instructions
 %   of the CM (i.e., for load, store, etc.)
-%\item non-trivial
+% \item non-trivial
 %   \begin{itemize}
@@
 %   \end{itemize}
-%\end{itemize}
+% \end{itemize}
 
-In order to give a logical account of the logical relation, we must first define a program logic for the capability machine language.
-Iris is a separation logic framework, which means we can embed any language into Iris by describing its small step semantics.
+In this section we give a brief overview of how we define a program logic
+for reasoning about the capability machine from \cite{skorstengaardESOP18} in the Iris framework.
 
-While Iris is a framework and supports many languages, it is geared towards models of higher-level languages.
-Embedding a language asks for a definition of expressions, values, and the state, which abstracts away from the fact that programs are stored in memory.
+While Iris is a framework and supports many languages, it is geared towards models of higher-level languages, which abstract from the fact that programs are stored in memory, and hence come equipped with notions of expressions and values, in addition to the program memory.
+In contrast, our low-level capability machine model has no notion of expression and values, it just consists of a memory and a register file.
+The program counter register contains a pointer to an address in memory.
+That address in turn will contain an integer, which can then be decoded to a machine instruction, such as Load, Store, Jump, etc.
+Once an instruction has been executed, the program counter is incremented and will then point to the next instruction in memory.
 
-Typically a machine will have a special register; the program counter, which contains a pointer to an address in memory.
-That address in turn will contain an integer, which can then be decoded to various machine dependent instructions, such as Load, Store, Jump, etc.
-Once an instruction has been executed, the program counter is incremented and will now point to the next instruction in memory.
+To capture the semantics of the capability machine in the Iris framework, we introduce an abstract notion of an instruction, whose operational meaning is to execute the instruction the program counter points to, and abstract notions of values, for halted and failed configurations.
 
-To capture the semantics of the Capability Machine in the Iris framework, we introduce an abstract notion of an instruction, whose operational meaning is to execute the instruction the program counter points to, and an abstract notion of values for halted and failed configurations.
+Next we use the Iris framework to \emph{prove} a Hoare triple for each instruction.
+As in \cite{Jensen:2013}, we define a points-to predicate for registers, denoted $r \mapsto_r w$.
+Since a capability machine replaces pointers with capabilities, we replace the conventional points-to predicate of separation logic for pointers with a points-to predicate with a permission attached to it, denoted
+$a \mapsto_a[p]~w$. 
+% \[
+%    \hspace{1cm} a \mapsto_a[p]~w.
+% \]
+This predicate states that address $a$ points to word $w$ with permission $p$.
+The permission restricts how the memory may be updated at address $a$.
+For instance, if $a \mapsto_a[\textsc{\tiny RO}]~w$, then the $\textsc{\tiny RO}$ (ReadOnly)
+permission only gives us permission to read from address $a$, but not to write to it.
 
-Next we use the Iris framework to prove Hoare triples for each instruction.
-Like in \cite{Jensen:2013}, we define a notion of points-to for registers.
-Since a capability machine replaces pointers with capabilities, we will also replace the conventional points-to for memory with a points-to assertion with a permission attached to it:
-$$ r \mapsto_r w \hspace{1cm} a \mapsto_a[p]~w $$
-The capability points-to predicate states that address $a$ points to word $w$ with permission $p$.
-The permission restricts the words an address can update to.
-For instance, in $a \mapsto_a[\textsc{\tiny RO}]~w$, the $\textsc{\tiny RO}$ permission does not allow for any updates to $a$.
-
-Similar to \cite{Jensen:2013}, the program logic rules for each instruction are of the following form: 
+% Similar to \cite{Jensen:2013}, the program logic rules for each instruction are of the following form: 
+The Hoare triples for basic instructions take the following form
 \begin{align*}
@@
 \end{align*} 
-Where the run time conditions correspond to the dynamic checks done by the capability machine, and \textsf{Instr Executable} is the abstract expression for executing the next instruction in memory.
-Unlike \cite{Jensen:2013}, the decoding function is in our case assumed.
-We now have the necessary building blocks for implementing the logical relation.
+Here, the run time conditions correspond to the dynamic checks done by the capability machine, and \textsf{Instr Executable} is the abstract expression for executing the next instruction in memory.
+This form of Hoare triple is similar to the one used in \cite{Jensen:2013} but
+unlike \cite{Jensen:2013}, the decoding function is in our case assumed.
 
+(Here we have only described the format for Hoare triples for individual instructions; we use a trick involving the standard bind rule of Iris to reason about programs consisting of many instructions, but we omit the description of that from this extended abstract.)
+
+% We now have the necessary building blocks for implementing the logical relation.
 \section{Logical Relation}
 
-\begin{figure*}[ht]
-	\begin{minipage}{\textwidth}
-	\begin{align*}
-		\mathcal{E}\interp{W}{pc} \triangleq&~\forall r, \mathcal{R}(W)(r) ~*~\textsf{context}(W)(r[\textsf{\small PC}:=pc])\\
-		&\sep~\textsc{WP}~\textsf{Seq (Instr Executable)}~\{ v, v = HaltedV \implies \exists W' r', W' \sqsubseteq_{priv} W * \textsf{context}(W')(r')\}
-	\end{align*}
-	\end{minipage}
-	\caption{Logical Relation for Expressions}
-	\label{fig:expr}
-\end{figure*}
+% \begin{figure*}[ht]
+% 	\begin{minipage}{\textwidth}
+% 	\begin{align*}
+% 		\mathcal{E}\interp{W}{pc} \triangleq&~\forall r, \mathcal{R}(W)(r) ~*~\textsf{context}(W)(r[\textsf{\small PC}:=pc])\\
+% 		&\sep~\textsc{WP}~\textsf{Seq (Instr Executable)}~\{ v, v = HaltedV \implies \exists W' r', W' \sqsubseteq_{priv} W * \textsf{context}(W')(r')\}
+% 	\end{align*}
+% 	\end{minipage}
+% 	\caption{Logical Relation for Expressions}
+% 	\label{fig:expr}
+% \end{figure*}
 
-\begin{figure*}[ht]
-	\begin{minipage}{\textwidth}
-	\begin{align*}
-		\mathcal{R}\interp{W}{r} \triangleq&~\forall (reg : \text{RegName} \backslash \textsf{PC}), \mathcal{V}\interp{W}{r(reg)}
-	\end{align*}
-	\end{minipage}
-	\caption{Logical Relation for Register States}
-	\label{fig:regs}
-\end{figure*}
+% \begin{figure*}[ht]
+% 	\begin{minipage}{\textwidth}
+% 	\begin{align*}
+% 		\mathcal{R}\interp{W}{r} \triangleq&~\forall (reg : \text{RegName} \backslash \textsf{PC}), \mathcal{V}\interp{W}{r(reg)}
+% 	\end{align*}
+% 	\end{minipage}
+% 	\caption{Logical Relation for Register States}
+% 	\label{fig:regs}
+% \end{figure*}
 
-\begin{figure*}[ht]
-	\begin{minipage}{\textwidth}
-	\begin{align*}
- 		\mathcal{V}\interp{W}{z} \triangleq&~\exists z' \in \mathds{Z}. z = z' \\
- 		\mathcal{V}\interp{W}{((\text{\tiny{O}},g),b,e,a)} \triangleq&~\top
- \\
- 		\left.
-		\begin{IEEEeqnarraybox}[\IEEEeqnarraystrutmode \IEEEeqnarraystrutsizeadd{2pt}{2pt}][c]{rCl} 
-			\mathcal{V}\interp{W}{((\text{\tiny{RO}},g),b,e,a)} \\
-			\mathcal{V}\interp{W}{((\text{\tiny{RW}},g),b,e,a)} \\
-			\mathcal{V}\interp{W}{((\text{\tiny{RWL}},g),b,e,a)}
-  \end{IEEEeqnarraybox}
-  \, \right\}  &\triangleq  \left.\exists~p', p \sqsubseteq p' * \textsf{read\_write\_cond}(\textsf{p',b,e})
-  \right. &\textit{where p is {\tiny{RO}}, {\tiny{RW}}, {\tiny{RWL}} resp.}
-  \\
-   		\left.
-		\begin{IEEEeqnarraybox}[\IEEEeqnarraystrutmode \IEEEeqnarraystrutsizeadd{2pt}{2pt}][c]{rCl} 
-			\mathcal{V}\interp{W}{((\text{\tiny{RX}},g),b,e,a)} \\
-			\mathcal{V}\interp{W}{((\text{\tiny{RWX}},g),b,e,a)} \\
-			\mathcal{V}\interp{W}{((\text{\tiny{RWLX}},g),b,e,a)}
-  \end{IEEEeqnarraybox}
-  \, \right\}  &\triangleq  \left.
-	\begin{IEEEeqnarraybox}[
-    \IEEEeqnarraystrutmode][c]{rCl}
-    \exists~p', p \sqsubseteq p' &*& \textsf{read\_write\_cond}(\textsf{p',b,e})\\
-  &*& ~\square~\textsf{exec\_cond(W)(\textsf{p,g,b,e})}
-  \end{IEEEeqnarraybox}  
-  \right. &\textit{where p is {\tiny{RX}}, {\tiny{RWX}}, {\tiny{RWLX}} resp.}
-  \\
- 		\mathcal{V}\interp{W}{((\text{\tiny{E}},g),b,e,a)} \triangleq&~\square~\textsf{enter\_cond}(W)(\textsf{g,b,e,a})
-	\end{align*}
-	\end{minipage}
-	\caption{Logical Relation for Words}
-	\label{fig:relation}
-\end{figure*}
+% \begin{figure*}[ht]
+% 	\begin{minipage}{\textwidth}
+% 	\begin{align*}
+%  		\mathcal{V}\interp{W}{z} \triangleq&~\exists z' \in \mathds{Z}. z = z' \\
+%  		\mathcal{V}\interp{W}{((\text{\tiny{O}},g),b,e,a)} \triangleq&~\top
+%  \\
+%  		\left.
+% 		\begin{IEEEeqnarraybox}[\IEEEeqnarraystrutmode \IEEEeqnarraystrutsizeadd{2pt}{2pt}][c]{rCl} 
+% 			\mathcal{V}\interp{W}{((\text{\tiny{RO}},g),b,e,a)} \\
+% 			\mathcal{V}\interp{W}{((\text{\tiny{RW}},g),b,e,a)} \\
+% 			\mathcal{V}\interp{W}{((\text{\tiny{RWL}},g),b,e,a)}
+%   \end{IEEEeqnarraybox}
+%   \, \right\}  &\triangleq  \left.\exists~p', p \sqsubseteq p' * \textsf{read\_write\_cond}(\textsf{p',b,e})
+%   \right. &\textit{where p is {\tiny{RO}}, {\tiny{RW}}, {\tiny{RWL}} resp.}
+%   \\
+%    		\left.
+% 		\begin{IEEEeqnarraybox}[\IEEEeqnarraystrutmode \IEEEeqnarraystrutsizeadd{2pt}{2pt}][c]{rCl} 
+% 			\mathcal{V}\interp{W}{((\text{\tiny{RX}},g),b,e,a)} \\
+% 			\mathcal{V}\interp{W}{((\text{\tiny{RWX}},g),b,e,a)} \\
+% 			\mathcal{V}\interp{W}{((\text{\tiny{RWLX}},g),b,e,a)}
+%   \end{IEEEeqnarraybox}
+%   \, \right\}  &\triangleq  \left.
+% 	\begin{IEEEeqnarraybox}[
+%     \IEEEeqnarraystrutmode][c]{rCl}
+%     \exists~p', p \sqsubseteq p' &*& \textsf{read\_write\_cond}(\textsf{p',b,e})\\
+%   &*& ~\square~\textsf{exec\_cond(W)(\textsf{p,g,b,e})}
+%   \end{IEEEeqnarraybox}  
+%   \right. &\textit{where p is {\tiny{RX}}, {\tiny{RWX}}, {\tiny{RWLX}} resp.}
+%   \\
+%  		\mathcal{V}\interp{W}{((\text{\tiny{E}},g),b,e,a)} \triangleq&~\square~\textsf{enter\_cond}(W)(\textsf{g,b,e,a})
+% 	\end{align*}
+% 	\end{minipage}
+% 	\caption{Logical Relation for Words}
+% 	\label{fig:relation}
+% \end{figure*}
 
-\begin{figure*}[ht]
-	\begin{minipage}{\textwidth}
-	\begin{align*}
- 		\textsf{read\_write\_cond}(p,b,e) \triangleq&~\bigsep_{a \in [b,e]} \text{rel}(a,p,\mathcal{V})\\
- 		\textsf{exec\_cond(W)(p,g,b,e)} \triangleq&~\forall a \in [b~e], W' \sqsubseteq^g W.~\rhd~\mathcal{E}\interp{W'}{((p,g),b,e,a)}\\
- 		\textsf{enter\_cond(W)(g,b,e,a)} \triangleq&~\forall W' \sqsubseteq^g W.~\rhd~\mathcal{E}\interp{W}{((\textsf{\tiny{RX}},g),b,e,a)}
-	\end{align*}
-	\end{minipage}
-	\\[2em]
-	where $W' \sqsubseteq^{\tiny Global} W \iff W' \sqsubseteq_{priv} W$\\ and 
-	$W' \sqsubseteq^{\tiny Local} W \iff W' \sqsubseteq_{pub} W$
+% \begin{figure*}[ht]
+% 	\begin{minipage}{\textwidth}
+% 	\begin{align*}
+%  		\textsf{read\_write\_cond}(p,b,e) \triangleq&~\bigsep_{a \in [b,e]} \text{rel}(a,p,\mathcal{V})\\
+%  		\textsf{exec\_cond(W)(p,g,b,e)} \triangleq&~\forall a \in [b~e], W' \sqsubseteq^g W.~\rhd~\mathcal{E}\interp{W'}{((p,g),b,e,a)}\\
+%  		\textsf{enter\_cond(W)(g,b,e,a)} \triangleq&~\forall W' \sqsubseteq^g W.~\rhd~\mathcal{E}\interp{W}{((\textsf{\tiny{RX}},g),b,e,a)}
+% 	\end{align*}
+% 	\end{minipage}
+% 	\\[2em]
+% 	where $W' \sqsubseteq^{\tiny Global} W \iff W' \sqsubseteq_{priv} W$\\ and 
+% 	$W' \sqsubseteq^{\tiny Local} W \iff W' \sqsubseteq_{pub} W$
 
-	\caption{Capability Conditions}
-	\label{fig:condition}
-\end{figure*}
+% 	\caption{Capability Conditions}
+% 	\label{fig:condition}
+% \end{figure*}
 
-\subsection{The Value Relation}
-%\begin{itemize}
+% \subsection{The Value Relation}
+% \begin{itemize}
 % 	\item definition of the relation and World
@@
 % 	\item now: standard islands are not used 
-%\end{itemize}
+% \end{itemize}
 
-The logical relation is a unary relation that relates a word to a notion of \textit{capability safety}.
-The value relation $\mathcal{V}$ is an Iris relation of type $World \to Word \to iProp$, where the World is a collection of state transition systems used to reason about local state.
+\newcommand{\World}{\mathrm{World}} \newcommand{\Word}{\mathrm{Word}} We now outline how we define a unary logical relation in Iris that captures capability safety.
+We define a value relation $\mathcal{V}$ as an Iris relation of type $\World \to \Word \to iProp$, where $\World$ is a collection of state transition systems used to reason about local state.
+(It is well-known how state transistion systems can be defined in Iris via Iris's notion of monoids and ghost state.)
 
-This notion of World is different from the logical relation given by \cite{skorstengaardESOP18}, where the World is a collection of invariants describing the behaviour of all memory.
+This notion of World is different from the logical relation given by \cite{skorstengaardESOP18}, where the World is a collection of invariants describing the behaviour of all memory, not just local state.
+We can use a simpler more abstract notion of world because XXX
+\todo{Lars: say why succintly, is it just because we use capability points-to predicates ?
+  But that could also have been done before\ldots}
+
+\todo{Lars: the following is too hard to read and understand -- I suggest to omit it and just add a succinct description above}
+\todo{begin}
 Skorstensgaard et.
@@
 We want the value relation to give us the points-to predicate, while asserting that what the predicate points to is valid. The challenge is now to define how that should be done.
-
-\subsection{Reasoning about Local State}
-%\begin{itemize}
+\todo{end}
+% \subsection{Reasoning about Local State}
+% \begin{itemize}
 % 	\item How to reason about local state in high level languages
@@
 % 	\item difficulties w/ iris invariants 
-%\end{itemize}
+% \end{itemize}
 
-Logical accounts of relations for high level languages with references have used Iris invariants to define the validity of a reference.
-In order to reason about local state, it suffices to add a future world relation only in the definition of the expression relation.
-This approach works in a high level language, where the calls are inherently well-bracketed.
+Logical accounts of logical relations for high level languages with references have used Iris invariants to define semantic validity of reference locations. This approach suffices for reasoning about 
+examples involving local state encapsulation in languages where all calls are well-bracketed.
 
-In our case however, calls are not always well-bracketed.
-\citeauthor{skorstengaardESOP18} distinguishes between well-bracketed and non well-bracketed calls by respectively using public- and private future world relations at the site of the callback.
+In our case, however, calls are not always well-bracketed (since the capability machine includes general jump instructions).
+\citeauthor{skorstengaardESOP18} distinguishes between well-bracketed and non well-bracketed calls by using notions of public- and private future worlds.
+Iris invariants do not make such a distinction and thus invariants alone are not sufficient for our purposes.
+Instead we explicitly define notions of public and private future world relations, and use these notions to distinguish between well-bracketed and non well-bracketed calls.
+We use these relations in combination with Iris' saved predicates, which allow us to save a predicate by associating it to a unique name.
+Using that unique name, we are then able to refer to the saved predicate somewhere else, and apply it to
+an appropriate future world argument.
 
-%Iris invariants do not make such a distinction.
-Using invariants alone is therefore not sufficient for our purposes.
-Instead we explicitly define notions of public and private future world relations, and use these notions to distinguish between well-bracketed and non well-bracketed calls.
+% A callback is an executable capability.
+% This distinction must therefore be done by the value relation.
+% However, if the value relation is indexed by a world, we again have difficulties using Iris invariants.
+% Asserting $\exists w, \mathcal{V}(W)(w)$ as invariant does not allow for $W$ to change.
 
-A callback is an executable capability.
-This distinction must therefore be done by the value relation.
-However, if the value relation is parametrized by a world, we have difficulties using Iris invariants.
-Asserting $\mathcal{V}(W)(w)$ as invariant does not allow for $W$ to change.
+% The difficulty is to support reasoning about local state, while distinguishing between public- and private future world relations in the value relation, rather than the expression relation.
+% We implement this by using Iris' saved predicates, which allows us to save a predicate by associating it to a unique name.
+% Using that unique name, we are then able to refer to that predicate somewhere else, and state it with the argument that we want.
+% That way, if we save the value relation predicate and associate it to a location, we can later on state that predicate and apply it to the future world in question.
 
-The difficulty lies in supporting reasoning about local state, while distinguishing between public- and private future world relations in the value relation, rather than the expression relation.
-We implement this by using Iris' saved predicates, which allows us to save a predicate by associating it to a unique name.
-Using that unique name, we are then able to refer to that predicate somewhere else, and state it with the argument that we want.
-That way, if we save the value relation predicate and associate it to a location, we can later on state that predicate and apply it to the future world in question.
+XXX Add that we prove FTLR and that the LR can be used to prove examples that rely
+on WBCF and LSE.
 

### 2019-10-17 prisc-2020/paper.tex

@@
 \todo{Lars: the following is too hard to read and understand -- I suggest to omit it and just add a succinct description above}
-\todo{begin}
-Skorstensgaard et.
-al.
-use the World to define validity of capabilities, by relating the invariant of the region in question to certain standard islands for capabilities.
+% \todo{begin}
+% Skorstensgaard et.
+% al.
+% use the World to define validity of capabilities, by relating the invariant of the region in question to certain standard islands for capabilities.
 
-In the logical account, we use the capability points-to assertion to both describe what we may get from the read permission of a capability, and the restrictions imposed on us by the write permission of a capability.
+% In the logical account, we use the capability points-to assertion to both describe what we may get from the read permission of a capability, and the restrictions imposed on us by the write permission of a capability.
 On the one hand, a points-to assertion lets us execute an instruction using that location.
@@
 
-We want the value relation to give us the points-to predicate, while asserting that what the predicate points to is valid. The challenge is now to define how that should be done.
-\todo{end}
+% We want the value relation to give us the points-to predicate, while asserting that what the predicate points to is valid. The challenge is now to define how that should be done.
+% \todo{end}
 % \subsection{Reasoning about Local State}

### 2019-10-18 prisc-2020/paper.tex
Some editing
- the PC is not incremented when we are jumping, so I replaced with "updated" instead

@@
   We present the first formalization of a capability machine that supports mechanized reasoning about deep semantic properties, including WBCF and LSE.
-  Our formalization is done in the Coq implementation of Iris, a state-of-the-art concurrent higher-order separation logic, and includes a formalization of the logical relation defined by Skorstensgaard et.
-al.
+  Our formalization is done in the Coq implementation of Iris, a state-of-the-art concurrent higher-order separation logic, and includes a formalization of the logical relation defined by Skorstensgaard et al.
 \cite{skorstengaardESOP18}, which can used to prove WBCF and LSE.
@@
 That address in turn will contain an integer, which can then be decoded to a machine instruction, such as Load, Store, Jump, etc.
-Once an instruction has been executed, the program counter is incremented and will then point to the next instruction in memory.
+Once an instruction has been executed, the program counter is updated and will then point to the next instruction in memory.
 
@@
 \small
-	\textit{run time conditions } \wedge &\textit{ decode}(w) = instr \Rightarrow  \\
+	\textit{run time conditions } \wedge &\textit{ decode}(w) = \emph{instr} \Rightarrow  \\
 	\{\{\{ \textsf{PC} \mapsto_r ((p,g),b,e,a) &* a \mapsto_a[p] w * ... \}\}\} \\ 
@@
 \newcommand{\World}{\mathrm{World}} \newcommand{\Word}{\mathrm{Word}} We now outline how we define a unary logical relation in Iris that captures capability safety.
-We define a value relation $\mathcal{V}$ as an Iris relation of type $\World \to \Word \to iProp$, where $\World$ is a collection of state transition systems used to reason about local state.
+We define a value relation $\mathcal{V}$ as an Iris relation of type $\World \to \Word \to \emph{iProp}$, where $\World$ is a collection of state transition systems used to reason about local state.
 (It is well-known how state transistion systems can be defined in Iris via Iris's notion of monoids and ghost state.)

### 2019-10-18 prisc-2020/paper.tex
added a final conclusion, changed from RO to RX, added a line about the size of the project in the introduction, only missing is to count the number of lines of code

--- 2019-10-17 prisc-2020/paper.tex
@@
 \end{itemize}
-All of the technical development is mechanized in Coq using the Coq implementation of Iris.
+Most of the technical development is mechanized in Coq using the Coq implementation of Iris (the remaining work consists of a few missing details).
+Currently the mechanization consists of XXX lines of code.
+Its substantial size can be slightly improved by some clever use of tactics, however such a mechanization will always be large, due to the low level nature of a machine: we want to consider every possible case of an instruction, both the fails and the successes.
 
-\todo{Alix: I think we need a short discussion (not specifically here) about the fact that we don't really have a theorem that says something like "being in the logical relation implies LSE and WBCF" as that's not really well defined, but we can prove things about specific programs that rely on it}
-\todo{Lars: I just tried to reduce the claim above}
+%\todo{Alix: I think we need a short discussion (not specifically here) about the fact that we don't really have a theorem that says something like "being in the logical relation implies LSE and WBCF" as that's not really well defined, but we can prove things about specific programs that rely on it}
+%\todo{Lars: I just tried to reduce the claim above}
 
@@
 The permission restricts how the memory may be updated at address $a$.
-For instance, if $a \mapsto_a[\textsc{\tiny RO}]~w$, then the $\textsc{\tiny RO}$ (ReadOnly)
-permission only gives us permission to read from address $a$, but not to write to it.
+For instance, if $a \mapsto_a[\textsc{\tiny RX}]~w$, then the $\textsc{\tiny RX}$ (ReadExecute)
+permission gives us permission to read from address $a$, to execute w, but it does not to permit us to write to address $a$.
 
@@
 This notion of World is different from the logical relation given by \cite{skorstengaardESOP18}, where the World is a collection of invariants describing the behaviour of all memory, not just local state.
-We can use a simpler more abstract notion of world because XXX
-\todo{Lars: say why succintly, is it just because we use capability points-to predicates ?
-  But that could also have been done before\ldots}
+We can use a simpler more abstract notion of world because the Iris model takes care of the world circularity problem.
+The necessary step indexing is abstracted away by the Iris saved predicates, which we use to define the validity of the regions a capability has authority over.
+%\todo{Lars: say why succintly, is it just because we use capability points-to predicates ?
+%  But that could also have been done before\ldots}
 
-\todo{Lars: the following is too hard to read and understand -- I suggest to omit it and just add a succinct description above}
+%\todo{Lars: the following is too hard to read and understand -- I suggest to omit it and just add a succinct description above}
 % \todo{begin}
@@
 
-XXX Add that we prove FTLR and that the LR can be used to prove examples that rely
-on WBCF and LSE.
+We prove the Fundamental Theorem of Logical Relations, in which we show that if we can read a region, we can execute it.
+We then use the fundamental theorem to prove examples with calls to unknown adversary code, which rely on local state encapsulation and well bracketed control flow.
 

### 2019-10-18 prisc-2020/paper.tex
Nitpicks

@@
 Most of the technical development is mechanized in Coq using the Coq implementation of Iris (the remaining work consists of a few missing details).
-Currently the mechanization consists of XXX lines of code.
+Currently the mechanization consists of 5.8k lines of spec and 15k lines of proof as reported by \textsf{coqwc}.
 Its substantial size can be slightly improved by some clever use of tactics, however such a mechanization will always be large, due to the low level nature of a machine: we want to consider every possible case of an instruction, both the fails and the successes.
+\todo{Alix: I would say it's more due to the nature of a capability machine with dynamic checks and the multiple ways they can fail}
 
@@
 
-Next we use the Iris framework to \emph{prove} a Hoare triple for each instruction.
+Next we use the Iris framework to \emph{prove} Hoare triples for each instruction.
 As in \cite{Jensen:2013}, we define a points-to predicate for registers, denoted $r \mapsto_r w$.
@@
 \end{align*} 
-Here, the run time conditions correspond to the dynamic checks done by the capability machine, and \textsf{Instr Executable} is the abstract expression for executing the next instruction in memory.
+Here, the runtime conditions correspond to the dynamic checks done by the capability machine, and \textsf{Instr Executable} is the abstract expression for executing the next instruction in memory.
 This form of Hoare triple is similar to the one used in \cite{Jensen:2013} but
@@
 In our case, however, calls are not always well-bracketed (since the capability machine includes general jump instructions).
-\citeauthor{skorstengaardESOP18} distinguishes between well-bracketed and non well-bracketed calls by using notions of public- and private future worlds.
+Skorstengaard et al.~\cite{skorstengaardESOP18} distinguish between well-bracketed and non well-bracketed calls by using notions of public- and private future worlds.
 Iris invariants do not make such a distinction and thus invariants alone are not sufficient for our purposes.

### 2019-10-18 prisc-2020/paper.tex
pass over sec 2 and 3

--- 2019-10-18 prisc-2020/paper.tex
@@
 For instance, if $a \mapsto_a[\textsc{\tiny RX}]~w$, then the $\textsc{\tiny RX}$ (ReadExecute)
-permission gives us permission to read from address $a$, to execute w, but it does not to permit us to write to address $a$.
+permission gives us permission to read from address $a$, to execute w, but it does not permit us to write to address $a$.
 
@@
 
-(Here we have only described the format for Hoare triples for individual instructions; we use a trick involving the standard bind rule of Iris to reason about programs consisting of many instructions, but we omit the description of that from this extended abstract.)
+Here we have only described the format for Hoare triples for individual instructions; we use a trick involving the standard bind rule of Iris to reason about programs consisting of many instructions, but we omit the description of that from this extended abstract.
 
@@
 \newcommand{\World}{\mathrm{World}} \newcommand{\Word}{\mathrm{Word}} We now outline how we define a unary logical relation in Iris that captures capability safety.
-We define a value relation $\mathcal{V}$ as an Iris relation of type $\World \to \Word \to \emph{iProp}$, where $\World$ is a collection of state transition systems used to reason about local state.
-(It is well-known how state transistion systems can be defined in Iris via Iris's notion of monoids and ghost state.)
+We define a value relation $\mathcal{V}$ as an Iris relation of type $\World \to \Word \to \emph{iProp}$, where $\World$ is a collection of state transition systems used to reason about local state and
+$\Word$ is the type of capability machine words. 
+It is well-known how state transistion systems can be defined in Iris via Iris' notion of monoids and ghost state \cite{iris}.
 
-This notion of World is different from the logical relation given by \cite{skorstengaardESOP18}, where the World is a collection of invariants describing the behaviour of all memory, not just local state.
-We can use a simpler more abstract notion of world because the Iris model takes care of the world circularity problem.
-The necessary step indexing is abstracted away by the Iris saved predicates, which we use to define the validity of the regions a capability has authority over.
+Our notion of World is simpler and more abstract than the 
+one used in the concrete logical relation given by \cite{skorstengaardESOP18}, where the World is a collection of invariants describing the behaviour of all memory, not just local state.
+We can use a simpler more abstract notion of world because the Iris model takes care of the world circularity problem. In particular, we make use of Iris' higher-order ghost state ---
+the ability to store arbitrary higher-order separation-logic predicates in ghost variables ---
+to define the validity of the regions a capability has authority over. The semantics of
+Iris' higher-order ghost state involves solving a recursive equation \cite{iris2}. 
 %\todo{Lars: say why succintly, is it just because we use capability points-to predicates ?
@@
 
-Logical accounts of logical relations for high level languages with references have used Iris invariants to define semantic validity of reference locations. This approach suffices for reasoning about 
-examples involving local state encapsulation in languages where all calls are well-bracketed.
+Logical accounts of logical relations for high level languages with references have used Iris invariants to define semantic validity of reference locations, see, e.g., \cite{ipm}.
+This approach suffices for reasoning about examples involving local state encapsulation in languages where all calls are well-bracketed.
 
 In our case, however, calls are not always well-bracketed (since the capability machine includes general jump instructions).
-\citeauthor{skorstengaardESOP18} distinguishes between well-bracketed and non well-bracketed calls by using notions of public- and private future worlds.
+\citeauthor{skorstengaardESOP18} distinguishes between well-bracketed and non well-bracketed calls by using notions of public- and private future worlds, following \cite{dreyer_neis_birkedal_2012}.
 Iris invariants do not make such a distinction and thus invariants alone are not sufficient for our purposes.
 Instead we explicitly define notions of public and private future world relations, and use these notions to distinguish between well-bracketed and non well-bracketed calls.
-We use these relations in combination with Iris' saved predicates, which allow us to save a predicate by associating it to a unique name.
+We use these relations in combination with Iris' higher-order ghost state, which allow us to save a predicate by associating it to a unique ghost name.
 Using that unique name, we are then able to refer to the saved predicate somewhere else, and apply it to
@@
 
-We prove the Fundamental Theorem of Logical Relations, in which we show that if we can read a region, we can execute it.
-We then use the fundamental theorem to prove examples with calls to unknown adversary code, which rely on local state encapsulation and well bracketed control flow.
+We prove the Fundamental Theorem of Logical Relations, which roughly says that if we have a read-execute capability, and it is capability safe to read it, then it is capability safe to execute it (use it as a program counter).
+We then use the fundamental theorem to prove functional correctness of examples with calls to unknown adversary code and whose correctness relies on local state encapsulation and well-bracketed control flow.
 

### 2019-10-18 prisc-2020/paper.tex
pass over sec 1

@@
 \end{itemize}
-Most of the technical development is mechanized in Coq using the Coq implementation of Iris (the remaining work consists of a few missing details).
-Currently the mechanization consists of 5.8k lines of spec and 15k lines of proof as reported by \textsf{coqwc}.
-Its substantial size can be slightly improved by some clever use of tactics, however such a mechanization will always be large, due to the low level nature of a machine: we want to consider every possible case of an instruction, both the fails and the successes.
-\todo{Alix: I would say it's more due to the nature of a capability machine with dynamic checks and the multiple ways they can fail}
+Currently, almost all of the technical development is mechanized in Coq using the Coq implementation of Iris.
+The mechanization presently consists of 5.8K lines of spec and 15K lines of proof as reported by \textsf{coqwc}.
+Its substantial size can probably be reduced by better use of tactics. However, such a mechanization will always be non-trivial due to the nature of the capability machine with dynamic checks
+and multiple ways in which instructions can fail.
 

### 2019-10-18 prisc-2020/paper.tex
removed appendix
future work

@@
 In our case, however, calls are not always well-bracketed (since the capability machine includes general jump instructions).
-\citeauthor{skorstengaardESOP18} distinguishes between well-bracketed and non well-bracketed calls by using notions of public- and private future worlds, following \cite{dreyer_neis_birkedal_2012}.
+\citeauthor{skorstengaardESOP18} \cite{skorstengaardESOP18} distinguishes between well-bracketed and non well-bracketed calls by using notions of public- and private future worlds, following \cite{dreyer_neis_birkedal_2012}.
 Iris invariants do not make such a distinction and thus invariants alone are not sufficient for our purposes.
-Instead we explicitly define notions of public and private future world relations, and use these notions to distinguish between well-bracketed and non well-bracketed calls.
+Instead we \textit{explicitly} define notions of public and private future world relations, rather than relying on Iris' \textit{implicit} future world, and use these notions to distinguish between well-bracketed and non well-bracketed calls.
 We use these relations in combination with Iris' higher-order ghost state, which allow us to save a predicate by associating it to a unique ghost name.
@@
 We then use the fundamental theorem to prove functional correctness of examples with calls to unknown adversary code and whose correctness relies on local state encapsulation and well-bracketed control flow.
+
+In the future we plan to finish the remaining details of the implementation, then use it as a starting point to prove full abstraction of a compiler from a high level language to our capability machine.
+Furthermore, it can also be used as a starting point for exploring different kinds of capabilities.
 
@@
 
-%% Appendix
-\appendix
-\section{Appendix}
+% %% Appendix
+% \appendix
+% \section{Appendix}
 
-Text of appendix \ldots
+% Text of appendix \ldots
 

### 2019-12-23 prisc-2020-talk/slides.tex
prisc 2020 presentation

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Beamer Presentation
% LaTeX Template
% Version 1.0 (10/11/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND THEMES
%----------------------------------------------------------------------------------------

\documentclass{beamer}

\mode<presentation> {

% The Beamer class comes with a number of default slide themes
% which change the colors and layouts of slides. Below this is a list
% of all the themes, uncomment each in turn to see what they look like.

\usetheme{default}
%\usetheme{AnnArbor}
%\usetheme{Antibes}
%\usetheme{Bergen}
%\usetheme{Berkeley}
%\usetheme{Berlin}
%\usetheme{Boadilla}
%\usetheme{CambridgeUS}
%\usetheme{Copenhagen}
%\usetheme{Darmstadt}
%\usetheme{Dresden}
%\usetheme{Frankfurt}
%\usetheme{Goettingen}
%\usetheme{Hannover}
%\usetheme{Ilmenau}
%\usetheme{JuanLesPins}
%\usetheme{Luebeck}
%\setheme{Madrid}
%\usetheme{Malmoe}
%\usetheme{Marburg}
%\usetheme{Montpellier}
%\usetheme{PaloAlto}
%\usetheme{Pittsburgh}
%\usetheme{Rochester}
%\usetheme{Singapore}
%\usetheme{Szeged}
%\usetheme{Warsaw}

% As well as themes, the Beamer class has a number of color themes
% for any slide theme. Uncomment each of these in turn to see how it
% changes the colors of your current slide theme.

%\usecolortheme{albatross}
%\usecolortheme{beaver}
%\usecolortheme{beetle}
%\usecolortheme{crane}
%\usecolortheme{dolphin}
%\usecolortheme{dove}
%\usecolortheme{fly}
%\usecolortheme{lily}
%\usecolortheme{orchid}
%\usecolortheme{rose}
%\usecolortheme{seagull}
\usecolortheme{seahorse}
%\usecolortheme{whale}
%\usecolortheme{wolverine}

%\setbeamertemplate{footline} % To remove the footer line in all slides uncomment this line
\setbeamertemplate{footline}[frame number] % To replace the footer line in all slides with a simple slide count uncomment this line

%\setbeamertemplate{navigation symbols}{} % To remove the navigation symbols from the bottom of all slides uncomment this line
}

\usepackage{graphicx} % Allows including images
\usepackage{booktabs} % Allows the use of \toprule, \midrule and \bottomrule in tables
\usepackage{tikz}
\usetikzlibrary{matrix}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{IEEEtrantools}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{dsfont}
\usepackage{longtable}
\usepackage{mathtools}

\newcommand{\sep}{-\kern-.6em\raisebox{-.659ex}{*}\ }
\newcommand{\bupd}[1]{=\kern-.6em\{#1\}\ \kern-.9em =\ \kern-.8em\raisebox{-.659ex}{*}\ }
\newcommand{\fupd}{=\kern-.3em=\ \kern-.8em\raisebox{-.659ex}{*}\ }
\newcommand{\nfupd}{=\kern-.6em \backslash\ \kern-.9em=\ \kern-.8em\raisebox{-.659ex}{*}\ }

\newcommand{\bigsep}{\mathop{\scalebox{2.5}{\raisebox{-0.4ex}{$*$}}}}%
\newcommand\myeq[1]{\stackrel{\mathclap{\normalfont\mbox{#1}}}{=}}

\newcommand{\ra}[1] {
  \begin{tikzpicture}
        \node[draw,dashed] {#1};
    \end{tikzpicture}}

\newcommand{\inv}[1] {
  \begin{tikzpicture}
        \node[draw] {#1};
    \end{tikzpicture}}

\newcommand{\interp}[2]{(#1)(#2)}
\newcommand\sepimp{\mathrel{-\mkern-6mu*}}

\usetikzlibrary{decorations.pathreplacing}

\newcommand\MemoryLayout[1]{
  \begin{tikzpicture}[scale=0.3]
     \draw[thick](0,0)--++(0,1);
     \foreach \pt/\col/\lab [remember=\pt as \tp (initially 0)] in {#1} {
     \if\lab\relax\relax\else
         \draw[thick,decorate, decoration={brace,amplitude=4mm}]
            (-\tp,-0.2)--node[below=4mm]{\lab} (-\pt+2,-0.2);
            \draw[fill=\col] (-\tp,0) rectangle (-\pt+2,1);
       \fi
       \foreach \a in {\tp,...,\pt} {
          \draw(-\a,0) rectangle ++(-1,1);
       }
       \draw[thick](-\pt,0)--++(0,1);

     }
     \node at (-4.5,0.6) {\small B};
  \end{tikzpicture}
}

\def\world{\emph{World}}
\def\word{\emph{Word}}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}

\lstset{style=mystyle,xleftmargin=.2\textwidth, xrightmargin=.2\textwidth}

\setbeamercovered{transparent}
\setbeamercolor{alerted text}{fg=red}

\AtBeginSection[]{
  \begin{frame}
  \vfill
  \centering
  \begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
    \usebeamerfont{title}\insertsectionhead\par%
  \end{beamercolorbox}
  \vfill
  \end{frame}
}

\AtBeginSubsection[]
{
    \begin{frame}
  	\vfill
  	\centering
  	\begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
    \usebeamerfont{title}\insertsubsectionhead\par%
  	\end{beamercolorbox}
  	\vfill
  \end{frame}
}
%----------------------------------------------------------------------------------------
%	TITLE PAGE
%----------------------------------------------------------------------------------------

\title[Capability Machine in Iris]{Implementing a Capability Machine model into Iris} % The short title appears at the bottom of every slide, the full title is only on the title page

\author{\textbf{A\"{i}na Linn Georges} \hspace{1cm} Alix Trieu \hspace{1cm} Lars Birkedal} % Your name
\institute[Aarhus University] % Your institution as it will appear on the bottom of every slide, may be shorthand to save space
{
Aarhus University \\ % Your institution for the title page
\medskip
\textit{ageorges@cs.au.dk} % Your email address
}
\date{\today} % Date, can be changed to a custom date

\begin{document}

\begin{frame}
\titlepage % Print the title page as the first slide
\end{frame}

\begin{frame}
\frametitle{Introduction}

\begin{itemize}
	\item<1> Capability machines allow for fine grained control over pointer permissions 
	\item<2> Good target for secure compilation
	\item<3> In particular: we are interested in enforcing certain higher level abstractions such as \textit{local state encapsulation} as \textit{well-bracketed control flow} at the lowest level of the machine
	\item<4> We need tools to reason about these subtle properties in a language that does not enforce them
	\item<5> These tools are elaborate and complex: we want to mechanize them, and facilitate the process of using them
\end{itemize}

\end{frame}

%----------------------------------------------------------------------------------------
%	PRESENTATION SLIDES
%----------------------------------------------------------------------------------------

\begin{frame}
\frametitle{Overview} % Table of contents slide, comment this block out to remove it
\tableofcontents[hideallsubsections]
 % Throughout your presentation, if you choose to use \section{} and \subsection{} commands, these will automatically be printed on this slide as an overview of your presentation
\end{frame}

%------------------------------------------------
\section{Capability Machines} % Sections can be created in order to organize your presentation into discrete blocks, all sections and subsections are automatically printed in the table of contents as an overview of the talk
%------------------------------------------------

\begin{frame}
\frametitle{Capability Machine}

\uncover<5->{\textbf{Capability: } An unforgeable token of authority
\\[1cm]}
\begin{columns}[c]

\column{.45\textwidth} % Left column and width
\begin{tikzpicture}
    \node<1> (img1) {\includegraphics[scale=0.7]{images/cap_1.pdf}};
    \node<2> (img2) {\includegraphics[scale=0.7]{images/cap_2.pdf}};
    \node<3> (img3) {\includegraphics[scale=0.7]{images/cap_3.pdf}};
    \node<4-> (img4) {\includegraphics[scale=0.7]{images/cap_4.pdf}};
\end{tikzpicture}

\column{.5\textwidth} % Right column and width
\uncover<4->{\begin{tikzpicture}

\matrix (a) [matrix of math nodes, ampersand replacement=\&, column sep=0.6cm, row sep=0.5cm]{
\& \textsc{\small RWX} \&\\
\textsc{\small RW}  \& \& \textsc{\small RX}  \\
\& \textsc{\small RO} \&\\};

\foreach \i/\j in {1-2/2-1, 1-2/2-3, 2-1/3-2, 2-3/3-2}
    \draw (a-\i) -- (a-\j);
%
\end{tikzpicture}}

\end{columns}
\end{frame}

%------------------------------------------------
\subsection{Enforcing Local Stack Encapsulation using Capabilities} % A subsection can be created just before a set of slides with a common theme to further break down your presentation into chunks

\begin{frame}[fragile]
\frametitle{Capability Safety}
\textbf{Local State Encapsulation}
\\[1cm]
\begin{columns}[c]

\column{.45\textwidth} % Left column and width
\begin{tikzpicture}
    \node<1> (img1) {\includegraphics[scale=0.7]{images/wbcf_1.pdf}};
    \node<2> (img2) {\includegraphics[scale=0.7]{images/wbcf_2.pdf}};
    \node<3> (img3) {\includegraphics[scale=0.7]{images/wbcf_3.pdf}};
\end{tikzpicture}

\column{.5\textwidth} % Right column and width
\begin{center}
\begin{lstlisting}
push r_stk 1
scall r 
pop r_stk r_1
assert r_1 1
halt
\end{lstlisting}
\end{center}

\end{columns}
\end{frame}

%------------------------------------------------
\subsection{Enforcing Well Bracketed Control Flow using Capabilities} % A subsection can be created just before a set of slides with a common theme to further break down your presentation into chunks

\begin{frame}[fragile]
\frametitle{Capability Safety}
\textbf{Well Bracketed Control Flow}
\\[1cm]
\begin{columns}[c]

\column{.45\textwidth} % Left column and width
\begin{tikzpicture}
    \node<1> (img1) {\includegraphics[scale=0.7]{images/wbcf_1.pdf}};
    \node<2> (img2) {\includegraphics[scale=0.7]{images/wbcf_2.pdf}};
    \node<3> (img3) {\includegraphics[scale=0.7]{images/wbcf_3.pdf}};
    \node<4> (img4) {\includegraphics[scale=0.7]{images/wbcf_4.pdf}};
    \node<5> (img5) {\includegraphics[scale=0.7]{images/wbcf_5.pdf}};
    \node<6> (img6) {\includegraphics[scale=0.7]{images/wbcf_6.pdf}};
\end{tikzpicture}

\column{.5\textwidth} % Right column and width
\begin{center}
\begin{lstlisting}
push r_stk 1
scall r
pop r_stk r_1
assert r_1 1
push r_stk 2
scall r 
halt
\end{lstlisting}
\end{center}

\end{columns}
\end{frame}

%------------------------------------------------
\subsection{Local Capabilities}
\begin{frame}
\frametitle{Local Capabilities}

\begin{center} (p,\textbf{Local},b,e,a) \hspace{2cm} (p,\textbf{Global},b,e,a) \end{center}

\uncover<2->{

\begin{tikzpicture}

\matrix (a) [matrix of math nodes, ampersand replacement=\&, column sep=0.6cm, row sep=0.5cm]{
\& \& \textsc{\small RWLX} \& \& \& \\
\& \textsc{\small RWX} \& \& \textsc{\small RWL} \& \& \\
\textsc{\small RX} \& \& \textsc{\small RW}  \&  \\
\& \textsc{\small RO} \& \\};

\foreach \i/\j in {1-3/2-2, 1-3/2-4, 2-2/3-1, 2-4/3-3, 2-2/3-3, 3-1/4-2, 3-3/4-2}
    \draw (a-\i) -- (a-\j);
%
\end{tikzpicture}
\begin{tikzpicture}

\matrix (a) [matrix of math nodes, ampersand replacement=\&, column sep=0.6cm, row sep=0.5cm]{
\textsc{\small GLOBAL}\\ \\ \\ \\
\textsc{\small LOCAL}\\};

\foreach \i/\j in {1-1/5-1}
    \draw (a-\i) -- (a-\j);
%
\end{tikzpicture}
}
\uncover<3>{
\begin{center} well-bracketed \hspace{2cm} not well-bracketed \end{center}}

\end{frame}

\begin{frame}
\frametitle{Calling Convention}
\begin{tikzpicture}
    \node<1> (img1) {\includegraphics[scale=0.7]{images/calling_1.pdf}};
    \node<2> (img2) {\includegraphics[scale=0.7]{images/calling_2.pdf}};
\end{tikzpicture}

\begin{longtable}[c]{|l|l|}
\hline
\texttt{r\_stk} & $(RWLX,Local,b,e,a)$ \\ \hline
\endfirsthead
%
\endhead
\end{longtable}

\end{frame}

%------------------------------------------------
\section{Reasoning about Capability Safety}

\begin{frame}
\frametitle{Expressing Capability Safety}

\begin{itemize}
	\item<2-> using a Program Logic 
	\item<3-> using a logical relation to capture invariants on the type system
	\item<4-> \textbf{using a logical relation on an \emph{untyped} (or \emph{uni-typed}) language to capture semantic properties of the language}
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Step-indexed Kripke Logical Relation}

$$ \mathcal{V}\uncover<2->{(W)}\triangleq~\{\uncover<5->{{\color{red}n}}, (RW, g, b, e, a) | \only<1-2>{\cdots} \only<3->{\exists r, W(r) \only<3-4>{=} \only<5->{\myeq{\small {\color{red} n}}} \iota_{[b,e]}} \} \cup \cdots $$

 \begin{itemize}
 	\item<4-> World-circularity problem
 		\begin{itemize}
 			\item<5-> {\color{red} Step indexing}
 		\end{itemize}
 	\item<6-> The world may evolve: we need future world relation
 		\begin{itemize}
 			\item<7-> Local capabilities are revoked whereas Global capabilities are not, \textit{the relation needs to model this distinction}:
 			$$ \sqsupseteq_{pub} \hspace{1 cm} and \hspace{1 cm} \sqsupseteq_{priv} $$
 		\end{itemize}
 \end{itemize}

\end{frame}

\begin{frame}
\frametitle{Expressing Capability Safety in Iris - an Iris primer}

\textbf{Iris:} Higher-order Concurrent Separation Logic Framework
\begin{itemize}
	\item<2-> Foundational
	\item<3-> Implemented in Coq -- equipped with an interactive proof mode
	\item<4-> Framework -- embed any language and its operational semantics into Iris
	\item<5-> Comes equipped with: 
		\begin{itemize}
			\item<5-> Invariants
			\item<6-> Ghost state
		\end{itemize}
\end{itemize}

\uncover<5->{We can take advantage of Iris' step-indexed model and invariants to mechanize step-indexed Kripke logical relations with recursive worlds in a succinct and elegant way.}

\end{frame}

\begin{frame}
\frametitle{Expressing Capability Safety in Iris}

\textbf{Roadmap}
	\begin{itemize}
		\item<2> embed the language into Iris
		\item<3> define a program logic by proving Hoare Triples
		\item<4> define the logical relation -- using Iris tools to solve the world circularity problem
		\item<5> prove the fundamental theorem of logical relations
		\item<6> use the logical relation to prove examples that rely on local state encapsulation and well-bracketed control flow with calls to unknown adversary
	\end{itemize}

\end{frame}

%------------------------------------------------
\section{Program Logic}
%------------------------------------------------

\begin{frame}
\frametitle{Abstract Instructions}

$$(reg,mem) \rightarrow (reg',mem')$$
%
\begin{itemize}
	\item<2-> \textsf{Instr Executable}
	\item<3-> \textsf{Instr Halted} \uncover<5->{$\rightarrow$ \textsf{HaltedV}}
	\item<4-> \textsf{Instr Failed} \uncover<5->{$\rightarrow$ \textsf{FailedV}}
\end{itemize}

\end{frame}

%------------------------------------------------
\subsection{A Capability Points-to Predicate}
\begin{frame}
\frametitle{Points-to Predicate with Permissions}

\begin{align*}
	a \mapsto_a[RWL] w \only<2->{&\fupd a \mapsto_a[RWL] ((p,Local),b,e,l) \\}
	\only<3->{&\fupd a \mapsto_a[RW] ((p,Local),b,e,l) \\}
	\only<4->{&\nfupd a \mapsto_a[RW] ((p',Local),b',e',l') \\}
\end{align*}

\end{frame}

%%------------------------------------------------
%\section{Proving Hoare Triples}
%
%%------------------------------------------------
%\subsection{Successful Execution}
%
%\begin{frame}
%\frametitle{Hoare Triples of the Program Logic: Success}
%
%\begin{align*}
%&decode(w) = \textsf{Load dst src} \\
%\wedge &\textsf{ isCorrectPC }((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc}) \\
%\wedge &\textsf{ readAllowed } p_{src} ~\wedge~  \textsf{withinBounds } (b_{src},e_{src},a_{src})\\ \\
%\{\{\{&~ \alert<2>{\textsf{PC} \mapsto_r ((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc})} * a_{pc} \mapsto_a[p_{pc}] w \\
%&* \alert<3>{dst \mapsto_r w_{dst}} * src \mapsto_r ((p_{src},g_{src}),b_{src},e_{src},a_{src}) \\
%&* a_{src} \mapsto_a[p_{src}] w_{src} \}\}\} \\
%&\textsf{Instr Executable} \\
%\{\{\{&~  \alert<2>{\textsf{PC} \mapsto_r ((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc} + 1)}
%			  * a_{pc} \mapsto_a[p_{pc}] w
%			  \\&* \alert<3>{dst \mapsto_r w_{src}}
%			  * src \mapsto_r ((p_{src},g_{src}),b_{src},e_{src},a_{src})
%			   \\&* a_{src} \mapsto_a[p_{src}] w_{src} \}\}\}
%\end{align*}
%
%\end{frame}
%
%%------------------------------------------------
%\subsection{Failed Execution}
%
%\begin{frame}
%\frametitle{Hoare Triples of the Program Logic: Failure}
%
%\begin{align*}
%&decode(w) = \textsf{Load dst src} \\
%\wedge &\textsf{ isCorrectPC }((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc}) \\
%\wedge &~\alert<2>{\neg \textsf{readAllowed } p_{src}} ~\vee~ \neg \textsf{withinBounds } (b_{src},e_{src},a_{src})\\
%\{\{\{&~\textsf{PC} \mapsto_r ((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc}) * a_{pc} \mapsto_a[p_{pc}] w \\
%&* src \mapsto_r ((p_{src},g_{src}),b_{src},e_{src},a_{src}) \}\}\} \\
%&\textsf{Instr Executable} \\
%\{\{\{&~ \alert<2>{\textsf{FailedV}}, \top \}\}\}
%\end{align*}
%
%\end{frame}

%------------------------------------------------
\section{A Unary Logical Relation for Reasoning about Semantic Properties of an Untyped Language}
%------------------------------------------------

%------------------------------------------------
\subsection{The Value Relation}

\begin{frame}[fragile]
\frametitle{The Value Relation}
\begin{center}
\textbf{A unary logical relation of an un-typed language}
\end{center}

$$\mathcal{V} : \only<6->{{\color{red}STS} \rightarrow~} 
			Word \rightarrow iProp~\Sigma $$
\\[1em]
\uncover<3->{\textbf{Challenge: }distinguish between Local and Global capabilities:
\begin{itemize}
	\item<4-> At the level of the value relation
	\item<5-> Model revocation
\end{itemize}
}
\uncover<6->{{\color{red} STS}: A collection of state transition systems}

\uncover<2->{
$$  \mathcal{V}\only<6->{{\color{red} (W)}}((\text{\tiny{RW}},g),b,e,a) \triangleq~ \bigsep_{a \in [b,e]} \fbox{$\exists w, a \mapsto_a [RW] w \sep \mathcal{V}\only<5->{{\color{red} (W)}}(w)$}$$}

\end{frame}

%------------------------------------------------
\subsection{The Execute Condition}

\begin{frame}
\frametitle{The Execute Condition}

\begin{equation*}
\textsf{exec\_cond(W)(\textsf{p,g,b,e})} \triangleq \begin{cases}
\forall a \in [b~e], W' \sqsubseteq_{pub} W.\\
~\rhd~\mathcal{E}\interp{W'}{((p,g),b,e,a)} &g = Local\\ \\
\forall a \in [b~e], W' \sqsubseteq_{priv} W.\\
~\rhd~\mathcal{E}\interp{W'}{((p,g),b,e,a)} &g = Global\\
\end{cases}
\end{equation*}

\end{frame}

%------------------------------------------------
\subsection{The Expression Relation}

\begin{frame}
\frametitle{The Expression Relation}

 	\begin{align*}
 		\mathcal{E}\interp{W}{pc} \triangleq&~\forall r, \mathcal{R}(W)(r) ~*~{\only<2->{\color{red}} \textsf{context}(W)(r[\textsf{\small PC}:=pc])}\\
 		&\sep~\textsc{WP}~\textsf{Seq (Instr Executable)}~\\
 		&\hspace{1cm}\{ v, v = HaltedV \implies \exists W' r', W' \sqsubseteq_{priv} W \\
 		&\hspace{1cm}* {\only<2->{\color{red}}\textsf{context}(W')(r')}\}
 	\end{align*}

\only<2->{
\begin{align*}
	{\color{red} \textsf{context}(W)(r) =~\only<2>{?}} &\only<3->{(\bigsep_{r_i \mapsto w \in r} r_i \mapsto_r w) \wedge \textsf{full\_map } r}\\
	\only<4->{&* \textsf{na\_inv } \gamma_{na} \top \\ }
	\only<5->{&* \textsf{sts\_full } W \\ }
	\only<6->{&* \textsf{region } W \\ }
\end{align*}
}

\end{frame}

%------------------------------------------------
\section{The Fundamental Theorem of Logical Relations}

\begin{frame}
\frametitle{The Fundamental Theorem of logical relations}
\begin{center}
If we can read a region, and every word in that region is safe, then we can safely execute it
\end{center}
\end{frame}

\begin{frame}
\begin{itemize}
	\item<2-> "If we can read a region" : $p = \textsc{\small RX} \vee p = \textsc{\small RWX} \vee p = \textsc{\small RWLX}$
	\item<3-> "and every word in that region is safe": $\textsf{read\_write\_cond }(p,b,e)$
	\item<4-> "then we can safely execute it": $\mathcal{E}\interp{W}{((p,g),b,e,a)}$
\end{itemize}

\begin{align*}
	&(p = \textsc{\small RX} \vee p = \textsc{\small RWX} \vee p = \textsc{\small RWLX}) \implies\\
	&\textsf{read\_write\_cond }(p,b,e) \implies \mathcal{E}\interp{W}{((p,g),b,e,a)}
\end{align*}
\end{frame}

%------------------------------------------------
\section{Reasoning about Unknown Code}

\begin{frame}[fragile]
\frametitle{Reasoning about Unknown Code}
\textbf{We use the fundamental theorem to reason about calls to an unknown adversary}

\begin{columns}[c]

\column{.45\textwidth} % Left column and width
\begin{tikzpicture}
    \node (img3) {\includegraphics[scale=0.7]{images/wbcf_3.pdf}};
\end{tikzpicture}

\column{.5\textwidth} % Right column and width
\begin{center}
\begin{lstlisting}
push r_stk 1
scall r 
pop r_stk r_1
assert r_1 1
halt
\end{lstlisting}
\end{center}

\end{columns}

\begin{align*}
 		\mathcal{E}\interp{W}{pc} \triangleq&~\forall r, \mathcal{R}(W)(r) ~*~{\textsf{context}(W)(r[\textsf{\small PC}:=pc])}\\
 		&\sep~\textsc{WP}~\textsf{Seq (Instr Executable)}~\\
 		&\hspace{1cm}\{ v, v = HaltedV \implies \exists W' r', W' \sqsubseteq_{priv} W \\
 		&\hspace{1cm}* {\textsf{context}(W')(r')}\}
 	\end{align*}

\end{frame}

%------------------------------------------------
%\section{Conclusion}
%------------------------------------------------

\begin{frame}
\frametitle{Conclusion}

\begin{itemize}
	\item Embed a capability machine into Iris
	\item Define its program logic 
	\item Mechanize a unary logical relation for an untyped capability machine language
	\item Prove the fundamental theorem of logical relations
	\item Reason about examples that rely on Local Stack Encapsulation and Well-Bracketed Control Flow with calls to an unknown adversary
\end{itemize}

\end{frame}

%------------------------------------------------

\begin{frame}
\frametitle{References}
\footnotesize{
\begin{thebibliography}{99} % Beamer does not support BibTeX so references must be inserted manually as below
\bibitem[Skorstensgaard, 2018]{p1} Lau Skorstengaard, 
Dominique Devriese, 
and Lars Birkedal (2018)
\newblock Reasoning About a Machine with Local Capabilities
\newblock ESOP \emph{Programming Languages and Systems} 475--501.

\bibitem[Dreyer, 2012]{p2} Derek Dreyer, Georg Neis, Lars Birkedal (2012)
\newblock The impact of higher-order state and control effects on local relational reasoning
\newblock \emph{Journal of Functional Programming} 22(4-5) 477--528.

\bibitem[Dreyer, 2011]{p3} Derek Dreyer, Amal Ahmed, Lars Birkedal (2011)
\newblock Logical Step-Indexed Logical Relations
\newblock \emph{LMCS} 7(2:16).

\end{thebibliography}
}
\end{frame}

%----------------------------------------------------------------------------------------

\end{document}

### 2020-01-01 prisc-2020-talk/slides.tex
added slides about challenges

@@
 			\item<6-> Ghost state
+			\item<7-> Always and Later Modalities
 		\end{itemize}
@@
 
-\uncover<5->{We can take advantage of Iris' step-indexed model and invariants to mechanize step-indexed Kripke logical relations with recursive worlds in a succinct and elegant way.}
+\uncover<8->{We can take advantage of Iris' step-indexed model and invariants to mechanize step-indexed Kripke logical relations with recursive worlds in a succinct and elegant way.}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Expressing Capability Safety in Iris - Challenges}
+
+\begin{itemize}
+	\item Region invariants: Iris invariants
+	\item Future world relation: frame preserving updates and world satisfaction 
+	\item Step indexing: later modality
+\end{itemize}
+
+\textbf{Challenges}
+\begin{itemize}
+	\item<2-> Iris was designed with more high level languages in mind, how do we embed a low level machine language into Iris
+	\item<3-> Iris abstracts away certain details we want to reason about directly
+	\item<4-> There is only one frame preserving update, we need to distinguish between two future world relations
+\end{itemize}
 

### 2020-01-11 prisc-2020-talk/slides.tex
altered the slides

@@
 \usepackage{mathtools}
+\usepackage{iris}
 
@@
 
-\newcommand{\ra}[1] {
-  \begin{tikzpicture}
-        \node[draw,dashed] {#1};
-    \end{tikzpicture}}
+%\newcommand{\ra}[1] {
+%  \begin{tikzpicture}
+%        \node[draw,dashed] {#1};
+%    \end{tikzpicture}}
 
@@
 
-\begin{frame}
-\frametitle{Overview} % Table of contents slide, comment this block out to remove it
-\tableofcontents[hideallsubsections]
- % Throughout your presentation, if you choose to use \section{} and \subsection{} commands, these will automatically be printed on this slide as an overview of your presentation
-\end{frame}
+%\begin{frame}
+%\frametitle{Overview} % Table of contents slide, comment this block out to remove it
+%\tableofcontents[hideallsubsections]
+% % Throughout your presentation, if you choose to use \section{} and \subsection{} commands, these will automatically be printed on this slide as an overview of your presentation
+%\end{frame}
 
@@
 %------------------------------------------------
-\subsection{Enforcing Local Stack Encapsulation using Capabilities} % A subsection can be created just before a set of slides with a common theme to further break down your presentation into chunks
-
-\begin{frame}[fragile]
-\frametitle{Capability Safety}
-\textbf{Local State Encapsulation}
-\\[1cm]
-\begin{columns}[c]
-
-\column{.45\textwidth} % Left column and width
-\begin{tikzpicture}
-    \node<1> (img1) {\includegraphics[scale=0.7]{images/wbcf_1.pdf}};
-    \node<2> (img2) {\includegraphics[scale=0.7]{images/wbcf_2.pdf}};
-    \node<3> (img3) {\includegraphics[scale=0.7]{images/wbcf_3.pdf}};
-\end{tikzpicture}
-
-\column{.5\textwidth} % Right column and width
-\begin{center}
-\begin{lstlisting}
-push r_stk 1
-scall r 
-pop r_stk r_1
-assert r_1 1
-halt
-\end{lstlisting}
-\end{center}
-
-\end{columns}
-\end{frame}
+%\subsection{Enforcing Local Stack Encapsulation using Capabilities} % A subsection can be created just before a set of slides with a common theme to further break down your presentation into chunks
+%
+%\begin{frame}[fragile]
+%\frametitle{Capability Safety}
+%\textbf{Local State Encapsulation}
+%\\[1cm]
+%\begin{columns}[c]
+%
+%\column{.45\textwidth} % Left column and width
+%\begin{tikzpicture}
+%    \node<1> (img1) {\includegraphics[scale=0.7]{images/wbcf_1.pdf}};
+%    \node<2> (img2) {\includegraphics[scale=0.7]{images/wbcf_2.pdf}};
+%    \node<3> (img3) {\includegraphics[scale=0.7]{images/wbcf_3.pdf}};
+%\end{tikzpicture}
+%
+%\column{.5\textwidth} % Right column and width
+%\begin{center}
+%\begin{lstlisting}
+%push r_stk 1
+%scall r 
+%pop r_stk r_1
+%assert r_1 1
+%halt
+%\end{lstlisting}
+%\end{center}
+%
+%
+%\end{columns}
+%\end{frame}
 
@@
 }
-\uncover<3>{
-\begin{center} well-bracketed \hspace{2cm} not well-bracketed \end{center}}
 
@@
 \end{tikzpicture}
-
-\begin{longtable}[c]{|l|l|}
-\hline
-\texttt{r\_stk} & $(RWLX,Local,b,e,a)$ \\ \hline
-\endfirsthead
 %
-\endhead
-\end{longtable}
+%\begin{longtable}[c]{|l|l|}
+%\hline
+%\texttt{r\_stk} & $(RWLX,Local,b,e,a)$ \\ \hline
+%\endfirsthead
+%%
+%\endhead
+%\end{longtable}
 
@@
 \begin{align*}
-	a \mapsto_a[RWL] w \only<2->{&\fupd a \mapsto_a[RWL] ((p,Local),b,e,l) \\}
-	\only<3->{&\fupd a \mapsto_a[RW] ((p,Local),b,e,l) \\}
-	\only<4->{&\nfupd a \mapsto_a[RW] ((p',Local),b',e',l') \\}
+	a \mapsto_a[RWL] w \only<2->{&\vs a \mapsto_a[RWL] ((p,Local),b,e,l) \\}
+	\only<3->{&\vs a \mapsto_a[RW] ((p,Local),b,e,l) \\}
+	\only<4->{&\not\vs a \mapsto_a[RW] ((p',Local),b',e',l') \\}
 \end{align*}
@@
 \uncover<2->{
-$$  \mathcal{V}\only<6->{{\color{red} (W)}}((\text{\tiny{RW}},g),b,e,a) \triangleq~ \bigsep_{a \in [b,e]} \fbox{$\exists w, a \mapsto_a [RW] w \sep \mathcal{V}\only<5->{{\color{red} (W)}}(w)$}$$}
+$$  \mathcal{V}\only<6->{{\color{red} (\Sigma)}}((\text{\tiny{RW}},g),b,e,a) \triangleq~ \bigsep_{a \in [b,e]} \fbox{$\exists w, a \mapsto_a [RW] w * \mathcal{V}\only<5->{{\color{red} (\Sigma)}}(w)$}$$}
 
@@
 \begin{equation*}
-\textsf{exec\_cond(W)(\textsf{p,g,b,e})} \triangleq \begin{cases}
-\forall a \in [b~e], W' \sqsubseteq_{pub} W.\\
-~\rhd~\mathcal{E}\interp{W'}{((p,g),b,e,a)} &g = Local\\ \\
-\forall a \in [b~e], W' \sqsubseteq_{priv} W.\\
-~\rhd~\mathcal{E}\interp{W'}{((p,g),b,e,a)} &g = Global\\
+\textsf{exec\_cond($\Sigma$)(\textsf{p,g,b,e})} \triangleq \begin{cases}
+\forall a \in [b~e], \Sigma ' \sqsupseteq_{pub} \Sigma.\\
+~\rhd~\mathcal{E}\interp{\Sigma '}{((p,g),b,e,a)} &g = Local\\ \\
+\forall a \in [b~e], \Sigma ' \sqsupseteq_{priv} \Sigma.\\
+~\rhd~\mathcal{E}\interp{\Sigma '}{((p,g),b,e,a)} &g = Global\\
 \end{cases}
@@
 %------------------------------------------------
+%\section{What's new/contributions}
+%------------------------------------------------
+
+\begin{frame}
+\frametitle{What's new}
+
+\textbf{What's new: capability machine viewpoint}
+
+\begin{itemize}
+	\item Mechanized formalization: currently $\sim 20,000$ lines of code
+	\item A higher level of abstraction
+\end{itemize}
+
+~\\[1em]
+\textbf{What's new: Iris formalization viewpoint}
+\begin{itemize}
+	\item Formalization of a machine level language, with no distinction between program and memory
+	\item Distinction between  well-bracketed and non well-bracketed calls: using public/private transitions
+\end{itemize}
+
+\end{frame}
+
+%------------------------------------------------
 %\section{Conclusion}

### 2020-01-11 prisc-2020-talk/slides.tex
moved the execute condition and expression relation to bonus slides

@@
 
-%------------------------------------------------
-\subsection{The Execute Condition}
-
 \begin{frame}
-\frametitle{The Execute Condition}
-
-\begin{equation*}
-\textsf{exec\_cond($\Sigma$)(\textsf{p,g,b,e})} \triangleq \begin{cases}
-\forall a \in [b~e], \Sigma ' \sqsupseteq_{pub} \Sigma.\\
-~\rhd~\mathcal{E}\interp{\Sigma '}{((p,g),b,e,a)} &g = Local\\ \\
-\forall a \in [b~e], \Sigma ' \sqsupseteq_{priv} \Sigma.\\
-~\rhd~\mathcal{E}\interp{\Sigma '}{((p,g),b,e,a)} &g = Global\\
-\end{cases}
-\end{equation*}
+\frametitle{From World to state transition system collection}
+On paper: 
+\begin{align*}
+	\text{Region} =~&\{Revoked\}~\uplus \\
+						     &\{Temporary\} \times {\color{red} \text{State} \times \text{Rels}} \\ 
+						     &\times (\text{State} \rightarrow (Wor \xrightarrow[\sqsupseteq_{pub}]{mon, ne} \text{UPred(MemSeg)}))~\uplus \\
+						     &\{Permanent\} \times {\color{red} \text{State} \times \text{Rels}}  \\ 
+						     &\times (\text{State} \rightarrow (Wor \xrightarrow[\sqsupseteq_{priv}]{mon, ne} \text{UPred(MemSeg)})) \\
+	\text{World} =~&\mathds{N} \rightharpoonup \text{Region}
+\end{align*}
+In the Iris mechanization, we use a collection of state transition systems: 
+$$\Sigma: \mathds{N} \rightharpoonup States \times \mathds{N} \rightharpoonup Rels $$
+\uncover<2->{The world circularity problem is now handled using Iris invariants and saved predicates.}
 
@@
 
-%------------------------------------------------
-\subsection{The Expression Relation}
+\begin{frame}
+\frametitle{Standard STS}
 
-\begin{frame}
-\frametitle{The Expression Relation}
+$$\Sigma: \mathds{N} \rightharpoonup States \times \mathds{N} \rightharpoonup Rels $$
 
- 	\begin{align*}
- 		\mathcal{E}\interp{W}{pc} \triangleq&~\forall r, \mathcal{R}(W)(r) ~*~{\only<2->{\color{red}} \textsf{context}(W)(r[\textsf{\small PC}:=pc])}\\
- 		&\sep~\textsc{WP}~\textsf{Seq (Instr Executable)}~\\
- 		&\hspace{1cm}\{ v, v = HaltedV \implies \exists W' r', W' \sqsubseteq_{priv} W \\
- 		&\hspace{1cm}* {\only<2->{\color{red}}\textsf{context}(W')(r')}\}
- 	\end{align*}
-
-\only<2->{
-\begin{align*}
-	{\color{red} \textsf{context}(W)(r) =~\only<2>{?}} &\only<3->{(\bigsep_{r_i \mapsto w \in r} r_i \mapsto_r w) \wedge \textsf{full\_map } r}\\
-	\only<4->{&* \textsf{na\_inv } \gamma_{na} \top \\ }
-	\only<5->{&* \textsf{sts\_full } W \\ }
-	\only<6->{&* \textsf{region } W \\ }
-\end{align*}
-}
-
-\end{frame}
-
-%------------------------------------------------
-\section{The Fundamental Theorem of Logical Relations}
-
-\begin{frame}
-\frametitle{The Fundamental Theorem of logical relations}
 \begin{center}
-If we can read a region, and every word in that region is safe, then we can safely execute it
+\begin{tikzpicture}[scale=0.2]
+\tikzstyle{every node}+=[inner sep=0pt]
+\draw [black] (17.8,-20.1) circle (3);
+\draw (17.8,-20.1) node {$temp$};
+\draw [black] (39.8,-20.1) circle (3);
+\draw (39.8,-20.1) node {$rev$};
+\draw [black] (29.2,-33) circle (3);
+\draw (29.2,-33) node {$perm$};
+\draw [black,dashed] (20.396,-18.602) arc (115.58004:64.41996:19.465);
+\fill [black] (37.2,-18.6) -- (36.7,-17.81) -- (36.27,-18.71);
+\draw [black] (37.061,-21.318) arc (-69.63899:-110.36101:23.742);
+\fill [black] (20.54,-21.32) -- (21.12,-22.07) -- (21.46,-21.13);
+\draw [black,dashed] (19.79,-22.35) -- (27.21,-30.75);
+\fill [black] (27.21,-30.75) -- (27.06,-29.82) -- (26.31,-30.48);
+\draw [black,dashed] (37.9,-22.42) -- (31.1,-30.68);
+\fill [black] (31.1,-30.68) -- (32,-30.38) -- (31.23,-29.75);
+\draw [black] (16.477,-17.42) arc (234:-54:2.25);
+\fill [black] (19.12,-17.42) -- (20,-17.07) -- (19.19,-16.48);
+\draw [black] (38.477,-17.42) arc (234:-54:2.25);
+\fill [black] (41.12,-17.42) -- (42,-17.07) -- (41.19,-16.48);
+\draw [black] (30.523,-35.68) arc (54:-234:2.25);
+\fill [black] (27.88,-35.68) -- (27,-36.03) -- (27.81,-36.62);
+\end{tikzpicture}
 \end{center}
-\end{frame}
-
-\begin{frame}
-\begin{itemize}
-	\item<2-> "If we can read a region" : $p = \textsc{\small RX} \vee p = \textsc{\small RWX} \vee p = \textsc{\small RWLX}$
-	\item<3-> "and every word in that region is safe": $\textsf{read\_write\_cond }(p,b,e)$
-	\item<4-> "then we can safely execute it": $\mathcal{E}\interp{W}{((p,g),b,e,a)}$
-\end{itemize}
-
-\begin{align*}
-	&(p = \textsc{\small RX} \vee p = \textsc{\small RWX} \vee p = \textsc{\small RWLX}) \implies\\
-	&\textsf{read\_write\_cond }(p,b,e) \implies \mathcal{E}\interp{W}{((p,g),b,e,a)}
-\end{align*}
-\end{frame}
-
-%------------------------------------------------
-\section{Reasoning about Unknown Code}
-
-\begin{frame}[fragile]
-\frametitle{Reasoning about Unknown Code}
-\textbf{We use the fundamental theorem to reason about calls to an unknown adversary}
-
-\begin{columns}[c]
-
-\column{.45\textwidth} % Left column and width
-\begin{tikzpicture}
-    \node (img3) {\includegraphics[scale=0.7]{images/wbcf_3.pdf}};
-\end{tikzpicture}
-
-\column{.5\textwidth} % Right column and width
-\begin{center}
-\begin{lstlisting}
-push r_stk 1
-scall r 
-pop r_stk r_1
-assert r_1 1
-halt
-\end{lstlisting}
-\end{center}
-
-\end{columns}
-
-\begin{align*}
- 		\mathcal{E}\interp{W}{pc} \triangleq&~\forall r, \mathcal{R}(W)(r) ~*~{\textsf{context}(W)(r[\textsf{\small PC}:=pc])}\\
- 		&\sep~\textsc{WP}~\textsf{Seq (Instr Executable)}~\\
- 		&\hspace{1cm}\{ v, v = HaltedV \implies \exists W' r', W' \sqsubseteq_{priv} W \\
- 		&\hspace{1cm}* {\textsf{context}(W')(r')}\}
- 	\end{align*}
 
@@
 \begin{itemize}
-	\item Mechanized formalization: currently $\sim 20,000$ lines of code
-	\item A higher level of abstraction
+	\item Mechanized formalization: currently $\sim 25 000$ lines of Iris code
+	\item At a higher level of abstraction
+	\begin{itemize}
+		\item<2-> Step index $\rightarrow$ later modality
+		\item<3-> World $\rightarrow$ collection of state transition systems
+	\end{itemize}
 \end{itemize}
 
-~\\[1em]
+~\\[0.5em]
 \textbf{What's new: Iris formalization viewpoint}
@@
 %----------------------------------------------------------------------------------------
+% Bonus Slides
+%------------------------------------------------
+\subsection{The Execute Condition}
+
+\begin{frame}
+\frametitle{The Execute Condition}
+
+\begin{equation*}
+\textsf{exec\_cond($\Sigma$)(\textsf{p,g,b,e})} \triangleq \begin{cases}
+\forall a \in [b~e], \Sigma ' \sqsupseteq_{pub} \Sigma.\\
+~\rhd~\mathcal{E}\interp{\Sigma '}{((p,g),b,e,a)} &g = Local\\ \\
+\forall a \in [b~e], \Sigma ' \sqsupseteq_{priv} \Sigma.\\
+~\rhd~\mathcal{E}\interp{\Sigma '}{((p,g),b,e,a)} &g = Global\\
+\end{cases}
+\end{equation*}
+
+\end{frame}
+
+%------------------------------------------------
+\subsection{The Expression Relation}
+
+\begin{frame}
+\frametitle{The Expression Relation}
+
+ 	\begin{align*}
+ 		\mathcal{E}\interp{\Sigma}{pc} \triangleq&~\forall r, \mathcal{R}(\Sigma)(r) ~*~{\only<2->{\color{red}} \textsf{context}(\Sigma)(r[\textsf{\small PC}:=pc])}\\
+ 		&\sep~\textsc{WP}~\textsf{Seq (Instr Executable)}~\\
+ 		&\hspace{1cm}\{ v, v = HaltedV \implies \exists \Sigma ' r', \Sigma ' \sqsupseteq_{priv} \Sigma \\
+ 		&\hspace{1cm}* {\only<2->{\color{red}}\textsf{context}(\Sigma ')(r')}\}
+ 	\end{align*}
+
+\only<2->{
+\begin{align*}
+	{\color{red} \textsf{context}(\Sigma)(r) =~\only<2>{?}} &\only<3->{(\bigsep_{r_i \mapsto w \in r} r_i \mapsto_r w) \wedge \textsf{full\_map } r}\\
+	\only<4->{&* \textsf{na\_inv } \gamma_{na} \top \\ }
+	\only<5->{&* \textsf{sts\_full } \Sigma \\ }
+	\only<6->{&* \textsf{region } \Sigma \\ }
+\end{align*}
+}
+
+\end{frame}
+
+%------------------------------------------------
+\section{The Fundamental Theorem of Logical Relations}
+
+\begin{frame}
+\frametitle{The Fundamental Theorem of logical relations}
+\begin{center}
+If we can read a region, and every word in that region is safe, then we can safely execute it
+\end{center}
+\end{frame}
+
+\begin{frame}
+\begin{itemize}
+	\item<2-> "If we can read a region" : $p = \textsc{\small RX} \vee p = \textsc{\small RWX} \vee p = \textsc{\small RWLX}$
+	\item<3-> "and every word in that region is safe": $\textsf{read\_write\_cond }(p,b,e)$
+	\item<4-> "then we can safely execute it": $\mathcal{E}\interp{\Sigma}{((p,g),b,e,a)}$
+\end{itemize}
+
+\begin{align*}
+	&(p = \textsc{\small RX} \vee p = \textsc{\small RWX} \vee p = \textsc{\small RWLX}) \implies\\
+	&\textsf{read\_write\_cond }(p,b,e) \implies \mathcal{E}\interp{\Sigma}{((p,g),b,e,a)}
+\end{align*}
+\end{frame}
+
+%------------------------------------------------
+\section{Reasoning about Unknown Code}
+
+\begin{frame}[fragile]
+\frametitle{Reasoning about Unknown Code}
+\textbf{We use the fundamental theorem to reason about calls to an unknown adversary}
+
+\begin{columns}[c]
+
+\column{.45\textwidth} % Left column and width
+\begin{tikzpicture}
+    \node (img3) {\includegraphics[scale=0.7]{images/wbcf_3.pdf}};
+\end{tikzpicture}
+
+\column{.5\textwidth} % Right column and width
+\begin{center}
+\begin{lstlisting}
+push r_stk 1
+scall r 
+pop r_stk r_1
+assert r_1 1
+halt
+\end{lstlisting}
+\end{center}
+
+\end{columns}
+
+\begin{align*}
+ 		\mathcal{E}\interp{\Sigma}{pc} \triangleq&~\forall r, \mathcal{R}(\Sigma)(r) ~*~{\textsf{context}(\Sigma)(r[\textsf{\small PC}:=pc])}\\
+ 		&\sep~\textsc{WP}~\textsf{Seq (Instr Executable)}~\\
+ 		&\hspace{1cm}\{ v, v = HaltedV \implies \exists \Sigma ' r', \Sigma ' \sqsupseteq_{priv} \Sigma \\
+ 		&\hspace{1cm}* {\textsf{context}(\Sigma ')(r')}\}
+ 	\end{align*}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Proving the Fundamental Theorem}
+
+\end{frame}
 

### 2020-01-13 prisc-2020-talk/slides.tex
progress on bonus slides that show the load case of the ftlr
finished Load case sketch in bonus slides

@@
 \usepackage{iris}
+\usepackage{arydshln}
+\usepackage[thinlines]{easytable}
 
@@
 \begin{itemize}
-	\item Formalization of a machine level language, with no distinction between program and memory
+	\item Formalization of a machine language, with no distinction between program and memory
 	\item Distinction between well-bracketed and non well-bracketed calls: using public/private transitions
@@
 
+{\small
+\begin{table}
+\begin{tabular}[t]{c}
+\parbox{8cm}{\begin{align}
+\only<4->{& decode(w) = \text{Load}~dst~src}
+\only<1-3>{\\ &\bigsep_{a \in [b,e]} \text{read\_write\_cond }a}
+\only<1-4>{\\ &\bigsep \mathcal{R}(r)}
+\only<5->{\\ &\bigsep \mathcal{V}(p,g,b,e,a)}
+\only<6->{\\ &\bigsep  \mathcal{V}(w_{src})}
+\end{align}} \\[1em]
+\hdashline
+\parbox{8cm}{\begin{align}
+\only<1-4>{&\bigsep_{reg \hookrightarrow w \in r\only<2->{\backslash \textsc{PC}}} reg \mapsto_r w}
+\only<5->{&\bigsep dst \mapsto_r \only<5-6>{w_{dst}} \only<7->{\color{red} w_{src}} \\
+				  &\bigsep src \mapsto_r (p,g,b,e,a)}
+\only<2->{\\ &\bigsep \textsc{PC} \mapsto_r (pc_g,pc_p,pc_b,pc_e,pc_a)}
+\only<3->{\\ &\bigsep pc_a \mapsto_a[pc_p] w}
+\only<6->{\\ &\bigsep a \mapsto_a[p] w_{src}}
+\end{align}} \\
+\hline
+\parbox{8cm}{\begin{align*}\textsc{WP}~\textsf{Seq (Instr Executable)}~
+ 		\{ v, &v = HaltedV \implies \\ &\exists \Sigma ' r',
+ 		\Sigma ' \sqsupseteq_{priv} \Sigma * {\textsf{context}(\Sigma ')(r')}\}
+ 		\end{align*}}
+\end{tabular}
+\end{table}
+}
+
 \end{frame}

### 2020-01-15 prisc-2020-talk/slides.tex
slide fixes

@@
 
+\textit{Reasoning about a Machine with Local Capabilities}: \cite{p1}
+
 \end{frame}
@@
     \node<3> (img3) {\includegraphics[scale=0.7]{images/cap_3.pdf}};
-    \node<4-> (img4) {\includegraphics[scale=0.7]{images/cap_4.pdf}};
+    \node<4-> (img4) {\includegraphics[scale=0.7]{images/cap_4'.pdf}};
 \end{tikzpicture}
@@
 \end{columns}
+\only<2-4>{
+\begin{itemize}
+	\item \only<2>{Range}
+			  \only<3>{Address}
+			  \only<4>{Permission}
+\end{itemize}}
+
 \end{frame}
@@
 %------------------------------------------------
-%\subsection{Enforcing Local Stack Encapsulation using Capabilities} % A subsection can be created just before a set of slides with a common theme to further break down your presentation into chunks
+%\subsection{Enforcing Local State Encapsulation using Capabilities} % A subsection can be created just before a set of slides with a common theme to further break down your presentation into chunks
 %
@@
 %------------------------------------------------
-\subsection{Enforcing Well Bracketed Control Flow using Capabilities} % A subsection can be created just before a set of slides with a common theme to further break down your presentation into chunks
+\subsection{Enforcing Well-Bracketed Control Flow using Capabilities} % A subsection can be created just before a set of slides with a common theme to further break down your presentation into chunks
 
@@
 \frametitle{Capability Safety}
-\textbf{Well Bracketed Control Flow}
+\textbf{Well-Bracketed Control Flow}
 \\[1cm]
@@
 \begin{tikzpicture}
-    \node<1> (img1) {\includegraphics[scale=0.7]{images/wbcf_1.pdf}};
-    \node<2> (img2) {\includegraphics[scale=0.7]{images/wbcf_2.pdf}};
-    \node<3> (img3) {\includegraphics[scale=0.7]{images/wbcf_3.pdf}};
-    \node<4> (img4) {\includegraphics[scale=0.7]{images/wbcf_4.pdf}};
-    \node<5> (img5) {\includegraphics[scale=0.7]{images/wbcf_5.pdf}};
-    \node<6> (img6) {\includegraphics[scale=0.7]{images/wbcf_6.pdf}};
+    \node<1> (img1) {\includegraphics[scale=0.7]{images/wbcf_1'.pdf}};
+    \node<2> (img2) {\includegraphics[scale=0.7]{images/wbcf_2'.pdf}};
+    \node<3> (img3) {\includegraphics[scale=0.7]{images/wbcf_3'.pdf}};
+    \node<4> (img4) {\includegraphics[scale=0.7]{images/wbcf_4'.pdf}};
+    \node<5> (img5) {\includegraphics[scale=0.7]{images/wbcf_5'.pdf}};
+    \node<6> (img6) {\includegraphics[scale=0.7]{images/wbcf_6'.pdf}};
 \end{tikzpicture}
@@
 \end{columns}
+
+\begin{itemize}
+	\item \only<1>{We start with a stack with range b to e}
+			  \only<2>{Push some local state}
+			  \only<3>{Prepare adversary stack}
+			  \only<4>{Adversary possesses a return capability}
+			  \only<5>{Once jumped to we get back original stack - we pop the stack and assert that local state did not change}
+			  \only<6>{Prepare the adversary stack for second call}
+\end{itemize}
 \end{frame}
@@
 \end{tikzpicture}
+\begin{itemize}
+	\item Local capabilities can only be stored where you have write local permission
+\end{itemize}
 }
@@
 %\end{longtable}
-
+\begin{itemize}
+	\item We want the adversary to lose any temporary capabilities (such as return capabilities) upon return of a function call
+\end{itemize}
 \end{frame}
@@
 \begin{itemize}
-	\item<2-> using a Program Logic 
+	\item<2-> using a program logic 
 	\item<3-> using a logical relation to capture invariants on the type system
@@
 		\item<2> embed the language into Iris
-		\item<3> define a program logic by proving Hoare Triples
+		\item<3> define a program logic by proving hoare triples
 		\item<4> define the logical relation -- using Iris tools to solve the world circularity problem
@@
 \end{frame}
-
-%------------------------------------------------
-\section{Program Logic}
-%------------------------------------------------
-
-\begin{frame}
-\frametitle{Abstract Instructions}
-
-$$(reg,mem) \rightarrow (reg',mem')$$
-%
-\begin{itemize}
-	\item<2-> \textsf{Instr Executable}
-	\item<3-> \textsf{Instr Halted} \uncover<5->{$\rightarrow$ \textsf{HaltedV}}
-	\item<4-> \textsf{Instr Failed} \uncover<5->{$\rightarrow$ \textsf{FailedV}}
-\end{itemize}
-
-\end{frame}
-
-%------------------------------------------------
-\subsection{A Capability Points-to Predicate}
-\begin{frame}
-\frametitle{Points-to Predicate with Permissions}
-
-\begin{align*}
-	a \mapsto_a[RWL] w \only<2->{&\vs a \mapsto_a[RWL] ((p,Local),b,e,l) \\}
-	\only<3->{&\vs a \mapsto_a[RW] ((p,Local),b,e,l) \\}
-	\only<4->{&\not\vs a \mapsto_a[RW] ((p',Local),b',e',l') \\}
-\end{align*}
-
-\end{frame}
-
-%%------------------------------------------------
-%\section{Proving Hoare Triples}
-%
-%%------------------------------------------------
-%\subsection{Successful Execution}
-%
-%\begin{frame}
-%\frametitle{Hoare Triples of the Program Logic: Success}
-%
-%\begin{align*}
-%&decode(w) = \textsf{Load dst src} \\
-%\wedge &\textsf{ isCorrectPC }((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc}) \\
-%\wedge &\textsf{ readAllowed } p_{src} ~\wedge~  \textsf{withinBounds } (b_{src},e_{src},a_{src})\\ \\
-%\{\{\{&~ \alert<2>{\textsf{PC} \mapsto_r ((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc})} * a_{pc} \mapsto_a[p_{pc}] w \\
-%&* \alert<3>{dst \mapsto_r w_{dst}} * src \mapsto_r ((p_{src},g_{src}),b_{src},e_{src},a_{src}) \\
-%&* a_{src} \mapsto_a[p_{src}] w_{src} \}\}\} \\
-%&\textsf{Instr Executable} \\
-%\{\{\{&~  \alert<2>{\textsf{PC} \mapsto_r ((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc} + 1)}
-%			  * a_{pc} \mapsto_a[p_{pc}] w
-%			  \\&* \alert<3>{dst \mapsto_r w_{src}}
-%			  * src \mapsto_r ((p_{src},g_{src}),b_{src},e_{src},a_{src})
-%			   \\&* a_{src} \mapsto_a[p_{src}] w_{src} \}\}\}
-%\end{align*}
-%
-%\end{frame}
-%
-%%------------------------------------------------
-%\subsection{Failed Execution}
-%
-%\begin{frame}
-%\frametitle{Hoare Triples of the Program Logic: Failure}
-%
-%\begin{align*}
-%&decode(w) = \textsf{Load dst src} \\
-%\wedge &\textsf{ isCorrectPC }((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc}) \\
-%\wedge &~\alert<2>{\neg \textsf{readAllowed } p_{src}} ~\vee~ \neg \textsf{withinBounds } (b_{src},e_{src},a_{src})\\
-%\{\{\{&~\textsf{PC} \mapsto_r ((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc}) * a_{pc} \mapsto_a[p_{pc}] w \\
-%&* src \mapsto_r ((p_{src},g_{src}),b_{src},e_{src},a_{src}) \}\}\} \\
-%&\textsf{Instr Executable} \\
-%\{\{\{&~ \alert<2>{\textsf{FailedV}}, \top \}\}\}
-%\end{align*}
-%
-%\end{frame}
 
@@
 
+\begin{itemize}
+	\item Dotted lines: private transitions
+	\item Continuous lines: public transitions
+\end{itemize}
+
 \end{frame}
@@
 
+%\begin{itemize}
+%	\item Embed a capability machine into Iris
+%	\item Define its program logic 
+%	\item Mechanize a unary logical relation for an untyped capability machine language
+%	\item Prove the fundamental theorem of logical relations
+%	\item Reason about examples that rely on Local State Encapsulation and Well-Bracketed Control Flow with calls to an unknown adversary
+%\end{itemize}
+
+\textbf{Contributions}
 \begin{itemize}
-	\item Embed a capability machine into Iris
-	\item Define its program logic 
-	\item Mechanize a unary logical relation for an untyped capability machine language
-	\item Prove the fundamental theorem of logical relations
-	\item Reason about examples that rely on Local Stack Encapsulation and Well-Bracketed Control Flow with calls to an unknown adversary
+	\item First mechanization of a core model of a capability machine. The mechanization includes: 
+		\begin{itemize}
+			\item Embedding of a capability machine language into Iris (first embedding of a machine language into Iris)
+			\item Mechanized proof of the fundamental theorem of logical relations 
+			\item Mechanized proof of capability safety of two non-trivial example programs 
+		\end{itemize}
+\end{itemize}
+\textbf{Future work}
+\begin{itemize}
+	\item Mechanize a proof of capability safety of the awkward example
+	\item Expand the mechanization with new capabilities and calling conventions that solve current shortcomings of local capabilities
 \end{itemize}
@@
 %------------------------------------------------
+%------------------------------------------------
+\section{Program Logic}
+%------------------------------------------------
+
+\begin{frame}
+\frametitle{Abstract Instructions}
+
+$$(reg,mem) \rightarrow (reg',mem')$$
+%
+\begin{itemize}
+	\item<2-> \textsf{Instr Executable}
+	\item<3-> \textsf{Instr Halted} \uncover<5->{$\rightarrow$ \textsf{HaltedV}}
+	\item<4-> \textsf{Instr Failed} \uncover<5->{$\rightarrow$ \textsf{FailedV}}
+\end{itemize}
+
+\end{frame}
+
+%------------------------------------------------
+\subsection{A Capability Points-to Predicate}
+\begin{frame}
+\frametitle{Points-to Predicate with Permissions}
+
+\begin{align*}
+	a \mapsto_a[RWL] w \only<2->{&\vs a \mapsto_a[RWL] ((p,Local),b,e,l) \\}
+	\only<3->{&\vs a \mapsto_a[RW] ((p,Local),b,e,l) \\}
+	\only<4->{&\not\vs a \mapsto_a[RW] ((p',Local),b',e',l') \\}
+\end{align*}
+
+\end{frame}
+
+%------------------------------------------------
+\section{Proving Hoare Triples}
+
+%------------------------------------------------
+\subsection{Successful Execution}
+
+\begin{frame}
+\frametitle{Hoare Triples of the Program Logic: Success}
+
+\begin{align*}
+&decode(w) = \textsf{Load dst src} \\
+\wedge &\textsf{ isCorrectPC }((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc}) \\
+\wedge &\textsf{ readAllowed } p_{src} ~\wedge~  \textsf{withinBounds } (b_{src},e_{src},a_{src})\\ \\
+\{\{\{&~ \alert<2>{\textsf{PC} \mapsto_r ((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc})} * a_{pc} \mapsto_a[p_{pc}] w \\
+&* \alert<3>{dst \mapsto_r w_{dst}} * src \mapsto_r ((p_{src},g_{src}),b_{src},e_{src},a_{src}) \\
+&* a_{src} \mapsto_a[p_{src}] w_{src} \}\}\} \\
+&\textsf{Instr Executable} \\
+\{\{\{&~  \alert<2>{\textsf{PC} \mapsto_r ((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc} + 1)}
+			  * a_{pc} \mapsto_a[p_{pc}] w
+			  \\&* \alert<3>{dst \mapsto_r w_{src}}
+			  * src \mapsto_r ((p_{src},g_{src}),b_{src},e_{src},a_{src})
+			   \\&* a_{src} \mapsto_a[p_{src}] w_{src} \}\}\}
+\end{align*}
+
+\end{frame}
+
+%------------------------------------------------
+\subsection{Failed Execution}
+
+\begin{frame}
+\frametitle{Hoare Triples of the Program Logic: Failure}
+
+\begin{align*}
+&decode(w) = \textsf{Load dst src} \\
+\wedge &\textsf{ isCorrectPC }((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc}) \\
+\wedge &~\alert<2>{\neg \textsf{readAllowed } p_{src}} ~\vee~ \neg \textsf{withinBounds } (b_{src},e_{src},a_{src})\\
+\{\{\{&~\textsf{PC} \mapsto_r ((p_{pc},g_{pc}),b_{pc},e_{pc},a_{pc}) * a_{pc} \mapsto_a[p_{pc}] w \\
+&* src \mapsto_r ((p_{src},g_{src}),b_{src},e_{src},a_{src}) \}\}\} \\
+&\textsf{Instr Executable} \\
+\{\{\{&~ \alert<2>{\textsf{FailedV}}, \top \}\}\}
+\end{align*}
+
+\end{frame}
+
 \subsection{The Execute Condition}
@@
 
+\begin{frame}
+\frametitle{Proving the Fundamental Theorem}
+
+{\small
+\begin{table}
+\begin{tabular}[t]{c}
+\parbox{8cm}{\begin{align}
+\only<4->{& decode(w) = \text{Load}~dst~src}
+\only<1-3>{\\ &\bigsep_{a \in [b,e]} \text{read\_write\_cond }a}
+\only<1-4>{\\ &\bigsep \mathcal{R}(r)}
+\only<5->{\\ &\bigsep \mathcal{V}(p,g,b,e,a)}
+\only<6->{\\ &\bigsep  \mathcal{V}(w_{src})}
+\end{align}} \\[1em]
+\hdashline
+\parbox{8cm}{\begin{align}
+\only<1-4>{&\bigsep_{reg \hookrightarrow w \in r\only<2->{\backslash \textsc{PC}}} reg \mapsto_r w}
+\only<5->{&\bigsep dst \mapsto_r \only<5-6>{w_{dst}} \only<7->{\color{red} w_{src}} \\
+				  &\bigsep src \mapsto_r (p,g,b,e,a)}
+\only<2->{\\ &\bigsep \textsc{PC} \mapsto_r (pc_g,pc_p,pc_b,pc_e,pc_a)}
+\only<3->{\\ &\bigsep pc_a \mapsto_a[pc_p] w}
+\only<6->{\\ &\bigsep a \mapsto_a[p] w_{src}}
+\end{align}} \\
+\hline
+\parbox{8cm}{\begin{align*}\textsc{WP}~\textsf{Seq (Instr Executable)}~
+ 		\{ v, &v = HaltedV \implies \\ &\exists \Sigma ' r',
+ 		\Sigma ' \sqsupseteq_{priv} \Sigma * {\textsf{context}(\Sigma ')(r')}\}
+ 		\end{align*}}
+\end{tabular}
+\end{table}
+}
+
+\end{frame}
+
 %------------------------------------------------
@@
 
-\begin{frame}
-\frametitle{Proving the Fundamental Theorem}
-
-{\small
-\begin{table}
-\begin{tabular}[t]{c}
-\parbox{8cm}{\begin{align}
-\only<4->{& decode(w) = \text{Load}~dst~src}
-\only<1-3>{\\ &\bigsep_{a \in [b,e]} \text{read\_write\_cond }a}
-\only<1-4>{\\ &\bigsep \mathcal{R}(r)}
-\only<5->{\\ &\bigsep \mathcal{V}(p,g,b,e,a)}
-\only<6->{\\ &\bigsep  \mathcal{V}(w_{src})}
-\end{align}} \\[1em]
-\hdashline
-\parbox{8cm}{\begin{align}
-\only<1-4>{&\bigsep_{reg \hookrightarrow w \in r\only<2->{\backslash \textsc{PC}}} reg \mapsto_r w}
-\only<5->{&\bigsep dst \mapsto_r \only<5-6>{w_{dst}} \only<7->{\color{red} w_{src}} \\
-				  &\bigsep src \mapsto_r (p,g,b,e,a)}
-\only<2->{\\ &\bigsep \textsc{PC} \mapsto_r (pc_g,pc_p,pc_b,pc_e,pc_a)}
-\only<3->{\\ &\bigsep pc_a \mapsto_a[pc_p] w}
-\only<6->{\\ &\bigsep a \mapsto_a[p] w_{src}}
-\end{align}} \\
-\hline
-\parbox{8cm}{\begin{align*}\textsc{WP}~\textsf{Seq (Instr Executable)}~
- 		\{ v, &v = HaltedV \implies \\ &\exists \Sigma ' r',
- 		\Sigma ' \sqsupseteq_{priv} \Sigma * {\textsf{context}(\Sigma ')(r')}\}
- 		\end{align*}}
-\end{tabular}
-\end{table}
-}
-
-\end{frame}
-
 \end{document}

### 2020-03-17 capability-meetings-notes/05-03/meeting-notes.tex

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}

\title{Capability Machine Meeting}

\begin{document}

\maketitle

\section*{Current Meeting}

Status report of mechanization:

\begin{itemize}
\item A looping version of the awkward example has been proven, which turned out not to exactly correspond to the original awkward example
\item Discussion on how we should build the WP rules on top of the semantics, and if we should model WP and semantics after each other or no
\item Most of the refactoring of the WP rules and the corresponding ftlr cases has been completed.
\item The ftlr case for Store, and rewriting parts of the WP for load and store in order to prove their concrete WP cases remain. 
\end{itemize}

Status report of IO-Drivers:

\begin{itemize}
\item A basic introduction to IO-related concepts of MMIO, Port-mapped IO, DMA, IOMMU
\item IO in Linux 2.5
\item Discussion on what model we could\textbackslash should aim for. The principles that were discussed in the above introduction were very general, and we should think about what things are easy\textbackslash hard to model in our current formalism, and given this set of features, look for a realistic architecture and corresponding driver we could perhaps base our implementation on. In this context, aRTOS, Contiki and CheriOS were mentioned.
\end{itemize}

\section*{Coming Week}

Work on the mechanization front during the coming week:

\begin{itemize}
\item A\"{i}na: finding a fix to convert the proof of the current awkward example to a proof of the classic awkward example.
\item Thomas, Alix, Arma\"el: finishing the refactoring of the WP cases. Thomas will write out the ftlr case for Store.
\item Alix will take a closer look at the WP rules for StoreU, LoadU and PromoteU in the uninitialized branch.
\item Thomas and Arma\"el will take a closer look at the systems that were mentioned (particularly CheriOS) and will write out a proposal annex spec.\ for the features our driver might want to support, its operational semantics and the corresponding statement of security.

\end{itemize}

\end{document}

### 2020-03-18 capability-meetings-notes/10-03/main.tex
(WIP) Meeting nodes for March 10

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{xspace}
\usepackage[dvipsnames]{xcolor}
\usepackage{iris}

\setlength{\parskip}{0.3em}

\newcommand{\Z}{\mathbb{Z}}
\newcommand{\X}[1]{\ensuremath{\textrm{#1}}}
\newcommand{\I}[1]{\ensuremath{\texttt{#1}}}
\newcommand{\SL}{Separation Logic\xspace}
\newcommand{\pure}[1]{\left\lceil #1 \right\rceil}

\newcommand{\FIXME}[1]{{\color{MidnightBlue} FIXME: #1}}

\newcommand{\MMIO}{\textlog{MMIO}}

\title{Capabilities, MMIO and Robust Safety}
\date{March 10, 2020}

\begin{document}

\maketitle

\section{Memory Mapped I/O: Operational Semantics}

The proposal is to simply represent read and writes to memory-mapped IO
addresses as events in a trace. This says nothing a priori about devices that
might be connected to these IO regions. In particular, without additional
assumptions, reading a byte from a memory-mapped region just returns an
arbitrary value.

If at some point we want to reason under the assumption that we are connected to
a specific device that reacts in a specific way, we can express that as an extra
\SL assertion, that we assume as a pre-condition, and that restricts the set of
different traces that we might observe.

\[
  \begin{array}{lcl}
    \X{EventTy} & := & \X{IOWrite} \; | \; \X{IORead} \\
    \X{Event} & := & \X{EventTy} \times \X{Addr} \times \Z \\
    \X{Trace} & := & \X{list} \; \X{Event} \\
    \X{State} & := & \underbrace{\X{Reg} \times \X{Mem}}_{\text{old state}}
                     \times \X{Trace} \\
  \end{array}
\]

Values of type \X{State} represent the state of a configuration in the
small-step operational semantics. In this setup, we assume the whole semantics
to be parameterized by the range of memory-mapped addresses: \MMIO.

\[
  \begin{array}{lcl}
    \MMIO & := & [\MMIO_{b}, \MMIO_{e}) \\
  \end{array}
\]

\textit{NB:} An alternative presentation would be to make this range an
immutable part of \X{State}. It probably does not really matter in the end, we
should do whatever makes our life easier for instantiating Iris with the
semantics.

In the (current) operational semantics without MMIO, the operational semantics
of the \I{Load} instruction is:

\begin{mathpar}
  \inferrule[Load]
  { r[\X{src}] = (p,g,b,e,a) \\ \X{readAllowed}\; p \\ a \in [b,e) }
  { (r, m) \xrightarrow{\I{Load} \; \X{dst} \; \X{src}} (r[\X{dst} := m[a]], m) }
\end{mathpar}

With MMIO, we obtain two rules:

\begin{mathpar}
  \inferrule[MemLoad]
  { r[\X{src}] = (p,g,b,e,a) \\ \X{readAllowed}\; p \\ a \in [b,e) \\ a \notin \MMIO}
  { (r, m, t) \xrightarrow{\I{Load} \; \X{dst} \; \X{src}}
    (r[\X{dst} := m[a]], m, t) }

  \inferrule[IOLoad]
  { r[\X{src}] = (p,g,b,e,a) \\ \X{readAllowed}\; p \\ a \in [b,e) \\ a \in \MMIO}
  { (r, m, t) \xrightarrow{\I{Load} \; \X{dst} \; \X{src}}
    (r[\X{dst} := x], m, (\X{IORead}, a, x) :: t) }
\end{mathpar}

Notice how in the second rule, we read an arbitrary integer $x$, and record it
in the trace.

\paragraph*{Remark:} With this presentation, the specification of the
``machine'' is very much decoupled from the model of the devices it might be
communicating with. This is a good thing, I think.

Nevertheless, one could consider an alternative presentation where the model of
the devices is more tightly integrated with the semantics of the machine. For
instance, one could make the operational semantics parameterized with the
devices' model, where each device is modeled as having some internal state, the
ability to react on reads or writes, or to perform an internal step. Then, the
operational semantics would either step the usual way, or whenever a device
steps.

I believe this would be somewhat similar to the semantics of I/O system calls
through a foreign function interfaces as formalized in
CakeML~\cite{cakeml-vstte17io}.

\paragraph*{Remark:} Alix says that the proposed style of operational semantics
(using a trace) is close to the semantics of \texttt{volatile} as in
CompCert---which is also how memory mapped addresses seem to be exposed to C
compilers in practice. So this is probably a good sign.

\section{\SL Resources}

We added a trace as part of the state, so we wish to also expose it as a \SL
assertion.

\newcommand{\tracefull}[1]{\ownGhost{\gamma_{T}}{\authfull{#1}}}
\newcommand{\tracefrag}[1]{\ownGhost{\gamma_{T}}{\authfrag{#1}}}

The simplest way to do that is to directly expose the complete trace. In that
case, the state interpretation additionally holds a resource for the trace:

\[
  \X{state\_interp} \; (r,m,t) := \ldots \ast \tracefull{t}
\]

The user then works with assertions of the form $\tracefrag{t'}$. Such an
assertion is not duplicable, and grants full ownership over the trace. In
particular, it allows one to \emph{update} the trace by emitting events, i.e. by
performing I/O operations.

The wp-rules for \I{Load} and \I{Store} need to be updated consequently. For
instance, the rule for a \I{Load} reading the integer $x$ on a MMIO address $a$
now requires $\tracefrag{t}$ in the pre-condition (for some $t$), and provides
$\tracefrag{(\X{IORead}, a, x) :: t}$ in the post-condition.

The corresponding resource algebra is $\authm(\exm(\X{Trace}))$. A few relevant
rules are:

\[
  \begin{array}{l}
    \tracefull{t} \ast \tracefrag{t'} \wand \pure{t = t'} \\
    \tracefrag{t} \ast \tracefrag{t'} \wand \FALSE \\
    \tracefull{t} \ast \tracefrag{t} \vsW \tracefull{t'} \ast \tracefrag{t'} \\
  \end{array}
\]

\paragraph*{Remark:} This is very coarse-grained: either one has the full
ownership for performing I/O and reasoning about it, or one cannot know anything
about it.

A first extension could be to allow observing prefixes of the trace (since
events can only be appended to the trace). The observation that the trace
has a given prefix would be duplicable.

Another extension, that seems very useful for modular reasoning, would be to
allow splitting the trace along separate range of addresses. Concretely, a trace
containing events about the range of MMIO addresses $[a,c)$ could be split into
two traces, granting ownership over events on addresses $[a,b)$ and $[b,c)$
respectively. Note that recombining these two traces would only yield some
unspecified interleaving of the events from the two traces, and not necessarily
yield the original trace.

One application of this second extension could be the verification of an example
involving ``multiplexed'' I/O, where two separate parts of the code are granted
separate ownership over separate MMIO addresses. These two separate pieces of
code would be verified separately; then, in the end, one could prove that one
gets \emph{some} interleaving of all the events emitted by both components.

\section{Toplevel Theorem (\emph{à la} OCPL)}

Let us start with a brief recap of the toplevel ``Robust Safety'' theorem for
OCPL itself (and its key ingredients), then the extension (by Thomas) of OCPL to
include a \I{print} capability, and finally move to the capability machine
setting.

\subsection{OCPL}

The {\sc RobustSafety} theorem of OCPL is as follows:

\begin{mathpar}
  \mprset{vskip=0.3em}
  \inferrule
  {C \in \textit{AdvCtx} \\
    e \; \X{closed} \\
    \hoare{\TRUE}{e}{x.\; \textlog{lowval}\; x} \\
    (C[e]);(\emptyset, \textlog{OK}) \longrightarrow^* T';(h',g')
  }
  {g' = \textlog{OK}}
\end{mathpar}

For any closed expression $e$, if $e$ has been verified to only return low
values, then running $e$ wrapped in an adversarial context $C$ from an initial
state, then we can observe that every reachable state is good
($g' = \textlog{OK}$ means that no assertion has failed in $e$).

$C \in AdvCtx$ means that $C$ cannot contain assertions (otherwise one could
trivially contradict the theorem by taking
$C[\cdot] = \textlog{assert false}$), and cannot contain references to raw
memory locations (otherwise one could directly access $e$'s private state,
invalidating the local state encapsulation mechanisms).

\newcommand{\lift}[2]{\textlog{lift}\; #1 \; #2}
\newcommand{\liftP}[1]{\lift{\Psi}{#1}}

$\textlog{lowval}\; x$ intuitively means that $x$ cannot be used to directly
access private (or ``high'') locations. It is formally defined using a logical
relation ``$\liftP{v}$'', which more generally asserts that the value $v$ only
gives direct access to locations that satisfy the predicate $\Psi$.

\[
  \begin{array}{lcl}
    \liftP (\textlog{rec}\; f\; x.\; e) & \eqdef
    & \later \forall v.\; \hoare{\liftP{v}}{e[v/x,\textlog{rec}\;f\;x.\; e/f]}{y.\; \liftP y} \\
    %
    \liftP (v_{1}, v_{2}) & \eqdef & \later (\liftP v_{1}, \liftP v_{2}) \\
    %
    \color{BrickRed} \liftP \ell & \color{BrickRed} \eqdef & \color{BrickRed} \Psi \; \ell \\
    %
    \ldots
  \end{array}
\]

Then, $\textlog{lowval}\; x$ is defined as $\lift{\textlog{lowloc}}{x}$, where
$\textlog{lowloc}$ is a predicate characterizing the region of memory containing
``low locations'' (distinct from the other region containing ``high
locations'').

\subsection{OCPL with \I{print}}

\newcommand{\OutV}{\textlog{Out}}

In Thomas' extension of OCPL, a new value $\OutV$ is added, denoting an
``output object capability'', as well as a $\textlog{print}$ primitive, where
$\textlog{print}\; \OutV\; v$ effectively ``prints'' the value $v$, i.e. adds it
to the trace of printed values.

One then wants to be able to encapsulate the use of $\OutV$, for instance by
defining object capabilities that enforce some invariants on the values being
printed. In that setting, $\OutV$ is considered as a ``high'' value:

\FIXME{Thomas, I realize that this is less general that what you have (I
  think?), since in your case $\OutV$ \emph{can} be low, if the trace predicate
  is equivalent to True. But I don't know how to state that in a setting where
  the predicate is not directly embedded in the definitions...}

\[
  \begin{array}{lcl}
    \liftP \OutV & \eqdef & \FALSE
  \end{array}
\]

The theorem then becomes:

\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  {C \in \textit{AdvCtx} \\
    e \; \X{closed} \\
    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)} \vdash \hoare{\TRUE}{e}{x.\; \textlog{lowval}\; x} \\
    (C[e]);(\emptyset, \textlog{OK}); \emptyset \longrightarrow^* T';(h',g'); t
  }
  {g' = \textlog{OK} \wedge P(t)}
\end{mathpar}

That is, if $e$ has been verified under the assumption that the predicate $P$
holds as a trace invariant, then executing $e$ in an adversarial context yields
a trace that does satisfy $P$.

$C \in \textit{AdvCtx}$ also has to be extended to forbid $C$ from containing
$\OutV$.

\subsection{Capability Machine with MMIO}

\newcommand{\VR}{\mathcal{V}}
\newcommand{\ER}{\mathcal{E}}
\newcommand{\RR}{\mathcal{R}}
\newcommand{\notMMIO}{\overline{\MMIO}}

Similarly to $\textlog{lift}$, we can generalize the existing logical relation
to thread a constraint on directly accessible memory locations. One would
parameterize the value, expression and register relations ($\VR$, $ER$ and $RR$)
with a predicate $\Psi$ on addresses.

Then, for any permission $p$ that includes either the R or W bit, $\VR$
is extended as follows:

\[
  \VR^{\Psi}(p, g, b, e, a) \eqdef \underbrace{\ldots}_{\text{as before}} \ast \; \pure{\forall a' \in [b,e).\; \Psi(a')}
\]

Then, $\ER^{\Psi}$ and $\RR^{\Psi}$ are simplify defined by threading the extra
$\Psi$ parameter through the existing definition.

Finally, for our relation to characterize ``low values'' that do not give direct
access to MMIO addresses, one would instantiate $\Psi$ with a predicate
$\notMMIO$ that excludes memory mapped addresses:

\[
  \notMMIO(a) \eqdef a \notin \MMIO
\]

Then, $\VR^{\notMMIO}$ is similar to the $\textlog{lowval}$ predicate of OCPL.
Intuitively, a value in the relation does not directly point to memory-mapped
locations, and also does not leak memory-mapped location through registers when
interacting with other code.

\paragraph*{Remark:} As stated above, the ``generalized'' value relation is in
fact not very useful for instantiation of $\Psi$ other than $\notMMIO$. Indeed,
even if $\Psi$ \emph{does} allow referring to addresses in the $\MMIO$ region,
the value relation does not grant any corresponding resources.
The fix would be to use the generalized trace resources mentioned previously,
and grant ownership for the part of the trace corresponding to addresses in
$[b,e) \cap \MMIO$.

Then, the toplevel theorem would be:

\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  { \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)} \vdash \VR^{\notMMIO}
    (E, g, b, e, a) \\
    (r[\X{PC} := (E, g, b, e, a)], m, \emptyset) \longrightarrow^*
    (r', m', t)
  }
  {P(t)}
\end{mathpar}

\FIXME{We need to add refine the assumptions of the theorem so that it has a
  chance of being true. One needs to add either a ``syntactic condition'' over
  thee state of the initial memory, or link with an initial boot-code and assume
  require the boot-code has a specification that entails the syntactic
  condition.}

\bibliographystyle{alpha}
\bibliography{biblio}

\appendix

\end{document}

### 2020-03-22 capability-meetings-notes/10-03/main.tex
Update Armaels meeting notes on the driver and robust safety
Small fixes
Concretization of top level theorems
Introduction of RA to describe MMIO

@@
 \usepackage{mathpartir}
+\usepackage{tensor}
 \usepackage{xspace}
@@
 \usepackage{iris}
+\usepackage{marvosym}
+\usepackage{xargs}
 
@@
 \newcommand{\Z}{\mathbb{Z}}
-\newcommand{\X}[1]{\ensuremath{\textrm{#1}}}
-\newcommand{\I}[1]{\ensuremath{\texttt{#1}}}
+\newcommand{\X}[1]{\ensuremath{\mathrm{#1}}}
+\newcommand{\V}[1]{\ensuremath{\mathit{#1}}}
+\newcommand{\I}[1]{\ensuremath{\mathtt{#1}}}
+\newcommand{\T}[1]{\texttt{#1}}
 \newcommand{\SL}{Separation Logic\xspace}
-\newcommand{\pure}[1]{\left\lceil #1 \right\rceil}
+\newcommand{\pure}[1]{\tensor[^{\ulcorner}]{#1{}}{^{\urcorner}}} %Hacky {} to avoid double superscript
 
@@
 \newcommand{\MMIO}{\textlog{MMIO}}
+
+\DeclareMathOperator{\initOKo}{init_{OK}}
+\newcommandx{\initOK}[4][3=r,4=m]{\initOKo(#1,#2,#3,#4)}
+\DeclareMathOperator{\initBCo}{init_{BC}}
+\newcommandx{\initBC}[4][3=r,4=m]{\initBCo(#1,#2,#3,#4)}
+
+\newcommand{\regdr}{\mathbin{!!}}
 
@@
   \begin{array}{lcl}
-    \MMIO & := & [\MMIO_{b}, \MMIO_{e}) \\
+    \MMIO & := & [\MMIO_{\X{b}}, \MMIO_{\X{e}}) \\
   \end{array}
@@
 
+The \I{Store} rule would be similar. Additionally, executing with a \X{PC}
+register pointing to an $\MMIO$ location should be disallowed in the operational
+semantics (I think this makes more sense than reading a random instruction
+through I/O).
+
 \paragraph*{Remark:} With this presentation, the specification of the
@@
 
-\newcommand{\tracefull}[1]{\ownGhost{\gamma_{T}}{\authfull{#1}}}
-\newcommand{\tracefrag}[1]{\ownGhost{\gamma_{T}}{\authfrag{#1}}}
+\newcommand{\tracefull}[1]{\ownGhost{\gamma_{\X{T}}}{\authfull{#1}}}
+\newcommand{\tracefrag}[1]{\ownGhost{\gamma_{\X{T}}}{\authfrag{#1}}}
 
@@
 events can only be appended to the trace). The observation that the trace
-has a given prefix would be duplicable.
+has a given prefix would be duplicable. This again seems to be an application of
+monotonicity, that could be realized by having a duplicable AtLeast part as part
+of $\tracefrag{t'}$,
+similarly to how we currently model Monotone References.
 
@@
 unspecified interleaving of the events from the two traces, and not necessarily
-yield the original trace.
+yield the original trace, since in our model, we cannot know the exact interleaving of IO operations issued by different parties.
 
@@
 gets \emph{some} interleaving of all the events emitted by both components.
+
+We use another resource algebra to model the MMIO locations. The operational
+semantics have been parameterized with a set $\MMIO$ of memory locations, but we
+need a way to specify what these locations are at a higher level of abstraction.
+A non-elegant solution is to index the WP definition with this set $\MMIO$. A
+better solution (that does not necessitate any changes to the definition of WP)
+is to add $\MMIO$ to the state, and add a conjunct to the state interpretation
+that expresses ownership of $\MMIO$ under agreement, i.e. add
+
+\newcommand{\MMIOag}{\ownGhost{\gamma_{\X{MMIO}}}{\agm (\X{MMIO})}}
+
+\[
+  \X{state\_interp} \; (r,m,t,\MMIO) := \ldots \ast \MMIOag
+\]
+
+to the state interpretation we had before.
+
+\paragraph*{Remark:} Another alternative, mentioned by Dominique, was to
+parameterize the entire language by the set of MMIO locations. This might come
+down to the same thing as the first solution mentioned above.
+
+In the future, we might be interested in the dynamic allocation of MMIO memory.
+The question is what this would mean, though, since the set of available devices
+allowing for MMIO access and the concrete buffers they provide will most likely
+still be modeled as fixed at runtime.
+Rather, this would allow us to take (un)mapped MMIO locations, connected to the different devices, and (un)map them anywhere in main memory.
+This would require a different way of modeling MMIO, where we need both a notion
+of the total pool of possible MMIO locations, and a description of the currently
+mapped locations. We could model this using an authoritative RA.
 
@@
 
-\FIXME{Thomas, I realize that this is less general that what you have (I
-  think?), since in your case $\OutV$ \emph{can} be low, if the trace predicate
-  is equivalent to True. But I don't know how to state that in a setting where
-  the predicate is not directly embedded in the definitions...}
-
 \[
@@
     \liftP \OutV & \eqdef & \FALSE
+  \end{array}
+\]
+
+To be more precise, we could actually allow sharing $\OutV$ in case the predicate P
+does not place any constraints on the output, i.e. is the $\TRUE$ predicate.
+Technically, we could hence have the following definition (although admittedly,
+it does look a bit hard-coded):
+\[
+  \begin{array}{lcl}
+    \liftP \OutV & \eqdef & \exists \iota \ldotp \knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast
+                            \pure{(\Lam{\_}.\TRUE) \spac t}}
   \end{array}
@@
     e \; \X{closed} \\
-    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)} \vdash \hoare{\TRUE}{e}{x.\; \textlog{lowval}\; x} \\
+    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast \pure{\! P(t)}} \vdash \hoare{\TRUE}{e}{x.\; \textlog{lowval}\; x} \\
     (C[e]);(\emptyset, \textlog{OK}); \emptyset \longrightarrow^* T';(h',g'); t
@@
 to thread a constraint on directly accessible memory locations. One would
-parameterize the value, expression and register relations ($\VR$, $ER$ and $RR$)
+parameterize the value, expression and register relations ($\VR$, $\ER$ and $\RR$)
 with a predicate $\Psi$ on addresses.
@@
 Intuitively, a value in the relation does not directly point to memory-mapped
-locations, and also does not leak memory-mapped location through registers when
-interacting with other code.
+locations, nor can it gain access to memory-mapped locations indirectly,
+similarly to how $\textlog{lowval}$s cannot grant access to high locations
+(in)directly. The relation $\ER^{\notMMIO}$ specifies that, if we fill the
+registers with values in $\VR^{\notMMIO}$, then the
+invariants in a private future world of our starting world will be satisfied at the end of execution.
 
@@
 The fix would be to use the generalized trace resources mentioned previously,
-and grant ownership for the part of the trace corresponding to addresses in
-$[b,e) \cap \MMIO$.
+and grant ownership for the part of the trace corresponding to the
+addresses in $[b,e) \cap \MMIO_{\X{pub}}$ with $\MMIO_{\X{pub}}$ the shareable subset of addresses in \MMIO.
 
-Then, the toplevel theorem would be:
+We will have to adapt the robust safety theorem to this new capability machine
+setting, \'and to the more concrete setup of the examples we want to gradually
+verify.
+
+Ignoring this second restriction for the time being, the general updated robust
+safety theorem might look something like this:
 
@@
   \inferrule
-  { \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)} \vdash \VR^{\notMMIO}
-    (E, g, b, e, a) \\
-    (r[\X{PC} := (E, g, b, e, a)], m, \emptyset) \longrightarrow^*
+  { \forall a_{i} \in A_{\X{entry}} \ldotp \left\{
+      {\begin{array}{ll}
+      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\! \ast \MMIOag
+         \vdash\\
+         \quad \forall W \ldotp\;(\ER^{\notMMIO} (RX, g, b, e, a_i))\spac W
+      \end{array}} \right. \\
+     \\
+    (r[\overline{r_{a_{i}} := (E, g, b, e, a_i)}], m, \emptyset) \longrightarrow^*
     (r', m', t)
@@
 
-\FIXME{We need to add refine the assumptions of the theorem so that it has a
-  chance of being true. One needs to add either a ``syntactic condition'' over
-  thee state of the initial memory, or link with an initial boot-code and assume
-  require the boot-code has a specification that entails the syntactic
-  condition.}
+The capability $(RX, g, b, e, a_i)$ points to the $i$th entry point of the code
+we want to gradually verify.
+The set of entry points is given by $A_{\X{entry}}$.
+The capability $(E, g, b, e, a_i)$ then represents
+a closure for this same piece of code. The adversary only gets access to the
+closure, in order to avoid them from executing or reading arbitrary lines of
+code within the trusted module.
+Notice the universal quantification over worlds in the precondition;
+this needs to be there, since we do not know in what world our code will be
+called.
+This universal quantification was not explicitly present in the OCPL
+formalization.
+The reason for that is that (I think so at least, do not quote me
+on this - Thomas) they
+used the built-in Iris worlds within their definition of weakest precondition,
+by leveraging the state interpretation, which is universally quantified over in
+the definition of weakest precondition.
+As we discussed before, if we do away
+with local capabilities, it should be possible to fall back onto Iris' notion of
+worlds, which would allow us to remove the universal quantification in the
+above definition too because it would be implicitly present in the WP inside $\ER$.
+
+\paragraph*{Remark:} it would be interesting to have a discussion about how to
+do away with the worlds if we remove local capabilities, regardless of whether
+we will in the end, since that might shed some light on why the worlds look the
+way they do right now.
+
+Also note that we might as well have written
+$\ldots \vdash \forall W \ldotp \; (\VR^{\notMMIO} \ldots $ %)
+in the above theorem, instead of having the expression relation, since this is
+the actual relation we will need, and easily unfolds to the former statement
+through the definition of the value relation. I wrote $\ER$ to place the focus
+on the execution of the driver being safe. I consider this a matter of taste.
+
+The above theorem is still lacking, since it has no notion of adversarial
+context \textit{AdvCtx} to limit the power of the attacker. Indeed, if we allow
+an attacker to have read or write access to any location in \MMIO, or possess
+any capability directly pointing into the driver, then all hope we might have of
+gradually enforcing safety properties is now lost. We need a check on the memory
+and registers that we will allow our driver closures to be used with.
+
+To enforce these syntactic restrictions, we have two options:
+\begin{enumerate}
+\item Add a syntactic check to the statement of our robust safety theorem,
+  ensuring that the adversary code cannot gain access to any non-E capabilities
+  pointing into the driver code or any capabilities pointing into $\MMIO$ directly.
+\item Assume the existence of a piece of boot code, that gets to run first when
+  the capability machine starts up, and makes sure the above condition actually
+  holds before passing control to the adversary. We could either assume that the
+  adversarial memory can contain any capability at startup, in which case the
+  driver has to erase the memory before passing control, or we could assume that
+  the memory cannot contain any capabilities whatsoever, in which case the boot
+  code does not have this obligation.
+\end{enumerate}
+
+If we were to follow option 2 and write concrete boot code, part of our proof
+would consist of reducing option 2 to the situation in option 1, since this is
+where unknown adversary code actually starts executing. In a way, option 1 is
+hence more fundamental than option 2. We therefore adapt the robust safety
+theorem to option 1 first.
+
+In this setting, we can subdivide the memory $m$ into 3 relevant subsections;
+$m = m_{\MMIO} \uplus m_{\X{driver}} \uplus m_{\X{adv}}$ %\uplus m_{\X{frame}}
+with the following meaning:
+\begin{itemize}
+\item $m_{\MMIO}$ contains the memory-mapped locations
+\item $m_{\X{driver}}$ contains the driver code
+\item  $m_{\X{adv}}$ contains the adversary's code and data (this will also
+  include the boot code in the general setting, since it does not contain any
+  inherently dangerous capabilities itself)
+\end{itemize}.
+
+\newcommand{\notMapsToR}[2]{\rightarrow^{\overline{#1}}\!\!(#2)}
+
+We can then implement the syntactic check using the following predicate
+$\notMapsToR{R}{w}$, that takes a word, memory region or register file and checks if it points into the forbidden region $R$ as follows:
+
+\[
+  \begin{array}{lcl}
+    \notMapsToR{R}{\X{inl}\spac z} & \eqdef
+    & \TRUE \\
+    %
+    \notMapsToR{R}{\X{inr}\spac (p,g,b,e,a)} & \eqdef & R \cap [b,e[\; =
+                                                 \emptyset  \\
+    %
+    \notMapsToR{R}{\V{reg}} & \eqdef & \forall r \in \dom(\V{reg}) \ldotp\spac \notMapsToR{R}{\V{reg}(r)} \\
+    %
+    \notMapsToR{R}{\V{mem}} & \eqdef & \forall l \in \dom(\V{mem}) \ldotp\spac   \notMapsToR{R}{\V{mem}(l)}
+  \end{array}
+\]
+
+Given this definition, we can now formalize the initialization constraints on
+registers $r$ and memory $m$, given the region $\MMIO$ and
+the driver's address space $[b,e[$ as follows;
+
+\begin{mathpar}
+  \mprset{vskip=0.4em}
+  \inferrule
+  { m = m_{\MMIO} \uplus m_{\X{driver}} \uplus m_{\X{adv}}\\
+    m_{\MMIO} = \MMIO \rightarrow \mbox{\Lightning} \\
+    \dom( m_{\X{driver}} ) = [b,e[\\
+    \dom( m_{\X{adv}} ) = [b_{\X{adv}},e_{\X{adv}}[\\
+    \notMapsToR{\MMIO \cup [b,e[}{\V{r}}\\
+    \notMapsToR{\MMIO \cup [b,e[}{\V{m_{\X{adv}}}} \\
+    r \regdr \X{PC} = (\X{RWX}, \X{G},b_{\X{adv}},e_{\X{adv}},b_{\X{adv}})\\
+  }
+  {\initOK{\MMIO}{[b,e[}}
+\end{mathpar}
+
+and subsequently formulate the robust safety theorem as follows:
+
+\begin{mathpar}
+  \mprset{vskip=0.4em}
+  \inferrule
+  { \forall a_{i} \in A_{\X{entry}} \ldotp \left\{
+      {\begin{array}{ll}
+      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\! \ast \MMIOag
+         \vdash\\
+         \quad \forall W \ldotp\;(\ER^{\notMMIO} (RX, g, b, e, a_i))\spac W
+       \end{array}} \right. \\
+   \initOK{\MMIO}{[b,e[} \\
+  (r[\overline{r_{a_{i}} := (E, g, b, e, a_i)}], m, \emptyset) \longrightarrow^* (r', m', t)\\
+  }
+  {P(t)}
+\end{mathpar}
+, where we have set the PC up to be a sensible capability (pointing to $b_{\X{adv}}$
+without loss of generality).
+
+Notice how the syntactic check for the registers is only performed on the set of registers $r$, allowing the closures in $\overline{r_{a_i}}$ to point into to the driver address space.
+
+Having this theorem in our toolbox, we can now take a closer look at scenario 2
+above, where a piece of boot code is used to set the memory up correctly,
+obviating the need for any other assumptions than that the boot code behaves
+according to some spec. Concretely, we want the boot code to uphold a spec that
+allows us to prove the above theorem.
+Alternatively, we could also just implement one specific instance of boot code, and prove that after it finishes executing, all preconditions to apply the above theorem are met.
+We can use the following predicate to specify that the starting configuration is
+valid:
+
+\begin{mathpar}
+  \mprset{vskip=0.4em}
+  \inferrule
+  { m = m_{\MMIO} \uplus m_{\X{driver}} \uplus m_{\X{boot}}\\
+    m_{\MMIO} = \MMIO \rightarrow \mbox{\Lightning} \\
+    \dom(m_{\X{driver}}) = [b,e[\\
+    \dom(m) = [0,\X{MEM_{MAX}}[\\
+    r \regdr \X{PC} = (\X{RWX}, \X{G},0,\X{MEM_{MAX}},l_{\X{boot}})\\
+  }
+  {\initBC{\MMIO}{[b,e[}}
+\end{mathpar}
+where $l_{\X{boot}}$ is the location where the machine boots, to be determined
+by the machines operational semantics.
+
+On the level of the operational semantics, we have to prove that the
+following holds, in order to use our concrete driver implementation with the
+robust safety theorem above (the driver still needs to be proven safe separately):
+
+\begin{mathpar}
+  \mprset{vskip=0.4em}
+  \inferrule
+  { \initBC{\MMIO}{[b,e[}\\
+  }
+  { \exists n\ldotp (r, m, \emptyset) \longrightarrow^n (r'[\overline{r_{a_{i}}
+      := (E, g, b, e, a_i)}], m', t) \land P(t) \land\\
+  \initOK{\MMIO}{[b,e[}[r'][m']}\quad (\textsc { BootCodeOK })
+\end{mathpar}
+
+Note that to link scenario 2 to scenario 1, we need a slightly adapted version of the robust safety theorem stated above
+where the initial trace is allowed to not be empty, but rather just satisfy
+$P(t)$, but that should not cause any problems.
+For a specific instance of boot code that provably satisfies the \textsc{BootCodeOK}
+rule, the resulting boot-code-based formulation of the robust safety theorem is now
+just a corollary of the general robust safety theorem, and has the following
+statement:
+
+\begin{mathpar}
+  \mprset{vskip=0.4em}
+  \inferrule
+  { \forall a_{i} \in A_{\X{entry}} \ldotp \left\{
+      {\begin{array}{ll}
+      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\! \ast \MMIOag
+         \vdash\\
+         \quad \forall W \ldotp\;(\ER^{\notMMIO} (RX, g, b, e, a_i))\spac W
+       \end{array}} \right. \\
+   \initBC{\MMIO}{[b,e[} \\
+  (r, m, \emptyset) \longrightarrow^* (r', m', t)\\
+  }
+  {P(t)}
+\end{mathpar}
+
+Now we have a closer look at the concrete example we would like to verify,
+including what its boot code looks like. Our goal is to prove that this concrete
+example satisfies \textsc{BootCodeOk} and this last robust safety theorem.
+
+The code for for the driver scenario we want to gradually verify can
+be found in \T{driver\_code.v}. This example file contains more details on the
+scenario in the comments. From a more high-level perspective, it contains a
+trusted part, consisting of a single MMIO location, boot code to properly set up
+the capability machine at start-up, and the code for the driver itself. On the
+other hand, the example also contains an untrusted part, consisting of a known,
+but untrusted code section and sandbox section (which we do not make assumptions
+over).
+The first address of the adversary code section is what the trusted part jumps
+to once it has finished setting up.
+
+The boot code is where execution starts off when the capability machine is
+powered on.
+It contains an omnipotent RWX capability (i.e. ranging over all of memory and
+providing full permissions over it) starts off execution.
+(\textbf{Design Alternative:} in the long run, it might be worthwhile to provide
+the boot code with a RWLX capability. Since RWX cannot be upgraded to RWLX, it
+is impossible in the current setting to develop a secure stack calling
+convention \'a la Lau, where the stack capability has to be local-WL and (in this
+case) the driver code has to ensure that there is no other WL than the stack.
+However, having a global stack capability is currently not an issue, since the interface to our driver is first order, i.e. it will
+never create another adversary stack frame on top of itself (currently, it does
+not even use the stack in the first place!), and since we are working in a
+single-threaded setting (not very relevant, but I think the multi-threaded
+setting is an interesting thought experiment). I avoided unnecessary
+complications and hence kept the omnipotent capability as RWX. If the omnipotent
+capability were to become RWLX in the future, we would have to be careful using
+a stack, since if we want our secure calling convention with a higher order
+interface to our driver API, there cannot be any adversary-accessible global
+write-local memory, i.e.
+the reduced omnipotent capability cannot be allowed to be global if we pass it to the adversary.
+We would thus have to pass a stack capability that we explicitly make local, and a global RWX capability for the rest of memory)
+
+The boot code has the following responsibilities:
+\begin{itemize}
+\item Generate the necessary closures (i.e.
+Enter capabilities) for the driver's read and write methods, so that the adversary can use safely use them to perform I/O, and any properties that the driver wants to uphold on the input-\textbackslash output stream can actually be enforced.
+\item Wipes all of the adversary's sandbox section, to make sure no remaining
+  capabilities can be found there (\textbf{Design Alternative:} maybe we should try to find out how CHERI handles this; is it possible that any lingering capabilities are left in RAM
+  on machine start-up? In any case, we are already being slightly unrealistic by assuming the adversary code section is just \emph{there}. We have the option
+  to not do any erasure, and just make the assumption that adversary memory
+  contains no capabilities pointing into MMIO or pointing into the driver. ).
+  It does n\'ot, however, wipe the untrusted, hard-coded adversary routine that reads $N$ lines of code from a single (and the only) MMIO location in memory.
+\item Reduces the omnipotent capability to provide RWX access to all of the
+  adversary's code, makes it point to the first element of the adversary's code
+  section and jumps to it, in order to hand control to the adversary. Before the
+  jump, it clears all registers except for the 2 containing the read and write
+  driver closures, in order to make sure that no extra permissions leak to the adversary.
+\end{itemize}
 

### 2020-03-24 capability-meetings-notes/10-03/main.tex
MMIO meeting notes: a few improvements / cleanups

@@
 
-\newcommand{\MMIO}{\textlog{MMIO}}
+\newcommand{\MMIO}{\textlog{MMIO}\xspace}
 
 \DeclareMathOperator{\initOKo}{init_{OK}}
-\newcommandx{\initOK}[4][3=r,4=m]{\initOKo(#1,#2,#3,#4)}
+\newcommandx{\initOK}[3][2=r,3=m]{\initOKo(#1,#2,#3)}
 \DeclareMathOperator{\initBCo}{init_{BC}}
-\newcommandx{\initBC}[4][3=r,4=m]{\initBCo(#1,#2,#3,#4)}
+\newcommandx{\initBC}[2][1=r,2=m]{\initBCo(#1,#2)}
 
-\newcommand{\regdr}{\mathbin{!!}}
+\newcommand{\app}{\mathbin{+\!\!+}}
+
+\newenvironment{remark}
+{ \bigskip\hrule\vspace{-1.3em}\nobreak
+  \paragraph*{Remark:}}
+{\vspace*{0.5em}\hrule\medskip}
 
@@
 
-\textit{NB:} An alternative presentation would be to make this range an
-immutable part of \X{State}. It probably does not really matter in the end, we
-should do whatever makes our life easier for instantiating Iris with the
-semantics.
-
 In the (current) operational semantics without MMIO, the operational semantics
@@
 
-With MMIO, we obtain two rules:
+With MMIO, we obtain two rules for \I{Load}:
 
@@
   { (r, m, t) \xrightarrow{\I{Load} \; \X{dst} \; \X{src}}
-    (r[\X{dst} := x], m, (\X{IORead}, a, x) :: t) }
+    (r[\X{dst} := x], m, t \app (\X{IORead}, a, x)) }
 \end{mathpar}
@@
 
-The \I{Store} rule would be similar. Additionally, executing with a \X{PC}
-register pointing to an $\MMIO$ location should be disallowed in the operational
-semantics (I think this makes more sense than reading a random instruction
-through I/O).
+The \I{Store} rule would be similar.
 
-\paragraph*{Remark:} With this presentation, the specification of the
+Notice that, for any address in \MMIO, which value is held by the map $m$ as
+that address is now irrelevant. Indeed, $m$ is never read nor modified for
+addresses in \MMIO. One could choose to enforce that a particular dummy value is
+stored in $m$ for these addresses. Instead, we choose to leave them
+unconstrained.
+
+\begin{remark}
+  Thomas: Additionally, we might want to disallow in the semantics an execution
+  with a \X{PC} register pointing to an $\MMIO$ location (I think this makes
+  more sense than reading a random instruction through I/O). Armaël: but it
+  makes the semantics of all instructions more complicated.
+\end{remark}
+
+\begin{remark} With this presentation, the specification of the
 ``machine'' is very much decoupled from the model of the devices it might be
@@
 through a foreign function interfaces as formalized in
-CakeML~\cite{cakeml-vstte17io}.
+CakeML~\cite{cakeml-vstte17io}, and also in Perennial~\cite{perennial-lang}.
+%
 
-\paragraph*{Remark:} Alix says that the proposed style of operational semantics
+\end{remark}
+
+\begin{remark} Alix says that the proposed style of operational semantics
 (using a trace) is close to the semantics of \texttt{volatile} as in
@@
 compilers in practice. So this is probably a good sign. 
+\end{remark}
 
@@
 We added a trace as part of the state, so we wish to also expose it as a \SL
-assertion.
+assertion. Recall the current definition of the state interpretation:
+\[
+  \X{state\_interp} \; (r,m) := \X{gen\_heap\_ctx}\; r \ast \X{gen\_heap\_ctx}\; m
+\]
 
@@
 
-The simplest way to do that is to directly expose the complete trace. In that
-case, the state interpretation additionally holds a resource for the trace:
-
+The simplest way to account for the trace is to directly expose it in a
+monolithic fashion. In that case, the state interpretation additionally holds a
+resource for the trace.
 \[
-  \X{state\_interp} \; (r,m,t) := \ldots \ast \tracefull{t}
+  \X{state\_interp} \; (r,m,t) :=
+    \X{gen\_heap\_ctx}\; r \ast
+    \X{gen\_heap\_ctx}\; (m \setminus \MMIO) \ast
+    {\color{BrickRed}\tracefull{t}}
 \]
 
-The user then works with assertions of the form $\tracefrag{t'}$. Such an
-assertion is not duplicable, and grants full ownership over the trace. In
-particular, it allows one to \emph{update} the trace by emitting events, i.e. by
-performing I/O operations.
+Note that we restrict the usual ``points-to'' assertions to be used only for
+non-MMIO addresses. Instead, to assert ownership of the MMIO region, the user
+works with assertions of the form $\tracefrag{t'}$. Such an assertion is not
+duplicable, and grants full ownership over the trace. In particular, it allows
+one to \emph{update} the trace by emitting events, i.e. by performing I/O
+operations.
 
@@
 
-\paragraph*{Remark:} This is very coarse-grained: either one has the full
+\begin{remark} This is very coarse-grained: either one has the full
 ownership for performing I/O and reasoning about it, or one cannot know anything
@@
 gets \emph{some} interleaving of all the events emitted by both components.
+\end{remark}
 
-We use another resource algebra to model the MMIO locations. The operational
-semantics have been parameterized with a set $\MMIO$ of memory locations, but we
-need a way to specify what these locations are at a higher level of abstraction.
-A non-elegant solution is to index the WP definition with this set $\MMIO$. A
-better solution (that does not necessitate any changes to the definition of WP)
-is to add $\MMIO$ to the state, and add a conjunct to the state interpretation
-that expresses ownership of $\MMIO$ under agreement, i.e. add
+\newcommand{\MMIOag}{\ownGhost{\gamma_{\X{MMIO}}}{\authfull (\X{MMIO})}}
 
-\newcommand{\MMIOag}{\ownGhost{\gamma_{\X{MMIO}}}{\agm (\X{MMIO})}}
-
+\begin{remark}
+  As an alternative to having the set of MMIO addresses (\MMIO) as a global
+  parameter, we might want to make it part of the state in the operational
+  semantics. In that case, we would need another resource algebra to model the
+  MMIO locations, i.e. add do the state interpretation we had before:
+%
 \[
@@
 
-to the state interpretation we had before.
+(where $\authm(\exm(\X{Trace}))$ is the resource algebra).
+\end{remark}
 
-\paragraph*{Remark:} Another alternative, mentioned by Dominique, was to
-parameterize the entire language by the set of MMIO locations. This might come
-down to the same thing as the first solution mentioned above.
-
+\begin{remark}
 In the future, we might be interested in the dynamic allocation of MMIO memory.
@@
 mapped locations. We could model this using an authoritative RA.
+\end{remark}
 
@@
 printed. In that setting, $\OutV$ is considered as a ``high'' value:
-
+%
 \[
@@
 
-To be more precise, we could actually allow sharing $\OutV$ in case the predicate P
-does not place any constraints on the output, i.e. is the $\TRUE$ predicate.
-Technically, we could hence have the following definition (although admittedly,
-it does look a bit hard-coded):
-\[
-  \begin{array}{lcl}
-    \liftP \OutV & \eqdef & \exists \iota \ldotp \knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast
-                            \pure{(\Lam{\_}.\TRUE) \spac t}}
-  \end{array}
-\]
-
 The theorem then becomes:
-
+%
 \begin{mathpar}
@@
 
+\medskip
+
+To be slightly more general, we could actually allow sharing $\OutV$ in case the
+predicate $P$ does not place any constraints on the output, i.e. it is the
+$\TRUE$ predicate. Technically, we could hence have the following alternative
+definition for $\textlog{lift}$ (although admittedly, it does look a bit
+hard-coded):
+%
+\[
+  \begin{array}{lcl}
+    \liftP \OutV & \eqdef & \exists \iota \ldotp \knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast
+                            \pure{(\Lam{\_}.\TRUE) \spac t}}
+  \end{array}
+\]
+
 \subsection{Capability Machine with MMIO}
+
+\subsubsection{Logical relation}
 
@@
 similarly to how $\textlog{lowval}$s cannot grant access to high locations
-(in)directly. The relation $\ER^{\notMMIO}$ specifies that, if we fill the
-registers with values in $\VR^{\notMMIO}$, then the
-invariants in a private future world of our starting world will be satisfied at the end of execution.
+directly. The relation $\ER^{\notMMIO}$ specifies that, if we fill the registers
+with values in $\VR^{\notMMIO}$, then the invariants in a private future world
+of our starting world will be satisfied at the end of execution.
 
-\paragraph*{Remark:} As stated above, the ``generalized'' value relation is in
-fact not very useful for instantiation of $\Psi$ other than $\notMMIO$. Indeed,
-even if $\Psi$ \emph{does} allow referring to addresses in the $\MMIO$ region,
-the value relation does not grant any corresponding resources.
-The fix would be to use the generalized trace resources mentioned previously,
-and grant ownership for the part of the trace corresponding to the
-addresses in $[b,e) \cap \MMIO_{\X{pub}}$ with $\MMIO_{\X{pub}}$ the shareable subset of addresses in \MMIO.
+\begin{remark} As it is defined above, the ``generalized'' value relation
+is in fact not very useful for instantiation of $\Psi$ other than $\notMMIO$.
+Indeed, even if $\Psi$ \emph{does} allow referring to addresses in the $\MMIO$
+region, the value relation does not grant any corresponding resources. The fix
+would be to use the generalized trace resources mentioned previously, and grant
+ownership for the part of the trace corresponding to the addresses in
+$[b,e) \cap \MMIO_{\X{pub}}$ with $\MMIO_{\X{pub}}$ the shareable subset of
+addresses in \MMIO.
+\end{remark}
 
-We will have to adapt the robust safety theorem to this new capability machine
-setting, \'and to the more concrete setup of the examples we want to gradually
-verify.
+\subsubsection{Robust safety theorem}
 
-Ignoring this second restriction for the time being, the general updated robust
-safety theorem might look something like this:
+A first attempt at adapting the OCPL robust safety theorem to the capability
+machine setting might look something like the following
+{\color{BrickRed}\textbf{(incomplete!)}} statement.
 
 \begin{mathpar}
-  \mprset{vskip=0.4em}
+  \mprset{vskip=0.5em}
   \inferrule
-  { \forall a_{i} \in A_{\X{entry}} \ldotp \left\{
-      {\begin{array}{ll}
-      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\! \ast \MMIOag
-         \vdash\\
-         \quad \forall W \ldotp\;(\ER^{\notMMIO} (RX, g, b, e, a_i))\spac W
-      \end{array}} \right. \\
+  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
+      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\!
+         \vdash
+         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
        \\
-    (r[\overline{r_{a_{i}} := (E, g, b, e, a_i)}], m, \emptyset) \longrightarrow^*
+     \\
+    (r[\overline{r_{a_{i}} := (\X{E}, g, b, e, a_i)}], m, \emptyset) \longrightarrow^*
     (r', m', t)
@@
 
-The capability $(RX, g, b, e, a_i)$ points to the $i$th entry point of the code
-we want to gradually verify.
-The set of entry points is given by $A_{\X{entry}}$.
-The capability $(E, g, b, e, a_i)$ then represents
-a closure for this same piece of code. The adversary only gets access to the
-closure, in order to avoid them from executing or reading arbitrary lines of
+The code that is verified (similar to the expression $e$ in OCPL) is here given
+as a set of \emph{entrypoints} (addresses) $A_{\X{entry}}$.
+%
+The capability $(\X{RX}, g, b, e, a_i)$ then points to the $i$th entry point of
+the code we want to gradually verify.
+
+The capability $(\X{E}, g, b, e, a_i)$ then represents a closure for this same piece
+of code. Adversarial code only gets access to the closure (as an enter
+capability), in order to avoid them from executing or reading arbitrary lines of
 code within the trusted module.
+
 Notice the universal quantification over worlds in the precondition;
@@
 called.
-This universal quantification was not explicitly present in the OCPL
+
+\begin{remark}
+  We could simplify the theorem statement by only considering the case of a
+  single entrypoint. This is equivalent: a piece of code that wants to expose
+  two distinct ``methods'' can implement a single toplevel entrypoint that
+  immediately returns two pointers that can then be used to invoke the two
+  methods.
+  %
+  This would make the toplevel theorem simpler, at the cost of somewhat more
+  contrieved calling convention between trusted and untrusted code.
+\end{remark}
+
+\begin{remark}
+The universal quantification on worlds was not explicitly present in the OCPL
 formalization.
@@
 
-\paragraph*{Remark:} it would be interesting to have a discussion about how to
-do away with the worlds if we remove local capabilities, regardless of whether
-we will in the end, since that might shed some light on why the worlds look the
-way they do right now.
+It would be interesting to have a discussion about how to do away with the
+worlds if we remove local capabilities, regardless of whether we will in the
+end, since that might shed some light on why the worlds look the way they do
+right now.
+\end{remark}
 
-Also note that we might as well have written
-$\ldots \vdash \forall W \ldotp \; (\VR^{\notMMIO} \ldots $ %)
-in the above theorem, instead of having the expression relation, since this is
-the actual relation we will need, and easily unfolds to the former statement
-through the definition of the value relation. I wrote $\ER$ to place the focus
-on the execution of the driver being safe. I consider this a matter of taste.
+\begin{remark}
+  Note that we could have written
+  $\ldots \vdash \forall W \ldotp \; (\VR^{\notMMIO} (\X{E}, \ldots))\; W$ %)
+  in the above theorem, instead of using the expression relation. This is
+  equivalent, and easily unfolds to the former statement through the definition
+  of the value relation. I wrote $\ER$ to place the focus on the execution of
+  the driver being safe. I consider this a matter of taste.
+\end{remark}
 
-The above theorem is still lacking, since it has no notion of adversarial
-context \textit{AdvCtx} to limit the power of the attacker. Indeed, if we allow
-an attacker to have read or write access to any location in \MMIO, or possess
-any capability directly pointing into the driver, then all hope we might have of
-gradually enforcing safety properties is now lost. We need a check on the memory
-and registers that we will allow our driver closures to be used with.
+\medskip
 
-To enforce these syntactic restrictions, we have two options:
+The theorem above is \emph{incomplete}. In other words, it is not provable as
+is, because it includes no notion of adversarial context (similar to OCPL's
+\textit{AdvCtx}), and therefore imposes no restriction on the adversarial code
+(i.e., the contents of the $r$ and $m$ other than our verified code).
+%
+As such, the statement can be trivially falsified by taking an attacker with
+direct read or write access to any location in \MMIO, or with direct access to
+the trusted code's internal state.
+
+We move to a theorem statement that correctly constrains the adversary in two
+steps:
+%
 \begin{enumerate}
-\item Add a syntactic check to the statement of our robust safety theorem,
-  ensuring that the adversary code cannot gain access to any non-E capabilities
-  pointing into the driver code or any capabilities pointing into $\MMIO$ directly.
-\item Assume the existence of a piece of boot code, that gets to run first when
-  the capability machine starts up, and makes sure the above condition actually
-  holds before passing control to the adversary. We could either assume that the
-  adversarial memory can contain any capability at startup, in which case the
-  driver has to erase the memory before passing control, or we could assume that
-  the memory cannot contain any capabilities whatsoever, in which case the boot
-  code does not have this obligation.
+  \item We add a syntactic condition (on the state of memory and registers) to
+    the tentative statement of our robust safety theorem, ensuring that the
+    adversary code cannot gain access to any non-$\X{E}$ capabilities pointing into
+    the driver code or any capabilities pointing into $\MMIO$ directly.
+%
+  \item Then, we provide an end-to-end statement that assumes the existence of a
+    piece of boot code, that gets to run first when the capability machine
+    starts up, and makes sure the above condition indeed holds before passing
+    control to the adversary.
+
+    The boot code is part of the trusted code base, and must satisfy a given
+    specification, as a precondition of the toplevel theorem.
+
+    The particular implementation of the boot code depends on the model of the
+    initial state of the machine. If the memory can contain any capability at
+    startup, then the boot code has to erase the memory before passing control.
+    If we assume that the initial memory cannot contain any capabilities
+    whatsoever, then the boot code does not need to do any cleanup.
 \end{enumerate}
-
-If we were to follow option 2 and write concrete boot code, part of our proof
-would consist of reducing option 2 to the situation in option 1, since this is
-where unknown adversary code actually starts executing. In a way, option 1 is
-hence more fundamental than option 2. We therefore adapt the robust safety
-theorem to option 1 first.
 
@@
   include the boot code in the general setting, since it does not contain any
-  inherently dangerous capabilities itself)
-\end{itemize}.
+  inherently dangerous capabilities itself).
+\end{itemize}
+
+\newcommand{\nonCap}[1]{\ensuremath{\mathrm{nonCap}(#1)}}
+
+We can then implement the syntactic check as requiring registers and memory to
+only contain non-capability values:
+
+\def \MathparLineskip {\lineskiplimit=0.7em\lineskip=0.7em}
+\begin{mathpar}
+  \nonCap{\X{inl}\; z} \eqdef \TRUE \and
+  \nonCap{\X{inr}\; (p,g,b,e,a)} \eqdef \FALSE \and
+  \nonCap{\V{reg}} \eqdef \forall r \in \dom(\V{reg}).\; \nonCap{\V{reg}(r)} \and
+  \nonCap{\V{mem}} \eqdef \forall l \in \dom(\V{mem}).\; \nonCap{\V{mem}(l)}
+\end{mathpar}
 
@@
 
-We can then implement the syntactic check using the following predicate
-$\notMapsToR{R}{w}$, that takes a word, memory region or register file and checks if it points into the forbidden region $R$ as follows:
+Given this definition, we can now formalize the initialization constraints on
+registers $r$ and memory $m$, given the region $\MMIO$ and
+the driver's address space $[b,e)$ as follows:
+
+\begin{mathpar}
+  \mprset{vskip=0.4em}
+  \inferrule
+  { m = m_{\MMIO} \uplus m_{\X{driver}} \uplus m_{\X{adv}}\\
+    \dom( m_{\MMIO} ) = \MMIO \\
+    \dom( m_{\X{driver}} ) = [b,e)\\
+    \dom( m_{\X{adv}} ) = [b_{\X{adv}},e_{\X{adv}})\\
+    \nonCap{\V{r}}\\
+    \nonCap{\V{m_{\X{adv}}}} \\
+    r[\X{PC}] = (\X{RWX}, \X{G},b_{\X{adv}},e_{\X{adv}},b_{\X{adv}})\\
+  }
+  {\initOK{[b,e)}}
+\end{mathpar}
+
+and subsequently formulate the robust safety theorem as follows:
+
+\begin{mathpar}
+  \mprset{vskip=0.5em}
+  \inferrule
+  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
+      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\!
+         \vdash
+         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
+         \\
+   \initOK{[b,e)} \\
+  (r[\overline{r_{a_{i}} := (\X{E}, g, b, e, a_i)}], m, \emptyset) \longrightarrow^* (r', m', t)\\
+  }
+  {P(t)}
+\end{mathpar}
+
+where we have set the PC up to be a sensible capability (pointing to
+$b_{\X{adv}}$ without loss of generality).
+
+Notice how the syntactic check for the registers is only performed on the set of registers $r$, allowing the closures in $\overline{r_{a_i}}$ to point into to the driver address space.
+
+\begin{remark}
+  If useful, the syntactic check can be relaxed to allow any capabilities that
+  do not point into the MMIO region or the driver's memory. One would define the
+  predicate $\notMapsToR{R}{w}$, that takes a word, memory region or register
+  file and checks if it points into the forbidden region $R$ as follows:
 
@@
     %
-    \notMapsToR{R}{\X{inr}\spac (p,g,b,e,a)} & \eqdef & R \cap [b,e[\; =
+    \notMapsToR{R}{\X{inr}\spac (p,g,b,e,a)} & \eqdef & R \cap [b,e)\; =
                                                  \emptyset  \\
@@
 
-Given this definition, we can now formalize the initialization constraints on
-registers $r$ and memory $m$, given the region $\MMIO$ and
-the driver's address space $[b,e[$ as follows;
-
-\begin{mathpar}
-  \mprset{vskip=0.4em}
-  \inferrule
-  { m = m_{\MMIO} \uplus m_{\X{driver}} \uplus m_{\X{adv}}\\
-    m_{\MMIO} = \MMIO \rightarrow \mbox{\Lightning} \\
-    \dom( m_{\X{driver}} ) = [b,e[\\
-    \dom( m_{\X{adv}} ) = [b_{\X{adv}},e_{\X{adv}}[\\
-    \notMapsToR{\MMIO \cup [b,e[}{\V{r}}\\
-    \notMapsToR{\MMIO \cup [b,e[}{\V{m_{\X{adv}}}} \\
-    r \regdr \X{PC} = (\X{RWX}, \X{G},b_{\X{adv}},e_{\X{adv}},b_{\X{adv}})\\
-  }
-  {\initOK{\MMIO}{[b,e[}}
-\end{mathpar}
-
-and subsequently formulate the robust safety theorem as follows:
-
-\begin{mathpar}
-  \mprset{vskip=0.4em}
-  \inferrule
-  { \forall a_{i} \in A_{\X{entry}} \ldotp \left\{
-      {\begin{array}{ll}
-      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\! \ast \MMIOag
-         \vdash\\
-         \quad \forall W \ldotp\;(\ER^{\notMMIO} (RX, g, b, e, a_i))\spac W
-       \end{array}} \right. \\
-   \initOK{\MMIO}{[b,e[} \\
-  (r[\overline{r_{a_{i}} := (E, g, b, e, a_i)}], m, \emptyset) \longrightarrow^* (r', m', t)\\
-  }
-  {P(t)}
-\end{mathpar}
-, where we have set the PC up to be a sensible capability (pointing to $b_{\X{adv}}$
-without loss of generality).
-
-Notice how the syntactic check for the registers is only performed on the set of registers $r$, allowing the closures in $\overline{r_{a_i}}$ to point into to the driver address space.
+Then, one would use $\notMapsToR{\MMIO \cup [b,e)}$ in place of $\nonCap{}$.
+\end{remark}
 
@@
 allows us to prove the above theorem.
-Alternatively, we could also just implement one specific instance of boot code, and prove that after it finishes executing, all preconditions to apply the above theorem are met.
-We can use the following predicate to specify that the starting configuration is
-valid:
+%
+(Alternatively, we could also just implement one specific instance of boot code,
+and prove that after it finishes executing, all preconditions to apply the above
+theorem are met.)
+
+Let $r_0$ and $m_0$ be the respective initial state of registers and memory
+immediately after booting. We assume that the following two properties hold,
+where $l_\X{boot}$ is some memory location (initially loaded in PC):
+%
+\begin{mathpar}
+  \dom(m_0) = [0,\X{MEM_{MAX}}) \and
+  r_0[\X{PC}] = (\X{RWX}, \X{G},0,\X{MEM_{MAX}},l_{\X{boot}})
+\end{mathpar}
+
+Depending on the specifics of the operiational semantics, we might have more
+information about $m_0$ and $r_0$, that can then be used in the implementation
+of the boot code (e.g., $m_0$ might be initialized to only zeroes).
+
+We also assume the boot-code to be initially loaded in memory at address
+$l_{\X{boot}}$.
+%
+Then, the specification for the boot code that we have to prove is stated as
+follows:
+
+\newcommand{\bigast}[2]{\underset{#1}\Sep{\!\!\! #2}\;}
+
+\[
+  \begin{array}{l}
+  \X{bootSpec}(P, g, b, e, A_{\X{entry}}) \eqdef {} \\[0.8em]
+    \quad
+  \begin{array}{l}
+    \left\{
+    \begin{array}{l}
+      \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, 0, \X{MEM_{MAX}}, l_{\X{boot}}) \ast {} \\[0.5em]
+      \bigast{r \in \X{Reg} \setminus \X{PC}}{r \mapsto_{r} \_}
+      \ast \bigast{m \in \X{Mem} \setminus \MMIO}{m \mapsto_{a} \_}
+      \ast \tracefrag{\emptyset}
+    \end{array}
+    \right\}
+    \\[2em]
+    \quad \I{Instr}\; \I{Executable}
+    \\[0.5em]
+    \left\{
+    \begin{array}{l}
+      \exists b_{adv}\; e_{adv}\; t.\; \\[0.5em]
+      \quad
+      \begin{array}{l}
+        \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{adv}, e_{adv}, b_{adv}) \ast {} \\[0.7em]
+        \bigast{a_{i} \in A_{\X{entry}}}{a_{i} \mapsto_{r} (\X{E},g,b,e,a_i)} \ast
+        \bigast{r \in Reg\setminus \{\X{PC}, A_{\X{entry}}\}}{r \mapsto_{r} \X{inl}\; \_} \ast {} \\[1.4em]
+        \bigast{m \in [b_{adv}, e_{adv})}{m \mapsto_{a} \X{inl}\; \_} \ast 
+        \tracefrag{t} \ast \pure{P(t)}
+      \end{array}
+    \end{array}
+    \right\}
+  \end{array}
+  \end{array}
+\]
+
+Intuitively, establishing this specification ensures that, after running the
+boot code, we arrive in a state that satisfies $\initOKo$.
+
+The resulting boot-code-base formulation of the robust safety theorem is now a
+corollary of the general robust safety theorem, and has the following statement:
 
 \begin{mathpar}
-  \mprset{vskip=0.4em}
+  \mprset{vskip=0.5em}
   \inferrule
-  { m = m_{\MMIO} \uplus m_{\X{driver}} \uplus m_{\X{boot}}\\
-    m_{\MMIO} = \MMIO \rightarrow \mbox{\Lightning} \\
-    \dom(m_{\X{driver}}) = [b,e[\\
-    \dom(m) = [0,\X{MEM_{MAX}}[\\
-    r \regdr \X{PC} = (\X{RWX}, \X{G},0,\X{MEM_{MAX}},l_{\X{boot}})\\
-  }
-  {\initBC{\MMIO}{[b,e[}}
-\end{mathpar}
-where $l_{\X{boot}}$ is the location where the machine boots, to be determined
-by the machines operational semantics.
-
-On the level of the operational semantics, we have to prove that the
-following holds, in order to use our concrete driver implementation with the
-robust safety theorem above (the driver still needs to be proven safe separately):
-
-\begin{mathpar}
-  \mprset{vskip=0.4em}
-  \inferrule
-  { \initBC{\MMIO}{[b,e[}\\
-  }
-  { \exists n\ldotp (r, m, \emptyset) \longrightarrow^n (r'[\overline{r_{a_{i}}
-      := (E, g, b, e, a_i)}], m', t) \land P(t) \land\\
-  \initOK{\MMIO}{[b,e[}[r'][m']}\quad (\textsc { BootCodeOK })
-\end{mathpar}
-
-Note that to link scenario 2 to scenario 1, we need a slightly adapted version of the robust safety theorem stated above
-where the initial trace is allowed to not be empty, but rather just satisfy
-$P(t)$, but that should not cause any problems.
-For a specific instance of boot code that provably satisfies the \textsc{BootCodeOK}
-rule, the resulting boot-code-based formulation of the robust safety theorem is now
-just a corollary of the general robust safety theorem, and has the following
-statement:
-
-\begin{mathpar}
-  \mprset{vskip=0.4em}
-  \inferrule
-  { \forall a_{i} \in A_{\X{entry}} \ldotp \left\{
-      {\begin{array}{ll}
-      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\! \ast \MMIOag
-         \vdash\\
-         \quad \forall W \ldotp\;(\ER^{\notMMIO} (RX, g, b, e, a_i))\spac W
-       \end{array}} \right. \\
-   \initBC{\MMIO}{[b,e[} \\
-  (r, m, \emptyset) \longrightarrow^* (r', m', t)\\
+  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
+      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\!
+         \vdash
+         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
+         \\
+   \X{bootSpec}(P,g,b,e,A_{\X{entry}}) \\
+  (r_0, m_0, \emptyset) \longrightarrow^* (r, m, t)\\
   }
@@
 
+Note that to link scenario 2 to scenario 1, we need a slightly adapted version
+of the robust safety theorem stated above where the initial trace is allowed to
+not be empty, but rather just satisfy $P(t)$, but that should not cause any
+problems.
+
+Note as well that here, if the boot-code performs I/O, it is required to satisfy
+the invariant $P$ on the trace. If needed, one could come up with a more general
+theorem statement, where the boot code is allowed to perform arbitrary I/O, and
+where $P$ is thus only enforced on the suffix of the trace \emph{after} the boot
+code has run...
+
+\begin{remark}
+  On the level of the operational semantics, proving the boot specification
+  corresponds to proving that the following holds, in order to use our concrete
+  driver implementation with the robust safety theorem above (the driver still
+  needs to be proven safe separately):
+\[
+  \begin{array}{l}
+    (\textsc {BootCodeOK}) \\[0.5em]
+%
+    \exists n\; b\; e\ldotp (r_0, m_0, \emptyset) \longrightarrow^n (r'[\overline{r_{a_{i}}
+    := (\X{E}, g, b, e, a_i)}], m', t) \\[0.5em]
+    {} \land P(t) \land \initOK{[b,e)}[r'][m']
+  \end{array}
+\]
+\end{remark}
+
+\subsubsection{Concrete boot and driver code}
+
 Now we have a closer look at the concrete example we would like to verify,
 including what its boot code looks like. Our goal is to prove that this concrete
-example satisfies \textsc{BootCodeOk} and this last robust safety theorem.
+example satisfies \X{bootSpec} (alternatively, \textsc{BootCodeOk}) and this
+last robust safety theorem.
 
@@
 \begin{itemize}
-\item Generate the necessary closures (i.e.
-Enter capabilities) for the driver's read and write methods, so that the adversary can use safely use them to perform I/O, and any properties that the driver wants to uphold on the input-\textbackslash output stream can actually be enforced.
-\item Wipes all of the adversary's sandbox section, to make sure no remaining
-  capabilities can be found there (\textbf{Design Alternative:} maybe we should try to find out how CHERI handles this; is it possible that any lingering capabilities are left in RAM
-  on machine start-up? In any case, we are already being slightly unrealistic by assuming the adversary code section is just \emph{there}. We have the option
-  to not do any erasure, and just make the assumption that adversary memory
-  contains no capabilities pointing into MMIO or pointing into the driver. ).
-  It does n\'ot, however, wipe the untrusted, hard-coded adversary routine that reads $N$ lines of code from a single (and the only) MMIO location in memory.
-\item Reduces the omnipotent capability to provide RWX access to all of the
-  adversary's code, makes it point to the first element of the adversary's code
-  section and jumps to it, in order to hand control to the adversary. Before the
-  jump, it clears all registers except for the 2 containing the read and write
-  driver closures, in order to make sure that no extra permissions leak to the adversary.
+  \item Generate the necessary closures (i.e. Enter capabilities) for the
+    driver's read and write methods, so that the adversary can use safely use
+    them to perform I/O, and any properties that the driver wants to uphold on
+    the input-output stream can actually be enforced.
+%
+  \item Wipe all of the adversary's sandbox section, to make sure no remaining
+  capabilities can be found there.
+  %
+    (\textbf{Design Alternative:} maybe we should try to find out how CHERI
+    handles this; is it possible that any lingering capabilities are left in RAM
+    on machine start-up? In any case, we are already being slightly unrealistic
+    by assuming the adversary code section is just \emph{there}. We have the
+    option to not do any erasure, and just make the assumption that adversary
+    memory contains no capabilities pointing into MMIO or pointing into the
+    driver.).
+  %
+    It does not, however, wipe the untrusted, hard-coded adversary routine that
+    reads $N$ lines of code from a single (and the only) MMIO location in
+    memory.
+%
+  \item Reduce the omnipotent capability to provide RWX access to all of the
+    adversary's code, makes it point to the first element of the adversary's
+    code section and jumps to it, in order to hand control to the adversary.
+    Before the jump, it clears all registers except for the 2 containing the
+    read and write driver closures, in order to make sure that no extra
+    permissions leak to the adversary.
 \end{itemize}

### 2020-03-24 capability-meetings-notes/10-03/main.tex
MMIO meeting notes: tweak

@@
 Note that to link scenario 2 to scenario 1, we need a slightly adapted version
-of the robust safety theorem stated above where the initial trace is allowed to
-not be empty, but rather just satisfy $P(t)$, but that should not cause any
-problems.
+of the robust safety theorem stated previously, where the initial trace is
+allowed to not be empty, but rather just satisfy $P(t)$, but that should not
+cause any problems.
 
-Note as well that here, if the boot-code performs I/O, it is required to satisfy
+Note as well that if the boot-code performs I/O, then it is required to preserve
 the invariant $P$ on the trace. If needed, one could come up with a more general
 theorem statement, where the boot code is allowed to perform arbitrary I/O, and
-where $P$ is thus only enforced on the suffix of the trace \emph{after} the boot
-code has run...
+where $P$ is thus only true of the suffix of the trace for events that occur
+\emph{after} the boot code has run...
 

### 2020-03-24 capability-meetings-notes/10-03/main.tex
meetitng notes: another fix

@@
 now requires $\tracefrag{t}$ in the pre-condition (for some $t$), and provides
-$\tracefrag{(\X{IORead}, a, x) :: t}$ in the post-condition.
+$\tracefrag{t \app (\X{IORead}, a, x)}$ in the post-condition.
 

### 2020-03-25 capability-meetings-notes/25-03/main.tex
meeting notes for meeting on March 25

\documentclass[a4paper, 10pt]{article}
\topmargin-4.0cm

\usepackage[english]{babel}
\usepackage{csquotes}
\usepackage[a4paper, margin=2cm]{geometry}
\usepackage{fancyhdr}
\usepackage{color}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{url}
\usepackage{hyperref}
\usepackage{lastpage}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}
\usepackage{datetime}

\usepackage{newpxtext}
\usepackage{newpxmath}

\begin{document}
\lhead{}
\rhead{}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}

\newdate{date}{17}{03}{2020}
\newdate{friday}{20}{03}{2020}
\newdate{next}{24}{03}{2020}

\begin{center}
{\LARGE \bf Meeting Notes}\\
\displaydate{date}
\end{center}

\section{Discussions about the notes}
\begin{itemize}
\item \underline{Uninitialised capabilities}

We agree that it looks good! Arma\"{e}l suggested it might be easier to make store initialise instead of storeu (where it currently increments it if we are on top). Would it make it complicated to just have store do the increment instead? proposal: not have a special case for storeU, but rather have a special case for store for U- capabilities that do the increment. The intuition is that we offset the load for proving everything in storeU, and split the work more evenly between store and storeU. Another alternative, is to enable store to write to U- capabilities, but have storeU be the specific store + increment. 
(Related work is making sure that all the regular stores be replaceable with storeU). 
For now we leave it like it is, but if the FTLR for storeU becomes too big, we might want to consider a split.

\item \underline{Device drivers}

Comments from Lars: 
\begin{itemize}
	\item The semantics are non deterministic because of the IOload, we should take note of all sources of non-determinism since they can cause trouble later on. 
	\item What is the simplest possible client program that one could consider? where one can think about instantiating P with something concrete (e.g. even numbers). It might be worth considering, since it forces us to compose these different parts. The original client Dominique presented is that we are not accessing certain parts of memory. A similar but contrived examples is that every element of the trace does not go above a certain threshold. 
Even though P is simple, it is interesting to see how these specs might fit together. 
	\item Reactive programming uses streams of outputs. An alternate to the non determinism is that Load takes the next element in some oracle stream.

Non determinism can be one way of modelling input, streams could be an alternative. In CakeML there is an abstract type for the state of the outside world, and an oracle and operations that change the state of the outside world. Whenever an IO operation happen, we keep track of it in a trace. This means we keep a model of the external world in the semantics. 
Which model is going to make the rest of the story the simplest here? 
What is the impact of non determinism? Here it’s still internally deterministic, but it’s with the external interactions with the world that non-determinism happen. There might a way to relax the determinism restriction for the WP lemmas we will need. 
To make it deterministic, we can make x be a parameter of the step (this parameter can be given by some kind of oracle).

Non determinism can lead to challenged down the road. We should be aware of this!

Let’s consider making it deterministic using some kind of oracle to avoid such challenges. 
\end{itemize}

Discussion about the spec on page 10/11: 
\begin{itemize}
\item If the boot code is doing IO, then we also need to state that the boot code produces a trace t for which P(t) holds.
\item We assume that code can produce multiple entry and exit points. Proposed alternative: one entry point and a registers that select what kind of thing we want to do. 
\item It might make things easier to read if we use notation for the fragmental trace resources. 
\item Some comments from Dominique:

Here the theorem does not make any assumptions about memory. (here we know nothing about memory. If memory contains any capabilities, the boot code will need to clear memory in order to satisfy the spec. What if we do know that memory is all 0’s at boot? then we cannot prove that the boot satisfies the step without clearing. This might be an issue? We need to mention that as a guarantee in the theorem. Let’s figure out what happens in real life before we add such an assumption).

spec on page 11: universal quantification over all W. No link between the world the boot code sets up and the world the expression relation is prover over. We need to consider all private future worlds of the world the boot code allocates.

page 10: we will need some assumptions over the memory that the entry points are valid.

can we generalise the entry points to be any permission, not just E-capabilities (i.e. closures). We could generalise the statement by stating validity. This might simplify the condition, the PC is kept as only of the registers.

\item Some comments from Frank: Does the boot code need to do IO? depending on the goal of the work: if we want to model what OS boot code does on IO, then yes, boot code does initialisation of devices which might leave a trace). Do we want to see this evolve into proving an actual boot code correct? Or do we want to focus on something where more fancy Iris stuff is needed. These might be two different directions for the project. 
\end{itemize}

\end{itemize}

\section{Next steps for the driver}

\begin{itemize}
	\item fix the statement of the theorem such that the world produced by the boot code is linked to the world the expression relation must hold over. 
	\item think of a client and a proof sketch of that minimal client (on paper) to get an idea of how these moving parts fit together.
\end{itemize}

\section{Next steps for the uninitialized capability}

\begin{itemize}
	\item prove the awkward example without clearing the stack upon return. This step will need further discussions (Aina will send an email thursday if such a discussion can already happen on friday)
	\item start proving WP's and FTLR for the now implemented semantics 
\end{itemize}

\end{document}

### 2020-03-25 capability-meetings-notes/10-03/main.tex
MMIO meeting notes: fix: boot code must prove that the driver is distinct from the adversary
MMIO writeup: Instr Executable -> Seq (Instr Executable)
Fix a typo

@@
       \ast \bigast{m \in \X{Mem} \setminus \MMIO}{m \mapsto_{a} \_}
-      \ast \tracefrag{\emptyset}
+      \ast \; \tracefrag{\emptyset}
     \end{array}
@@
     \\[2em]
-    \quad \I{Instr}\; \I{Executable}
+    \quad \I{Seq}\; (\I{Instr}\; \I{Executable})
     \\[0.5em]
@@
         \bigast{m \in [b_{adv}, e_{adv})}{m \mapsto_{a} \X{inl}\; \_} \ast
-        \tracefrag{t} \ast \pure{P(t)}
+        \bigast{m \in [b, e)}{m \mapsto_{a} \_} \ast
+        \; \tracefrag{t} \ast \pure{P(t)}
       \end{array}
@@
 
-The resulting boot-code-base formulation of the robust safety theorem is now a
+The resulting boot-code formulation of the robust safety theorem is now a
 corollary of the general robust safety theorem, and has the following statement:

### 2020-03-31 capability-meetings-notes/10-03/main.tex
MMIO writeup: new formalization of memory + non-determinism

--- 2020-03-24 capability-meetings-notes/10-03/main.tex
@@
 
-Notice that, for any address in \MMIO, which value is held by the map $m$ as
+\newcommand{\stdin}{\X{in}}
+\begin{remark}
+As Lars remarked, the read rule introduces non-determinism into our operational
+semantics.
+This has lead to some complications in the past (the nature of which is still
+mysterious to me; n.b. that we use\\ $\I{ wp\_lift\_atomic\_head\_step\_no\_fork}$
+in our WP rules, and that one would allow non-determinism).
+(one question I -Thomas- had in this respect; does the $\I{Alloc}$ rule in HeapLang
+not introduce non-determinism in a similar way? Why is this not problematic?)
+Lars' suggestion was to additionally parameterize the operational semantics by
+an input-stream $\stdin$.
+The updated \textsc{IOLoad} load rule might then look something like this:
+\begin{mathpar}
+  \inferrule[IOLoad]
+  { r[\X{src}] = (p,g,b,e,a) \\ \X{readAllowed}\; p \\ a \in [b,e) \\ a \in \MMIO
+  \\ n = \#\{i \mid t[i] = ( \X{IORead},\_\;,\_) \}  }
+  { (r, m, t) \xrightarrow{\I{Load} \; \X{dst} \; \X{src}}
+    (r[\X{dst} := \stdin[n]], m, t \app (\X{IORead}, a, \stdin[n])) }
+\end{mathpar}
+
+When parameterizing our development with multiple values, we could either write
+them out as different Coq \texttt{Parameters} where we need them, or store them
+as fields in our \texttt{DriverG} typeclass (the latter is the approach taken
+in e.g.\ the code-base of ~\cite{sammler-2020}).
+\end{remark}
+
+We now have two options when it comes to modeling the memory $m$:
+\begin{enumerate}
+\item Include the \MMIO addresses into the map. Notice that, for any address in \MMIO, the specific value held by the map $m$ for
 that address is now irrelevant. Indeed, $m$ is never read nor modified for
-addresses in \MMIO. One could choose to enforce that a particular dummy value is
-stored in $m$ for these addresses. Instead, we choose to leave them
+addresses in \MMIO (possible exception: see next remark). One could choose to enforce that a particular dummy value is stored in $m$ for these addresses. Instead, we choose to leave them
 unconstrained.
+\item Have $m$ model the physical RAM addresses only. This model is closer to
+  the conceptual model of state that we wish to implement, since it does not
+  require us to talk about the values in the heap that \MMIO locations are
+  associated with. In case it is not harder to implement (which I currently do not think
+  it is), it is to be preferred.
+\end{enumerate}
+Both approaches require us to make sure that no points-to chunks $l \mapsto \_$
+with $l \in \MMIO$ can be created. In the next section, we assume the second
+approach (until someone proves me wrong about it working).
 
 \begin{remark}
-  Thomas: Additionally, we might want to disallow in the semantics an execution
-  with a \X{PC} register pointing to an $\MMIO$ location (I think this makes
-  more sense than reading a random instruction through I/O). Armaël: but it
-  makes the semantics of all instructions more complicated.
+  Additionally, we have the option of disallowing an execution
+  with a \X{PC} register pointing to a $\MMIO$ location in the semantics.
+  First off, consider that this is an edge case; since we always have points-to
+  chunks describing the layout of code in memory (and a points-to chunk for
+  location $l$ doubles as a proof that $l \not\in \MMIO$), it is trivial to prove that
+  this case would not occur in any of our current examples.
+  If we make no changes to the current code base, this behavior would be
+  allowed, and the following would happen in both cases described above:
+  \begin{enumerate}
+  \item An unconstrained value is read from the location. Any instruction could
+    be executed.
+  \item The predicate $\I{MemLocate}$ in $\I{lang.v}$ will make sure
+  that the default value $0$ is read and decoded into an instruction when we try
+  to execute an \MMIO location.
+  This would not cause a conflict with the current WP rules,
+  since a points-to chunk is required in the precondition, ensuring that this
+  case does not occur.
+  Another alternative is to read the encoding of a $\X{NOP}$ instruction, instead of
+  just the value $0$.
+\end{enumerate}
+  If we were to simply disallow execution when the PC points into \MMIO, we would make the semantics of all instructions slightly more complicated, but in a
+  uniform way.
+  We could simply add $a \not\in \MMIO$ to the predicate $\I{isCorrectPC}$ (or
+  add it as an additional predicate), which is used to define failing execution.
+  It would again not be a drastic change, since a points-to for the
+  address $a$ the PC points to is required by the WP rules regardless, from which we
+  can prove that $a \not\in \MMIO$.
 \end{remark}
@@
 monolithic fashion. In that case, the state interpretation additionally holds a
-resource for the trace.
+resource for the trace:
+
+\begin{align*}
+    \X{state\_interp} \; (r,m,t) :=& \X{gen\_heap\_ctx}\; r \ast\\
+    &\X{gen\_heap\_ctx}\; m \ast \dom(m) \cap \MMIO = \emptyset \ast
+    {\color{BrickRed}\tracefull{t}}
+\end{align*}
+
+\begin{remark}
+For the case where the \MMIO locations are still part of the heap $m$, this
+would look as follows:
 \[
@@
 \]
+
+Sanity check: we never have allocation of new locations $l$ in our capability machine, but if we
+did, that would work in both settings by requiring that $l \not\in \MMIO$ in the
+operational semantics rule for \textsc{Alloc}.
+\end{remark}
 
@@
 
-We also assume the boot-code to be initially loaded in memory at address
-$l_{\X{boot}}$.
+We also assume the capability machine to initially start execution in the boot code  at address $l_{\X{boot}}$.
 %
-Then, the specification for the boot code that we have to prove is stated as
-follows:
+Then, the syntactic specification for the boot code that we have to prove is stated as follows:
 

### 2020-03-31 capability-meetings-notes/10-03/main.tex
MMIO writeup: started writing section on clients

@@
         \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{adv}, e_{adv}, b_{adv}) \ast {} \\[0.7em]
-        \bigast{a_{i} \in A_{\X{entry}}}{a_{i} \mapsto_{r} (\X{E},g,b,e,a_i)} \ast
+        \bigast{a_{i} \in A_{\X{entry}}}{r_{a_{i}} \mapsto_{r} (\X{E},g,b,e,a_i)} \ast
         \bigast{r \in Reg\setminus \{\X{PC}, A_{\X{entry}}\}}{r \mapsto_{r} \X{inl}\; \_} \ast {} \\[1.4em]
@@
   { \forall a_{i} \in A_{\X{entry}} \ldotp\,
-      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\!
+    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast \pure{P(t)}}\!
          \vdash
@@
 \subsubsection{Concrete boot and driver code}
-
+\label{subsec:driverimpl}
 Now we have a closer look at the concrete example we would like to verify,
@@
 
+\section{Potential driver clients/properties}
+The concrete driver that we described in \texttt{subsec:driverimpl} does not yet enforce any useful
+safety properties on the traces that it outputs.
+This section describes a couple such properties that we might want to enforce,
+modifications to the driver that they require and example clients that make use of our modified driver.
+
+\subsection{Stateless properties}
+
+\paragraph{Printing values $\le 1000$}
+
+\paragraph{Modelling channel-specific safety properties}
+
+\subsection{Stateful properties}
+
+\paragraph{Performing $\le 1000$ I/O operations}
+
+To ensure that no more than 1000 values will be read or sent, the driver needs to
+keep some internal state, e.g.\ a single integer value $n$ (at location $l_n$) representing the number of times it has been called.
+
+To accommodate this local state, the initial world will need to contain one
+extra region for location $l_n$, that allows any integer values $\ge 0$. Since
+this region only has a single state, we can again short-circuit the world and
+enforce the constraints we want on $n$ through an invariant directly. We now
+discuss the concrete constraint that we are looking for.
+
+We have to tie this integer $n$ to the number of I/O events
+present in the trace $t$, otherwise it will prove impossible to verify our
+driver. Concretely, the invariant we will verify our driver under changes from
+$\knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast \pure{|t| \le 1000}}$ to
+$\knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast \pure{|t| \le 1000} \ast
+  \exists n\ldotp l_n \mapsto n \ast n = |t| }$. Concretely, this means that
+instead of having a pure predicate $P$ in our robust safety theorem, we should
+also allow non-pure predicates that entail $P$, since we might have to enforce
+other invariants relating to local state over the trace as well.
+
+\begin{remark}
+ I swept this under the rug in the above exposition, but if we allow boot code to run first, we have to
+ make sure that the value $n$ indeed corresponds to the number of I/O events
+ that have occurred when control is passed to the adversary.
+\end{remark}
+
+To make sure we can prove the driver specification, we will have to update the
+robust safety theorem slightly to incorporate these changes. It will then look
+as follows:
+
+\paragraph{Never performing I/O again after some stop token has been read on $\stdin$}
+
 \bibliographystyle{alpha}

### 2020-03-31 capability-meetings-notes/10-03/main.tex
MMIO notes; verbal description of semantic bootspec

@@
 
+It should be noted that a more general, semantic version of this specification
+is also possible. It would contain the following elements, that should all
+follow directly from the specification above:
+\begin{itemize}
+\item All registers except \X{PC} are in the value relation with respect to
+  either the empty world (if we do not want to allow any non-empty, non-enter
+  capabilities in registers), or a world containing a single standard region for
+  all of the adversary's memory.
+  The \X{PC} register satisfies both the read- and write conditions (note the
+  parallel with the fundamental theorem) with respect to this same world, if we
+  picked the non-empty world, or just points to the adversary's memory, which
+  does not contain capabilities (if we
+  picked the empty world, a capability pointing into memory can never be valid).
+  It feels slightly less artificial to pick the non-empty world option here.
+\item We have possession of the predicates $\I{region}$ and
+  $\I{sts\_full\_world}$ for the world we picked in the previous bullet. we will
+  need these to apply the fundamental theorem.
+\item A description of the exact driver code, still in terms of points-to chunks
+  (or, in this case equivalently, an invariant)
+\end{itemize}
+
 The resulting boot-code-base formulation of the robust safety theorem is now a

### 2020-03-31 capability-meetings-notes/10-03/main.tex
Allow the boot-spec precondition to know about r_0 and m_0

@@
       \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, 0, \X{MEM_{MAX}}, l_{\X{boot}}) \ast {} \\[0.5em]
-      \bigast{r \in \X{Reg} \setminus \X{PC}}{r \mapsto_{r} \_}
-      \ast \bigast{m \in \X{Mem} \setminus \MMIO}{m \mapsto_{a} \_}
+      \bigast{r \in \X{Reg} \setminus \X{PC}}{r \mapsto_{r} r_{0}[r]}
+      \ast \bigast{m \in \X{Mem} \setminus \MMIO}{m \mapsto_{a} m_{0}[m]}
       \ast \; \tracefrag{\emptyset}
@@
 
-The resulting boot-code-base formulation of the robust safety theorem is now a
-corollary of the general robust safety theorem, and has the following statement:
+The resulting boot-code formulation of the robust safety theorem is now a
 
@@
 \section{Potential driver clients/properties}
-The concrete driver that we described in \texttt{subsec:driverimpl} does not yet enforce any useful
-safety properties on the traces that it outputs.
-This section describes a couple such properties that we might want to enforce,
-modifications to the driver that they require and example clients that make use of our modified driver.
+The concrete driver that we described in Section~\ref{subsec:driverimpl} does
+not yet enforce any useful safety properties on the traces that it outputs. This
+section describes a couple such properties that we might want to enforce,
+modifications to the driver that they require and example clients that make use
+of our modified driver.
 
@@
 
-\paragraph{Never performing I/O again after some stop token has been read on $\stdin$}
+\paragraph{Never performing I/O again after some stop token has been read}
 

### 2020-03-31 capability-meetings-notes/10-03/main.tex
MMIO notes: better description of how we handle Worlds + remarks

--- 2020-03-31 capability-meetings-notes/10-03/main.tex
@@
 \DeclareMathOperator{\initOKo}{init_{OK}}
+\DeclareMathOperator{\initOKos}{init^{\X{sem}}_{OK}}
 \newcommandx{\initOK}[3][2=r,3=m]{\initOKo(#1,#2,#3)}
+\newcommandx{\initOKS}[4][2=r,3=m,4=W]{\initOKos(#1,#2,#3,#4)}
 \DeclareMathOperator{\initBCo}{init_{BC}}
+\DeclareMathOperator{\initBCos}{init^{\X{sem}}_{BC}}
 \newcommandx{\initBC}[2][1=r,2=m]{\initBCo(#1,#2)}
+\newcommandx{\initBCS}[3][1=r,2=m,3=W]{\initBCos(#1,#2,#3)}
+\DeclareMathOperator{\driverC}{code}
 
+\newcommand{\bigast}[2]{\underset{#1}\Sep{\!\!\! #2}\;}
 \newcommand{\app}{\mathbin{+\!\!+}}
@@
 
+\newcommand{\Wpub}{\sqsupseteq^{\X{pub}}}
+\newcommand{\Wpriv}{\sqsupseteq^{\X{priv}}}
 \begin{mathpar}
@@
   contrieved calling convention between trusted and untrusted code.
+  Another alternative is to provide an opcode to specify the operation that
+  needs to be performed.
+  This is slightly less general, since it does not allow providing read access
+  only to an adversary.
 \end{remark}
@@
     the driver code or any capabilities pointing into $\MMIO$ directly.
+    Alternatively, we can formulate this condition in a semantic way, see below.
+    Additionally, we specify what exact code the driver contains, so that we can
+    use this in our proof that the driver is secure. This will involve adding an
+    invariant to the robust safety theorem.
 %
@@
 \newcommand{\notMapsToR}[2]{\rightarrow^{\overline{#1}}\!\!(#2)}
+\newcommand{\Winit}{\X{W_{init}}}
+\newcommand{\Wemp}{\emptyset}
 
@@
 registers $r$ and memory $m$, given the region $\MMIO$ and
-the driver's address space $[b,e)$ as follows:
+the driver's address space $[b,e)$ as follows (for a resource-aware formulation;
+see the bootSpec postcondition below):
 
@@
 
-and subsequently formulate the robust safety theorem as follows:
+It should be noted that a more general, semantic version of this specification
+is also possible. It would contain the following elements, that should all
+follow directly from the specification above (TODO: write this out properly and
+do the same for the semantic bootspec):
+\begin{itemize}
+\item All registers except \X{PC} are in the value relation with respect to
+  either the empty world $\Wemp$ (if we do not want to allow any non-empty, non-enter
+  capabilities in registers), or a world $\Winit$ containing a single standard region for
+  all of the adversary's memory.
+  The \X{PC} register satisfies both the read- and write conditions (note the
+  parallel with the fundamental theorem) with respect to this same world, if we
+  picked the non-empty world, or just points to the adversary's memory, which
+  does not contain capabilities (if we
+  picked the empty world, a capability pointing into memory can never be valid).
+  It feels slightly less artificial to pick the non-empty world option here.
+\item We have possession of the predicates $\I{region}$ and
+  $\I{sts\_full\_world}$ for the world we picked in the previous bullet. we will
+  need these to apply the fundamental theorem.
+\item A description of the exact driver code, still in terms of points-to chunks
+  (or, in this case equivalently, an invariant). We can write this invariant as
+      $\driverC(b,e,\overline{\V{op}_l}) = \knowInv{\iota}{\bigast{l \in [b,e)}{l \mapsto_{a} \mathit{\V{op}_l}}}$, with $\V{op}_l$
+      some hard-coded driver instruction at location $l$.
+\end{itemize}
+
+and subsequently formulate the robust safety theorem as follows (using the
+semantic specification for $\initOKo$, which references the initial world):
 
@@
   \inferrule
-  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
-      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\!
-         \vdash
-         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
-         \\
-   \initOK{[b,e)} \\
+  { \forall a_{i} \in A_{\X{entry}} \ldotp\,\left\{
+    {\begin{array}{ll}
+    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\ast
+    \driverC(b,e,\overline{\V{op}_l})\!
+         \vdash \\
+         \forall W \Wpriv \Winit \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
+        \end{array}} \right. \\
+   \initOKS{[b,e)}[r][m][\Winit] \\
   (r[\overline{r_{a_{i}} := (\X{E}, g, b, e, a_i)}], m, \emptyset) \longrightarrow^* (r', m', t)\\
@@
 
-Notice how the syntactic check for the registers is only performed on the set of registers $r$, allowing the closures in $\overline{r_{a_i}}$ to point into to the driver address space.
+Notice how the syntactic check for the registers is only performed on the set of
+registers $r$, allowing the closures in $\overline{r_{a_i}}$ to point into to
+the driver address space (TODO: we should change this though and add a
+requirement to $\initOKo$).
+(TODO: introduce notation for trace-owndership and predicate P holding)
 
@@
 
-\newcommand{\bigast}[2]{\underset{#1}\Sep{\!\!\! #2}\;}
-
 \[
@@
 
-It should be noted that a more general, semantic version of this specification
-is also possible. It would contain the following elements, that should all
-follow directly from the specification above:
-\begin{itemize}
-\item All registers except \X{PC} are in the value relation with respect to
-  either the empty world (if we do not want to allow any non-empty, non-enter
-  capabilities in registers), or a world containing a single standard region for
-  all of the adversary's memory.
-  The \X{PC} register satisfies both the read- and write conditions (note the
-  parallel with the fundamental theorem) with respect to this same world, if we
-  picked the non-empty world, or just points to the adversary's memory, which
-  does not contain capabilities (if we
-  picked the empty world, a capability pointing into memory can never be valid).
-  It feels slightly less artificial to pick the non-empty world option here.
-\item We have possession of the predicates $\I{region}$ and
-  $\I{sts\_full\_world}$ for the world we picked in the previous bullet. we will
-  need these to apply the fundamental theorem.
-\item A description of the exact driver code, still in terms of points-to chunks
-  (or, in this case equivalently, an invariant)
-\end{itemize}
-
 The resulting boot-code-base formulation of the robust safety theorem is now a
@@
   \inferrule
-  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
-    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast \pure{P(t)}}\!
-         \vdash
-         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
-         \\
-   \X{bootSpec}(P,g,b,e,A_{\X{entry}}) \\
+  {  \forall a_{i} \in A_{\X{entry}} \ldotp\,\left\{
+    {\begin{array}{ll}
+    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\ast
+    \driverC(b,e,\overline{\V{op}_l})\!
+         \vdash \\
+         \forall W \Wpriv \Winit \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
+        \end{array}} \right. \\   \X{bootSpec}^{\X{sem}}(P,g,b,e,A_{\X{entry}}, \Winit) \\
   (r_0, m_0, \emptyset) \longrightarrow^* (r, m, t)\\
@@
 \end{remark}
-
+\begin{remark}
+ Dominique made the remark that we might want to allow capabilities pointing
+ into the driver code that are not enter capabilities, but can just generally be
+ proven safe wrt.\ the driver code. This scenario (which we currently do not
+ support out of simplicity considerations) would force us to add the driver code
+ as a static region to the world. While this would work on paper, it does not in the
+ current Coq formalization, since all regions that standard regions point to are
+ forced to be standard as well (through the conditions in the different cases of
+ the $\I{interp1}$ predicate). Essentially, this behavior is disallowed because
+ the Coq development merged the notion of read- and write condition, and there
+ is hence no room to define a static region which defines a stricter reading
+ policy, but hence disallowes writing. More generally, it is currently
+ impossible to plug non-standard regions into the world, and allow standard
+ regions to depend on them in any way whatsoever. This was not a problem up
+ until now, since our examples never required this sort of dependency.
+\end{remark}
 \subsubsection{Concrete boot and driver code}

### 2020-03-31 capability-meetings-notes/10-03/main.tex
mmio meeting notes: \VR and \ER: fix the order of arguments (world goes first)

@@
 \[
-  \VR^{\Psi}(p, g, b, e, a) \eqdef \underbrace{\ldots}_{\text{as before}} \ast \; \pure{\forall a' \in [b,e).\; \Psi(a')}
+  \VR^{\Psi}(W)(p, g, b, e, a) \eqdef \underbrace{\ldots}_{\text{as before}} \ast \; \pure{\forall a' \in [b,e).\; \Psi(a')}
 \]
@@
          \vdash
-         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
+         \forall W \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
        \\
@@
   Note that we could have written
-  $\ldots \vdash \forall W \ldotp \; (\VR^{\notMMIO} (\X{E}, \ldots))\; W$ %)
+  $\ldots \vdash \forall W \ldotp \; \VR^{\notMMIO} (W) (\X{E}, \ldots)$
   in the above theorem, instead of using the expression relation. This is
@@
          \vdash \\
-         \forall W \Wpriv \Winit \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
+         \forall W \Wpriv \Winit \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
         \end{array}} \right. \\
@@
          \vdash \\
-         \forall W \Wpriv \Winit \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
+         \forall W \Wpriv \Winit \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
         \end{array}} \right. \\   \X{bootSpec}^{\X{sem}}(P,g,b,e,A_{\X{entry}}, \Winit) \\

### 2020-03-31 capability-meetings-notes/10-03/main.tex
Some notes on potential clients

--- 2020-03-31 capability-meetings-notes/10-03/main.tex
@@
 \usepackage{xargs}
+\usepackage{listings}
 
@@
 \DeclareMathOperator{\driverC}{code}
+\newcommand{\wrap}[1]{\operatorname{wrap}_{\mathrm{#1}}}
 
@@
 \section{Potential driver clients/properties}
-The concrete driver that we described in \texttt{subsec:driverimpl} does not yet enforce any useful
+The concrete driver that we described in \ref{subsec:driverimpl} does not yet enforce any useful
 safety properties on the traces that it outputs.
@@
 
+To only print values $\le 1000$, we need to wrap the driver's read method in a
+function that checks this condition, before passing control to the actual read
+method.
+On an abstract level, given a function $f$, we would have the following wrapper:
+\[
+  \wrap{\le 1000} f \eqdef \lambda x\ldotp \X{assert}(x \le 1000); f\; x
+\]
+, and the closure we share with the environment is hence $\wrap{\le
+  1000}(\X{read})$ instead of just $\X{read}$.
+
+A trivial misbehaving client could look as follows:
+(TODO: ask Dominique for Lau's code on environments)
+\begin{lstlisting}
+  move r1 1001
+  jmp r2
+\end{lstlisting}
+It simply loads an argument to write and jumps to the write routine.
+
+A well-behaved client might look as follows:
+\begin{lstlisting}
+  move r1 0
+  jmp r2
+\end{lstlisting}
+
 \paragraph{Modelling channel-specific safety properties}
+Here, we could encode different channels into our integers (in the spirit of
+what Frank mentioned), and enforce different safety properties on the different
+channels (by ensuring that we can split $P$).
+On a conceptual level, this does not require any changes whatsoever.
 
@@
 instead of having a pure predicate $P$ in our robust safety theorem, we should
-also allow non-pure predicates that entail $P$, since we might have to enforce
+also allow non-pure predicates $P_{\X{np}}$ that entail $P$, i.e. $\forall t
+\ldotp P_{\X{np}}(t) \wand \pure{P(t)} $, since we might have to enforce
 other invariants relating to local state over the trace as well.
@@
 To make sure we can prove the driver specification, we will have to update the
-robust safety theorem slightly to incorporate these changes. It will then look
-as follows:
+robust safety theorem slightly to incorporate these changes.
+
+Given a macro instruction to perform loops, a well/mis-behaving client might
+look as follows: (TODO: write this out,
+and come up with some macro formulation)
 
 \paragraph{Never performing I/O again after some stop token has been read on $\stdin$}
+Again, this requires no extra conceptual changes.
+
+\paragraph{Verifying OCP's}
+As Arma\"el noted at some point, it should be possible to simulate the
+proof-style we see in OCPL by emitting values onto the trace, instead of making
+assertions over them. This should give us the same proving power as the OCPL
+paper, and allow us to emulate their examples.
+
+\section{Proof sketch}
+This section describes the steps involved in proving the robust safety theorem
+stated above in some more detail.
 

### 2020-03-31 capability-notes/driver-notes/main.tex
Moved meeting notes to new folder; created driver notes folder

--- 2020-03-31 capability-meetings-notes/10-03/main.tex
@@
 \section{Potential driver clients/properties}
-The concrete driver that we described in \ref{subsec:driverimpl} does not yet enforce any useful
-safety properties on the traces that it outputs.
-This section describes a couple such properties that we might want to enforce,
-modifications to the driver that they require and example clients that make use of our modified driver.
+The concrete driver that we described in Section~\ref{subsec:driverimpl} does
+not yet enforce any useful safety properties on the traces that it outputs. This
+section describes a couple such properties that we might want to enforce,
+modifications to the driver that they require and example clients that make use
+of our modified driver.
 

### 2020-03-31 capability-notes/meeting-notes/17-03/main.tex
Moved meeting notes to new folder; created driver notes folder

--- 2020-03-25 capability-meetings-notes/25-03/main.tex
@@
 
-\section{Discussions about the notes}
+\section{Talk of the day}
+
 \begin{itemize}
-\item \underline{Uninitialised capabilities}
+\item A\"ina has finished proving a version of the awkward example without the preamble and with stack clearing upon return.
 
-We agree that it looks good!
-Arma\"{e}l suggested it might be easier to make store initialise instead of storeu (where it currently increments it if we are on top).
-Would it make it complicated to just have store do the increment instead?
-proposal: not have a special case for storeU, but rather have a special case for store for U- capabilities that do the increment.
-The intuition is that we offset the load for proving everything in storeU, and split the work more evenly between store and storeU.
-Another alternative, is to enable store to write to U- capabilities, but have storeU be the specific store + increment.
-(Related work is making sure that all the regular stores be replaceable with storeU). 
-For now we leave it like it is, but if the FTLR for storeU becomes too big, we might want to consider a split.
+\item Alix, Arma\"el and Thomas have finished the refactoring.
 
-\item \underline{Device drivers}
+\item Alix has rebased the uninitialized capabilities branch over the refactoring and started fixing proofs.
 
-Comments from Lars: 
-\begin{itemize}
-	\item The semantics are non deterministic because of the IOload, we should take note of all sources of non-determinism since they can cause trouble later on. 
-	\item What is the simplest possible client program that one could consider?
-where one can think about instantiating P with something concrete (e.g.
-even numbers).
-It might be worth considering, since it forces us to compose these different parts.
-The original client Dominique presented is that we are not accessing certain parts of memory.
-A similar but contrived examples is that every element of the trace does not go above a certain threshold.
-Even though P is simple, it is interesting to see how these specs might fit together. 
-	\item Reactive programming uses streams of outputs. An alternate to the non determinism is that Load takes the next element in some oracle stream.
-
-Non determinism can be one way of modelling input, streams could be an alternative.
-In CakeML there is an abstract type for the state of the outside world, and an oracle and operations that change the state of the outside world.
-Whenever an IO operation happen, we keep track of it in a trace.
-This means we keep a model of the external world in the semantics.
-Which model is going to make the rest of the story the simplest here? 
-What is the impact of non determinism?
-Here it’s still internally deterministic, but it’s with the external interactions with the world that non-determinism happen.
-There might a way to relax the determinism restriction for the WP lemmas we will need.
-To make it deterministic, we can make x be a parameter of the step (this parameter can be given by some kind of oracle).
-
-Non determinism can lead to challenged down the road. We should be aware of this!
-
-Let’s consider making it deterministic using some kind of oracle to avoid such challenges. 
+\item Thomas has given a presentation of what the driver code that would be verified could look like.
 \end{itemize}
 
-Discussion about the spec on page 10/11: 
-\begin{itemize}
-\item If the boot code is doing IO, then we also need to state that the boot code produces a trace t for which P(t) holds.
-\item We assume that code can produce multiple entry and exit points. Proposed alternative: one entry point and a registers that select what kind of thing we want to do. 
-\item It might make things easier to read if we use notation for the fragmental trace resources. 
-\item Some comments from Dominique:
-
-Here the theorem does not make any assumptions about memory.
-(here we know nothing about memory.
-If memory contains any capabilities, the boot code will need to clear memory in order to satisfy the spec.
-What if we do know that memory is all 0’s at boot?
-then we cannot prove that the boot satisfies the step without clearing.
-This might be an issue?
-We need to mention that as a guarantee in the theorem.
-Let’s figure out what happens in real life before we add such an assumption).
-
-spec on page 11: universal quantification over all W.
-No link between the world the boot code sets up and the world the expression relation is prover over.
-We need to consider all private future worlds of the world the boot code allocates.
-
-page 10: we will need some assumptions over the memory that the entry points are valid.
-
-can we generalise the entry points to be any permission, not just E-capabilities (i.e.
-closures).
-We could generalise the statement by stating validity.
-This might simplify the condition, the PC is kept as only of the registers.
-
-\item Some comments from Frank: Does the boot code need to do IO?
-depending on the goal of the work: if we want to model what OS boot code does on IO, then yes, boot code does initialisation of devices which might leave a trace).
-Do we want to see this evolve into proving an actual boot code correct?
-Or do we want to focus on something where more fancy Iris stuff is needed.
-These might be two different directions for the project.
-\end{itemize}
-
-\end{itemize}
-
-\section{Next steps for the driver}
+\section{Things to do for next meeting}
 
 \begin{itemize}
-	\item fix the statement of the theorem such that the world produced by the boot code is linked to the world the expression relation must hold over. 
-	\item think of a client and a proof sketch of that minimal client (on paper) to get an idea of how these moving parts fit together.
-\end{itemize}
+\item A\"ina will give a tour of the Coq code on \displaydate{friday} to showcase how to go through the examples.
+She will also give a higher level presentation of the awkward example, and how to reason about the world/revocation in the Iris setting next meeting.
 
-\section{Next steps for the uninitialized capability}
+\item Arma\"el and Thomas will write down something about the operational semantics for I/O, and the expected changes to the value relation.
 
-\begin{itemize}
-	\item prove the awkward example without clearing the stack upon return. This step will need further discussions (Aina will send an email thursday if such a discussion can already happen on friday)
-	\item start proving WP's and FTLR for the now implemented semantics 
+\item Alix will write down something about uninitialized capabilities, how they are modeled, what is impacted by their introduction, and try to continue working on the uninitialized capabilities branch.
 \end{itemize}

### 2020-03-31 capability-notes/driver-notes/main.tex
Some minor notational fixes

@@
 \newcommand{\I}[1]{\ensuremath{\mathtt{#1}}}
+\newcommand{\Sf}[1]{\ensuremath{\mathsf{#1}}}
 \newcommand{\T}[1]{\texttt{#1}}
@@
 
-\newcommand{\MMIO}{\textlog{MMIO}\xspace}
+\newcommand{\MMIO}{\Sf{MMIO}\xspace}
 
@@
 \DeclareMathOperator{\driverC}{code}
+\DeclareMathOperator{\trace}{tr}
+\DeclareMathOperator{\lowval}{\Sf{lowval}}
+\DeclareMathOperator{\lifto}{\Sf{lift}}
+\newcommandx{\trInv}[2][1=P,2=\iota]{\operatorname{trInv}^{#2}(#1)}
 \newcommand{\wrap}[1]{\operatorname{wrap}_{\mathrm{#1}}}
-
 \newcommand{\bigast}[2]{\underset{#1}\Sep{\!\!\! #2}\;}
@@
     e \; \X{closed} \\
-    \hoare{\TRUE}{e}{x.\; \textlog{lowval}\; x} \\
-    (C[e]);(\emptyset, \textlog{OK}) \longrightarrow^* T';(h',g')
+    \hoare{\TRUE}{e}{x\ldotp \lowval x} \\
+    (C[e]);(\emptyset, \Sf{OK}) \longrightarrow^* T';(h',g')
   }
-  {g' = \textlog{OK}}
+  {g' = \Sf{OK}}
 \end{mathpar}
@@
 state, then we can observe that every reachable state is good
-($g' = \textlog{OK}$ means that no assertion has failed in $e$).
+($g' = \Sf{OK}$ means that no assertion has failed in $e$).
 
@@
 trivially contradict the theorem by taking
-$C[\cdot] = \textlog{assert false}$), and cannot contain references to raw
+$C[\cdot] = \Sf{assert}\spac \X{false}$), and cannot contain references to raw
 memory locations (otherwise one could directly access $e$'s private state,
@@
 
-\newcommand{\lift}[2]{\textlog{lift}\; #1 \; #2}
+\newcommand{\lift}[2]{\lifto #1 \: #2}
 \newcommand{\liftP}[1]{\lift{\Psi}{#1}}
 
-$\textlog{lowval}\; x$ intuitively means that $x$ cannot be used to directly
+$\Sf{lowval}\; x$ intuitively means that $x$ cannot be used to directly
 access private (or ``high'') locations. It is formally defined using a logical
@@
   \begin{array}{lcl}
-    \liftP (\textlog{rec}\; f\; x.\; e) & \eqdef
-    & \later \forall v.\; \hoare{\liftP{v}}{e[v/x,\textlog{rec}\;f\;x.\; e/f]}{y.\; \liftP y} \\
+    \liftP (\Sf{rec}\; f\; x.\; e) & \eqdef
+    & \later \forall v.\; \hoare{\liftP{v}}{e[v/x,\Sf{rec}\;f\;x.\; e/f]}{y.\; \liftP y} \\
     %
@@
 
-Then, $\textlog{lowval}\; x$ is defined as $\lift{\textlog{lowloc}}{x}$, where
-$\textlog{lowloc}$ is a predicate characterizing the region of memory containing
+Then, $\lowval x$ is defined as $\lift{\Sf{lowloc}}{x}$, where
+$\Sf{lowloc}$ is a predicate characterizing the region of memory containing
 ``low locations'' (distinct from the other region containing ``high
@@
 
-\newcommand{\OutV}{\textlog{Out}}
+\newcommand{\OutV}{\Sf{Out}}
 
 In Thomas' extension of OCPL, a new value $\OutV$ is added, denoting an
-``output object capability'', as well as a $\textlog{print}$ primitive, where
-$\textlog{print}\; \OutV\; v$ effectively ``prints'' the value $v$, i.e. adds it
+``output object capability'', as well as a $\Sf{print}$ primitive, where
+$\Sf{print}\; \OutV\; v$ effectively ``prints'' the value $v$, i.e. adds it
 to the trace of printed values.
@@
     e \; \X{closed} \\
-    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast \pure{\! P(t)}} \vdash \hoare{\TRUE}{e}{x.\; \textlog{lowval}\; x} \\
-    (C[e]);(\emptyset, \textlog{OK}); \emptyset \longrightarrow^* T';(h',g'); t
+    \trInv \vdash \hoare{\TRUE}{e}{x.\; \operatorname{\Sf{lowval}} x} \\
+    (C[e]);(\emptyset, \Sf{OK}); \emptyset \longrightarrow^* T';(h',g'); t
   }
-  {g' = \textlog{OK} \wedge P(t)}
+  {g' = \Sf{OK} \wedge P(t)}
 \end{mathpar}
+, where we introduced the following notation:
+%
+\[
+  \begin{array}{lcl}
+    \trace(t) & \eqdef & \tracefrag{t}\\
+    \trInv & \eqdef & \knowInv{\iota}{\exists t.\; \trace(t) \ast \pure{\! P(t)}}
+  \end{array}
+\]
 
@@
 $\TRUE$ predicate. Technically, we could hence have the following alternative
-definition for $\textlog{lift}$ (although admittedly, it does look a bit
+definition for $\Sf{lift}$ (although admittedly, it does look a bit
 hard-coded):
@@
   \begin{array}{lcl}
-    \liftP \OutV & \eqdef & \exists \iota \ldotp \knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast
-                            \pure{(\Lam{\_}.\TRUE) \spac t}}
+    \liftP \OutV & \eqdef & \trInv[\Lam{\_}.\TRUE]
   \end{array}
@@
 
-Similarly to $\textlog{lift}$, we can generalize the existing logical relation
+Similarly to $\Sf{lift}$, we can generalize the existing logical relation
 to thread a constraint on directly accessible memory locations. One would
@@
 
-Then, $\VR^{\notMMIO}$ is similar to the $\textlog{lowval}$ predicate of OCPL.
+Then, $\VR^{\notMMIO}$ is similar to the $\lowval$ predicate of OCPL.
 Intuitively, a value in the relation does not directly point to memory-mapped
 locations, nor can it gain access to memory-mapped locations indirectly,
-similarly to how $\textlog{lowval}$s cannot grant access to high locations
+similarly to how $\lowval$s cannot grant access to high locations
 directly. The relation $\ER^{\notMMIO}$ specifies that, if we fill the registers
@@
 Given this definition, we can now formalize the initialization constraints on
-registers $r$ and memory $m$, given the region $\MMIO$ and
+registers $r$ and memory $m$, given the region \MMIO and
 the driver's address space $[b,e)$ as follows (for a resource-aware formulation;

### 2020-03-31 capability-notes/driver-notes/main.tex
Added comment by Frank

@@
   (or, in this case equivalently, an invariant). We can write this invariant as
-      $\driverC(b,e,\overline{\V{op}_l}) = \knowInv{\iota}{\bigast{l \in [b,e)}{l \mapsto_{a} \mathit{\V{op}_l}}}$, with $\V{op}_l$
-      some hard-coded driver instruction at location $l$.
+      $\driverC(b,e,\overline{\V{op}_l}) = \knowInv{\iota}{\bigast{l \in [b,e)}{l \mapsto_{a} \mathit{\V{w}_l}}}$, with $\V{w}_l$
+      some hard-coded driver word (an instruction or a capability) at location $l$.
 \end{itemize}
@@
     \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\ast
-    \driverC(b,e,\overline{\V{op}_l})\!
+    \driverC(b,e,\overline{\V{w}_l})\!
          \vdash \\
@@
 \]
+
+\begin{remark}
+As Frank mentioned, if we do not enforce any validity constraints on the part of the adversary's memory that is not reachable
+from the valid registers themselves, the driver cannot read from this memory in
+a useful way during execution (since the memory might contain any value whatsoever).
+There is currently no concrete use case for this scenario yet, so we left such
+condition on the adversary's memory out, but it is good to be aware of.
+\end{remark}
 

### 2020-03-31 capability-notes/driver-notes/main.tex
Updated notation for (im)pure invariants over traces

@@
 \DeclareMathOperator{\lifto}{\Sf{lift}}
-\newcommandx{\trInv}[2][1=P,2=\iota]{\operatorname{trInv}^{#2}(#1)}
+\newcommandx{\trInv}[3][1=P,2=\iota,3]{\operatorname{trInv}^{#2}_{#3}(#1)}
+\newcommandx{\trInvNp}[2][1=P,2=\iota]{\trInv[#1][#2][\mathrm{np}]}
 \newcommand{\wrap}[1]{\operatorname{wrap}_{\mathrm{#1}}}
@@
   { \forall a_{i} \in A_{\X{entry}} \ldotp\,
-      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\!
+      \trInv\!
          \vdash
@@
   (or, in this case equivalently, an invariant). We can write this invariant as
-      $\driverC(b,e,\overline{\V{op}_l}) = \knowInv{\iota}{\bigast{l \in [b,e)}{l \mapsto_{a} \mathit{\V{w}_l}}}$, with $\V{w}_l$
+      $\driverC(b,e,\overline{\V{w}_l}) = \knowInv{\iota}{\bigast{l \in [b,e)}{l \mapsto_{a} \mathit{\V{w}_l}}}$, with $\V{w}_l$
       some hard-coded driver word (an instruction or a capability) at location $l$.
@@
     {\begin{array}{ll}
-    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\ast
+    \trInv \ast
     \driverC(b,e,\overline{\V{w}_l})\!
@@
     {\begin{array}{ll}
-    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\ast
-    \driverC(b,e,\overline{\V{op}_l})\!
+    \trInv \ast
+    \driverC(b,e,\overline{\V{w}_l})\!
          \vdash \\
@@
 driver. Concretely, the invariant we will verify our driver under changes from
-$\knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast \pure{|t| \le 1000}}$ to
-$\knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast \pure{|t| \le 1000} \ast
-  \exists n\ldotp l_n \mapsto n \ast n = |t| }$. Concretely, this means that
+$\trInv[|t| \le 1000]$ to
+$\trInvNp[\pure{|t| \le 1000} \ast
+  \exists n\ldotp l_n \mapsto n \ast n = |t| ]$, where we defined
+%
+\[
+  \begin{array}{lcl}
+    \trInvNp[P_{\X{np}}] & \eqdef & \knowInv{\iota}{\exists t.\; \trace(t) \ast P_{\X{np}}(t)}
+  \end{array}
+\]
+
+(i.e., the predicate over the trace does not need to be pure any longer).
+Concretely, this means that
 instead of having a pure predicate $P$ in our robust safety theorem, we should

### 2020-04-01 capability-notes/driver-notes/main.tex
First attempt at writing a semantic bootSpec

@@
 %
-Then, the syntactic specification for the boot code that we have to prove is stated as follows:
+Then, the syntactic specification for the boot code that we have to prove is
+stated as follows (we give both a specific, syntactic specification
+\X{bootSpec_{synt}} and a more
+general, semantic one \X{bootSpec_{sem}} that should follow from the syntactic specification):
 
@@
   \begin{array}{l}
-  \X{bootSpec}(P, g, b, e, A_{\X{entry}}) \eqdef {} \\[0.8em]
+
+  \X{bootSpec}(P, g, b, e, A_{\X{entry}},\V{bootCont}) \eqdef \\[0.8em]
     \quad
@@
       \ast \bigast{m \in \X{Mem} \setminus \MMIO}{m \mapsto_{a} m_{0}[m]}
-      \ast \; \tracefrag{\emptyset}
+      \ast \; \tracefrag{\emptyset} \ast \\[1.2em]
+      \later  \V{bootCont}
     \end{array}
     \right\}
-    \\[2em]
+    \\[2.5em]
     \quad \I{Seq}\; (\I{Instr}\; \I{Executable})
-    \\[0.5em]
-    \left\{
+    \\[0.2em]
+    \left\{ \Phi \right\}
+  \end{array}\\[4em]
+
+    \X{bootCont_{synt}}(P, g, b, e, A_{\X{entry}},\Phi) \eqdef  \\[0.2em]
     \begin{array}{l}
-      \exists b_{adv}\; e_{adv}\; t.\; \\[0.5em]
+       (\exists b_{adv}\; e_{adv}\; t.\; \\[0.5em]
       \quad
@@
         \bigast{m \in [b_{adv}, e_{adv})}{m \mapsto_{a} \X{inl}\; \_} \ast
-        \bigast{m \in [b, e)}{m \mapsto_{a} \_} \ast
-        \; \tracefrag{t} \ast \pure{P(t)}
-      \end{array}
-    \end{array}
-    \right\}
-  \end{array}
-  \end{array}
-\]
+        %\bigast{m \in [b, e)}{m \mapsto_{a} \_} \ast
+        \driverC(b,e,\overline{\V{w}_l})\ast
+        \; \tracefrag{t} \ast \pure{P(t)})
+      \end{array}
+    \end{array} \\[4em]
+    \quad\wand
+    \begin{array}{l}
+     \wpre{\I{Seq}\; (\I{Instr}\; \I{Executable})}{\Phi}
+    \end{array}\\[1.2em]
+
+  \X{bootSpec_{synt}}(P, g, b, e, A_{\X{entry}}) \eqdef\\
+     \quad \forall \Phi\ldotp  \X{bootSpec}(P, g, b, e, A_{\X{entry}},\X{bootCont_{synt}}(P, g, b, e, A_{\X{entry}},\Phi))\\[0.8em]
+
+ \X{InitW}(\Winit,b,e) \eqdef\\
+\quad \text{Predicate that forces  $\Winit$ to be the world with one
+    permanent, standard region at $[b,e[$}\\[0.8em]
+
+ \X{bootCont_{sem}}(P, g, b, e, A_{\X{entry}},\Phi,\Winit) \eqdef  \\[0.2em]
+    \begin{array}{l}
+       (\exists b_{\X{adv}}\; e_{\X{adv}}\; t.\; \\[0.5em]
+      \quad
+      \begin{array}{l}
+        \X{InitW}(\Winit,b_{\X{adv}}, e_{\X{adv}})\ast \\[0.5em]
+        \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{\X{PC}}, e_{\X{PC}}, a_{\X{PC}})
+        \ast \VR^{\notMMIO} (\Winit)(\X{RW}, \X{G}, b_{\X{PC}}, e_{\X{PC}}, a_{\X{PC}})) \ast {} \\[1.2em]
+        \bigast{r \in Reg\setminus \{\X{PC}\}}{(r \mapsto_{r} w_r \ast
+        \VR^{\notMMIO} (\Winit)(w_r))} \ast {} \\[1.2em]
+        \driverC(b,e,\overline{\V{w}_l})\ast
+        \; \tracefrag{t} \ast  \pure{P(t)}\ast \\[0.5em]
+        \X{region}(\Winit) \ast  \X{sts\_full\_world}\; \X{sts\_std} \;
+        \Winit)\\[0.5em]       \end{array}
+    \end{array} \\[4em]
+    \quad\wand
+    \begin{array}{l}
+     \wpre{\I{Seq}\; (\I{Instr}\; \I{Executable})}{\Phi}
+    \end{array}\\[1.2em]
+
+  \X{bootSpec_{sem}}(P, g, b, e, A_{\X{entry}},\Winit)
+    \eqdef\\
+     \quad \forall \Phi\ldotp  \X{bootSpec}(P, g, b, e, A_{\X{entry}},\X{bootCont_{sem}}(P, g, b, e, A_{\X{entry}},\Phi,\Winit))\\[0.8em]
+
+  \end{array}
+\]
+
+\begin{remark}
+  We will have to allocate the above predicates\\ $\X{region}(\Winit) \ast
+  \X{sts\_full\_world}\; \X{sts\_std} \; \Winit$ from our Coq development under
+  a different ghost name $\gamma$ than we usually do.
+  I do not know how exactly that would work on a technical level, but it should somehow.
+\end{remark}
 
@@
 from the valid registers themselves, the driver cannot read from this memory in
-a useful way during execution (since the memory might contain any value whatsoever).
-There is currently no concrete use case for this scenario yet, so we left such
-condition on the adversary's memory out, but it is good to be aware of.
+a useful way during execution (since the memory might contain any value
+whatsoever, \'and we cannot obtain a points-to chunk to it).
+There is currently no concrete use case for this scenario yet, but it is good to be aware of.
+In the general case, we would have to make sure that the parts of the adversary
+code that are not pointed to by registers but are needed by the driver for some
+reason are present in the world (otherwise, we have no way to obtain the
+points-to chunk either), or present in an invariant.
 \end{remark}

### 2020-04-02 capability-notes/driver-notes/main.tex
driver notes: move most of the remarks to an appendix to ease the reading of the main document

@@
 
+\newcommand{\rk}[1]{\hyperref[{rk:#1}]{\textbf{[Rk~\ref*{rk:#1}]}}}
+
 \title{Capabilities, MMIO and Robust Safety}
@@
 Notice how in the second rule, we read an arbitrary integer $x$, and record it
-in the trace.
-
+in the trace~\rk{opsem:nondet}.
+%
 The \I{Store} rule would be similar.
@@
 \newcommand{\stdin}{\X{in}}
-\begin{remark}
-As Lars remarked, the read rule introduces non-determinism into our operational
-semantics.
-This has lead to some complications in the past (the nature of which is still
-mysterious to me; n.b. that we use\\ $\I{ wp\_lift\_atomic\_head\_step\_no\_fork}$
-in our WP rules, and that one would allow non-determinism).
-(one question I -Thomas- had in this respect; does the $\I{Alloc}$ rule in HeapLang
-not introduce non-determinism in a similar way? Why is this not problematic?)
-Lars' suggestion was to additionally parameterize the operational semantics by
-an input-stream $\stdin$.
-The updated \textsc{IOLoad} load rule might then look something like this:
-\begin{mathpar}
-  \inferrule[IOLoad]
-  { r[\X{src}] = (p,g,b,e,a) \\ \X{readAllowed}\; p \\ a \in [b,e) \\ a \in \MMIO
-  \\ n = \#\{i \mid t[i] = ( \X{IORead},\_\;,\_) \}  }
-  { (r, m, t) \xrightarrow{\I{Load} \; \X{dst} \; \X{src}}
-    (r[\X{dst} := \stdin[n]], m, t \app (\X{IORead}, a, \stdin[n])) }
-\end{mathpar}
-
-When parameterizing our development with multiple values, we could either write
-them out as different Coq \texttt{Parameters} where we need them, or store them
-as fields in our \texttt{DriverG} typeclass (the latter is the approach taken
-in e.g.\ the code-base of ~\cite{sammler-2020}).
-\end{remark}
 
@@
 \begin{enumerate}
-\item Include the \MMIO addresses into the map. Notice that, for any address in \MMIO, the specific value held by the map $m$ for
-that address is now irrelevant. Indeed, $m$ is never read nor modified for
-addresses in \MMIO (possible exception: see next remark). One could choose to enforce that a particular dummy value is stored in $m$ for these addresses. Instead, we choose to leave them
-unconstrained.
+  \item Include the \MMIO addresses into the map. Notice that, for any address
+    in \MMIO, the specific value held by the map $m$ for that address is now
+    irrelevant. Indeed, $m$ is never read nor modified for addresses in \MMIO
+    (possible exception: see \rk{opsem:pc_in_mmio}). One could choose to enforce
+    that a particular dummy value is stored in $m$ for these addresses or to
+    leave them unconstrained.
+%
   \item Have $m$ model the physical RAM addresses only. This model is closer to
@@
     require us to talk about the values in the heap that \MMIO locations are
-  associated with. In case it is not harder to implement (which I currently do not think
-  it is), it is to be preferred.
+    associated with. It however probably requires threading through the
+    operational semantics the invariant that $m$ is disjoint from \MMIO.
 \end{enumerate}
+%
 Both approaches require us to make sure that no points-to chunks $l \mapsto \_$
 with $l \in \MMIO$ can be created. In the next section, we assume the second
-approach (until someone proves me wrong about it working).
-
-\begin{remark}
-  Additionally, we have the option of disallowing an execution
-  with a \X{PC} register pointing to a $\MMIO$ location in the semantics.
-  First off, consider that this is an edge case; since we always have points-to
-  chunks describing the layout of code in memory (and a points-to chunk for
-  location $l$ doubles as a proof that $l \not\in \MMIO$), it is trivial to prove that
-  this case would not occur in any of our current examples.
-  If we make no changes to the current code base, this behavior would be
-  allowed, and the following would happen in both cases described above:
-  \begin{enumerate}
-  \item An unconstrained value is read from the location. Any instruction could
-    be executed.
-  \item The predicate $\I{MemLocate}$ in $\I{lang.v}$ will make sure
-  that the default value $0$ is read and decoded into an instruction when we try
-  to execute an \MMIO location.
-  This would not cause a conflict with the current WP rules,
-  since a points-to chunk is required in the precondition, ensuring that this
-  case does not occur.
-  Another alternative is to read the encoding of a $\X{NOP}$ instruction, instead of
-  just the value $0$.
-\end{enumerate}
-  If we were to simply disallow execution when the PC points into \MMIO, we would make the semantics of all instructions slightly more complicated, but in a
-  uniform way.
-  We could simply add $a \not\in \MMIO$ to the predicate $\I{isCorrectPC}$ (or
-  add it as an additional predicate), which is used to define failing execution.
-  It would again not be a drastic change, since a points-to for the
-  address $a$ the PC points to is required by the WP rules regardless, from which we
-  can prove that $a \not\in \MMIO$.
-\end{remark}
-
-\begin{remark} With this presentation, the specification of the
-``machine'' is very much decoupled from the model of the devices it might be
-communicating with. This is a good thing, I think.
-
-Nevertheless, one could consider an alternative presentation where the model of
-the devices is more tightly integrated with the semantics of the machine. For
-instance, one could make the operational semantics parameterized with the
-devices' model, where each device is modeled as having some internal state, the
-ability to react on reads or writes, or to perform an internal step. Then, the
-operational semantics would either step the usual way, or whenever a device
-steps.
-
-I believe this would be somewhat similar to the semantics of I/O system calls
-through a foreign function interfaces as formalized in
-CakeML~\cite{cakeml-vstte17io}, and also in Perennial~\cite{perennial-lang}.
-%
-
-\end{remark}
-
-\begin{remark} Alix says that the proposed style of operational semantics
-(using a trace) is close to the semantics of \texttt{volatile} as in
-CompCert---which is also how memory mapped addresses seem to be exposed to C
-compilers in practice. So this is probably a good sign. 
-\end{remark}
+approach~\rk{opsem:pc_in_mmio}.
 
@@
 monolithic fashion. In that case, the state interpretation additionally holds a
-resource for the trace:
+resource for the trace~\rk{resources:mmio_in_m}:
 
@@
 \end{align*}
-
-\begin{remark}
-For the case where the \MMIO locations are still part of the heap $m$, this
-would look as follows:
-\[
-  \X{state\_interp} \; (r,m,t) :=
-    \X{gen\_heap\_ctx}\; r \ast
-    \X{gen\_heap\_ctx}\; (m \setminus \MMIO) \ast
-    {\color{BrickRed}\tracefull{t}}
-\]
-
-Sanity check: we never have allocation of new locations $l$ in our capability machine, but if we
-did, that would work in both settings by requiring that $l \not\in \MMIO$ in the
-operational semantics rule for \textsc{Alloc}.
-\end{remark}
 
@@
 
-The corresponding resource algebra is $\authm(\exm(\X{Trace}))$. A few relevant
+The corresponding resource algebra is
+$\authm(\exm(\X{Trace}))$~\rk{resources:fine_grained_trace}. A few relevant
 rules are:
@@
 
-\begin{remark} This is very coarse-grained: either one has the full
-ownership for performing I/O and reasoning about it, or one cannot know anything
-about it.
-
-A first extension could be to allow observing prefixes of the trace (since
-events can only be appended to the trace). The observation that the trace
-has a given prefix would be duplicable. This again seems to be an application of
-monotonicity, that could be realized by having a duplicable AtLeast part as part
-of $\tracefrag{t'}$,
-similarly to how we currently model Monotone References.
-
-Another extension, that seems very useful for modular reasoning, would be to
-allow splitting the trace along separate range of addresses. Concretely, a trace
-containing events about the range of MMIO addresses $[a,c)$ could be split into
-two traces, granting ownership over events on addresses $[a,b)$ and $[b,c)$
-respectively. Note that recombining these two traces would only yield some
-unspecified interleaving of the events from the two traces, and not necessarily
-yield the original trace, since in our model, we cannot know the exact interleaving of IO operations issued by different parties.
-
-One application of this second extension could be the verification of an example
-involving ``multiplexed'' I/O, where two separate parts of the code are granted
-separate ownership over separate MMIO addresses. These two separate pieces of
-code would be verified separately; then, in the end, one could prove that one
-gets \emph{some} interleaving of all the events emitted by both components.
-\end{remark}
+To simplify notation, let us introduce the following definitions for user-level
+ownership over the trace, and preservation of an invariant on the trace,
+respectively:
+
+\[
+  \begin{array}{rcl}
+    \trace(t) & \eqdef & \tracefrag{t} \\
+    \trInv[P][\iota] & \eqdef & \knowInv{\iota}{\exists t.\; \trace(t) \ast \pure{P(t)}}
+  \end{array}
+\]
+
+\rk{resources:mmio_in_heap} \rk{resources:dynamic_mmio}
 
 \newcommand{\MMIOag}{\ownGhost{\gamma_{\X{MMIO}}}{\authfull (\X{MMIO})}}
-
-\begin{remark}
-  As an alternative to having the set of MMIO addresses (\MMIO) as a global
-  parameter, we might want to make it part of the state in the operational
-  semantics. In that case, we would need another resource algebra to model the
-  MMIO locations, i.e. add do the state interpretation we had before:
-%
-\[
-  \X{state\_interp} \; (r,m,t,\MMIO) := \ldots \ast \MMIOag
-\]
-
-(where $\authm(\exm(\X{Trace}))$ is the resource algebra).
-\end{remark}
-
-\begin{remark}
-In the future, we might be interested in the dynamic allocation of MMIO memory.
-The question is what this would mean, though, since the set of available devices
-allowing for MMIO access and the concrete buffers they provide will most likely
-still be modeled as fixed at runtime.
-Rather, this would allow us to take (un)mapped MMIO locations, connected to the different devices, and (un)map them anywhere in main memory.
-This would require a different way of modeling MMIO, where we need both a notion
-of the total pool of possible MMIO locations, and a description of the currently
-mapped locations. We could model this using an authoritative RA.
-\end{remark}
 
@@
 \end{mathpar}
-, where we introduced the following notation:
-%
-\[
-  \begin{array}{lcl}
-    \trace(t) & \eqdef & \tracefrag{t}\\
-    \trInv & \eqdef & \knowInv{\iota}{\exists t.\; \trace(t) \ast \pure{\! P(t)}}
-  \end{array}
-\]
 
@@
 of our starting world will be satisfied at the end of execution.
-
-\begin{remark} As it is defined above, the ``generalized'' value relation
-is in fact not very useful for instantiation of $\Psi$ other than $\notMMIO$.
-Indeed, even if $\Psi$ \emph{does} allow referring to addresses in the $\MMIO$
-region, the value relation does not grant any corresponding resources. The fix
-would be to use the generalized trace resources mentioned previously, and grant
-ownership for the part of the trace corresponding to the addresses in
-$[b,e) \cap \MMIO_{\X{pub}}$ with $\MMIO_{\X{pub}}$ the shareable subset of
-addresses in \MMIO.
-\end{remark}
+%
+\rk{toplevel:v_psi}
 
@@
 the code we want to gradually verify.
+%
+\rk{toplevel:single_entrypoint}
 
@@
 called.
-
-\begin{remark}
-  We could simplify the theorem statement by only considering the case of a
-  single entrypoint. This is equivalent: a piece of code that wants to expose
-  two distinct ``methods'' can implement a single toplevel entrypoint that
-  immediately returns two pointers that can then be used to invoke the two
-  methods.
-  %
-  This would make the toplevel theorem simpler, at the cost of somewhat more
-  contrieved calling convention between trusted and untrusted code.
-  Another alternative is to provide an opcode to specify the operation that
-  needs to be performed.
-  This is slightly less general, since it does not allow providing read access
-  only to an adversary.
-\end{remark}
-
-\begin{remark}
-The universal quantification on worlds was not explicitly present in the OCPL
-formalization.
-The reason for that is that (I think so at least, do not quote me
-on this - Thomas) they
-used the built-in Iris worlds within their definition of weakest precondition,
-by leveraging the state interpretation, which is universally quantified over in
-the definition of weakest precondition.
-As we discussed before, if we do away
-with local capabilities, it should be possible to fall back onto Iris' notion of
-worlds, which would allow us to remove the universal quantification in the
-above definition too because it would be implicitly present in the WP inside $\ER$.
-
-It would be interesting to have a discussion about how to do away with the
-worlds if we remove local capabilities, regardless of whether we will in the
-end, since that might shed some light on why the worlds look the way they do
-right now.
-\end{remark}
-
-\begin{remark}
-  Note that we could have written
-  $\ldots \vdash \forall W \ldotp \; \VR^{\notMMIO} (W) (\X{E}, \ldots)$
-  in the above theorem, instead of using the expression relation. This is
-  equivalent, and easily unfolds to the former statement through the definition
-  of the value relation. I wrote $\ER$ to place the focus on the execution of
-  the driver being safe. I consider this a matter of taste.
-\end{remark}
+%
+\rk{toplevel:ocpl_explicit_worlds}
 
@@
 \end{mathpar}
+
+\rk{toplevel:less_restricted_syntactic_check_adverary}
 
@@
 requirement to $\initOKo$).
-(TODO: introduce notation for trace-owndership and predicate P holding)
-
-\begin{remark}
-  If useful, the syntactic check can be relaxed to allow any capabilities that
-  do not point into the MMIO region or the driver's memory. One would define the
-  predicate $\notMapsToR{R}{w}$, that takes a word, memory region or register
-  file and checks if it points into the forbidden region $R$ as follows:
-
-\[
-  \begin{array}{lcl}
-    \notMapsToR{R}{\X{inl}\spac z} & \eqdef
-    & \TRUE \\
-    %
-    \notMapsToR{R}{\X{inr}\spac (p,g,b,e,a)} & \eqdef & R \cap [b,e)\; =
-                                                 \emptyset  \\
-    %
-    \notMapsToR{R}{\V{reg}} & \eqdef & \forall r \in \dom(\V{reg}) \ldotp\spac \notMapsToR{R}{\V{reg}(r)} \\
-    %
-    \notMapsToR{R}{\V{mem}} & \eqdef & \forall l \in \dom(\V{mem}) \ldotp\spac   \notMapsToR{R}{\V{mem}(l)}
-  \end{array}
-\]
-
-Then, one would use $\notMapsToR{\MMIO \cup [b,e)}$ in place of $\nonCap{}$.
-\end{remark}
 
@@
 
+\appendix
+
+\section{Remarks}
+
+\subsection{Operational Semantics}
+
+\subsubsection{Non-determinism}
+\label{rk:opsem:nondet}
+
+With this presentation, the specification of the ``machine'' is very much
+decoupled from the model of the devices it might be communicating with. This is
+a good thing, I think.
+
+One possible downside is that this makes the operational semantics
+non-deterministic, which might or might not be an issue in the future.
+
+There are several ways in which the operational semantics could be made ``more
+deterministic''.
+
+One could additionally parameterize the operational semantics by an input-stream
+$\stdin$. The updated \textsc{IOLoad} load rule might then look something like
+this:
+\begin{mathpar}
+  \inferrule[IOLoad]
+  { r[\X{src}] = (p,g,b,e,a) \\ \X{readAllowed}\; p \\ a \in [b,e) \\ a \in \MMIO
+  \\ n = \#\{i \mid t[i] = ( \X{IORead},\_\;,\_) \}  }
+  { (r, m, t) \xrightarrow{\I{Load} \; \X{dst} \; \X{src}}
+    (r[\X{dst} := \stdin[n]], m, t \app (\X{IORead}, a, \stdin[n])) }
+\end{mathpar}
+
+When parameterizing our development with multiple values, we could either write
+them out as different Coq \texttt{Parameters} where we need them, or store them
+as fields in a \texttt{DriverG} typeclass (the latter is the approach taken in
+e.g.\ the code-base of~\cite{sammler-2020}).
+
+One could also take as a parameter a predicate that restrict the set of traces
+that are allowed (that is what~\cite{sammler-2020} do).
+
+We could also express more precisely the devices' model, and equip each device
+with some internal state, the ability to react on reads or writes, or to perform
+an internal step. Then, the operational semantics would either step the usual
+way, or whenever a device steps.
+
+I believe this would be somewhat similar to the semantics of I/O system calls
+through a foreign function interfaces as formalized in
+CakeML~\cite{cakeml-vstte17io}, and also in Perennial~\cite{perennial-lang}.
+
+\subsubsection{PC in MMIO}
+\label{rk:opsem:pc_in_mmio}
+
+Additionally, we have the option of disallowing an execution with a \X{PC}
+register pointing to a $\MMIO$ location in the semantics.
+%
+First off, consider that this is an edge case; since we always have points-to
+chunks describing the layout of code in memory (and a points-to chunk for
+location $l$ doubles as a proof that $l \not\in \MMIO$), it is trivial to prove
+that this case would not occur in any of our current examples.
+%
+If we make no changes to the current code base, this behavior would be allowed,
+and the following would happen in both cases described above:
+%
+\begin{enumerate}
+  \item An unconstrained value is read from the location. Any instruction could
+    be executed.
+%
+  \item The predicate $\I{MemLocate}$ in $\I{lang.v}$ will make sure that the
+    default value (e.g. $0$) is read and decoded into an instruction when we try
+    to execute an \MMIO location. This would not cause a conflict with the
+    current WP rules, since a points-to chunk is required in the precondition,
+    ensuring that this case does not occur.
+    %
+    Another alternative is to read the encoding of a $\X{NOP}$ instruction,
+    instead of just the value $0$.
+\end{enumerate}
+%
+If we were to simply disallow execution when the PC points into \MMIO, we would
+make the semantics of all instructions slightly more complicated, but in a
+uniform way.
+%
+We could simply add $a \not\in \MMIO$ to the predicate $\I{isCorrectPC}$ (or add
+it as an additional predicate), which is used to define failing execution. It
+would again not be a drastic change, since a points-to for the address $a$ the
+PC points to is required by the WP rules regardless, from which we can prove
+that $a \not\in \MMIO$.
+
+\subsection{Resources}
+
+\subsubsection{MMIO overlapping with $m$}
+\label{rk:resources:mmio_in_m}
+
+For the case where the \MMIO locations are still part of the heap $m$, this
+would look as follows:
+\[
+  \X{state\_interp} \; (r,m,t) :=
+    \X{gen\_heap\_ctx}\; r \ast
+    \X{gen\_heap\_ctx}\; (m \setminus \MMIO) \ast
+    {\color{BrickRed}\tracefull{t}}
+\]
+
+Sanity check: we never have allocation of new locations $l$ in our capability
+machine, but if we did, that would work in both settings by requiring that
+$l \not\in \MMIO$ in the operational semantics rule for \textsc{Alloc}.
+
+\subsubsection{Fine-grained trace resources}
+\label{rk:resources:fine_grained_trace}
+
+This is very coarse-grained: either one has the full
+ownership for performing I/O and reasoning about it, or one cannot know anything
+about it.
+
+A first extension could be to allow observing prefixes of the trace (since
+events can only be appended to the trace). The observation that the trace
+has a given prefix would be duplicable. This again seems to be an application of
+monotonicity, that could be realized by having a duplicable AtLeast part as part
+of $\tracefrag{t'}$,
+similarly to how we currently model Monotone References.
+
+Another extension, that seems very useful for modular reasoning, would be to
+allow splitting the trace along separate range of addresses. Concretely, a trace
+containing events about the range of MMIO addresses $[a,c)$ could be split into
+two traces, granting ownership over events on addresses $[a,b)$ and $[b,c)$
+respectively. Note that recombining these two traces would only yield some
+unspecified interleaving of the events from the two traces, and not necessarily
+yield the original trace, since in our model, we cannot know the exact
+interleaving of IO operations issued by different parties.
+
+One application of this second extension could be the verification of an example
+involving ``multiplexed'' I/O, where two separate parts of the code are granted
+separate ownership over separate MMIO addresses. These two separate pieces of
+code would be verified separately; then, in the end, one could prove that one
+gets \emph{some} interleaving of all the events emitted by both components.
+
+\subsubsection{MMIO in the heap}
+\label{rk:resources:mmio_in_heap}
+
+As an alternative to having the set of MMIO addresses (\MMIO) as a global
+parameter, we could make it part of the state in the operational semantics. In
+that case, we would need another resource algebra to model the MMIO locations,
+i.e. add do the state interpretation we had before:
+%
+\[
+  \X{state\_interp} \; (r,m,t,\MMIO) := \ldots \ast \MMIOag
+\]
+
+(where $\authm(\exm(\X{Trace}))$ is the resource algebra).
+
+\subsubsection{Dynamic MMIO}
+\label{rk:resources:dynamic_mmio}
+
+In the future, we might be interested in the dynamic allocation of MMIO memory.
+The question is what this would mean, though, since the set of available devices
+allowing for MMIO access and the concrete buffers they provide will most likely
+still be modeled as fixed at runtime.
+Rather, this would allow us to take (un)mapped MMIO locations, connected to the different devices, and (un)map them anywhere in main memory.
+This would require a different way of modeling MMIO, where we need both a notion
+of the total pool of possible MMIO locations, and a description of the currently
+mapped locations. We could model this using an authoritative RA.
+
+\subsection{Toplevel theorem}
+
+\subsubsection{$\VR^\Psi$ with arbitrary $\Psi$}
+\label{rk:toplevel:v_psi}
+
+As it is defined above, the ``generalized'' value relation
+is in fact not very useful for instantiation of $\Psi$ other than $\notMMIO$.
+Indeed, even if $\Psi$ \emph{does} allow referring to addresses in the $\MMIO$
+region, the value relation does not grant any corresponding resources. The fix
+would be to use the generalized trace resources mentioned previously, and grant
+ownership for the part of the trace corresponding to the addresses in
+$[b,e) \cap \MMIO_{\X{pub}}$ with $\MMIO_{\X{pub}}$ the shareable subset of
+addresses in \MMIO.
+
+\subsubsection{Driver with single entrypoint}
+\label{rk:toplevel:single_entrypoint}
+
+We could simplify the theorem statement by only considering the case of a
+single entrypoint. This is equivalent: a piece of code that wants to expose
+two distinct ``methods'' can implement a single toplevel entrypoint that
+immediately returns two pointers that can then be used to invoke the two
+methods.
+%
+This would make the toplevel theorem simpler, at the cost of somewhat more
+contrieved calling convention between trusted and untrusted code.
+Another alternative is to provide an opcode to specify the operation that
+needs to be performed.
+This is slightly less general, since it does not allow providing read access
+only to an adversary.
+
+\subsubsection{OCPL and explicit worlds}
+\label{rk:toplevel:ocpl_explicit_worlds}
+
+The universal quantification on worlds was not explicitly present in the OCPL
+formalization.
+The reason for that is that (I think so at least, do not quote me
+on this - Thomas) they
+used the built-in Iris worlds within their definition of weakest precondition,
+by leveraging the state interpretation, which is universally quantified over in
+the definition of weakest precondition.
+As we discussed before, if we do away
+with local capabilities, it should be possible to fall back onto Iris' notion of
+worlds, which would allow us to remove the universal quantification in the
+above definition too because it would be implicitly present in the WP inside $\ER$.
+
+It would be interesting to have a discussion about how to do away with the
+worlds if we remove local capabilities, regardless of whether we will in the
+end, since that might shed some light on why the worlds look the way they do
+right now.
+
+\subsubsection{Less restricted but still syntactic check for the adversary memory}
+\label{rk:toplevel:less_restricted_syntactic_check_adverary}
+
+If useful, the syntactic check can be relaxed to allow any capabilities that
+do not point into the MMIO region or the driver's memory. One would define the
+predicate $\notMapsToR{R}{w}$, that takes a word, memory region or register
+file and checks if it points into the forbidden region $R$ as follows:
+
+\[
+  \begin{array}{lcl}
+    \notMapsToR{R}{\X{inl}\spac z} & \eqdef
+    & \TRUE \\
+    %
+    \notMapsToR{R}{\X{inr}\spac (p,g,b,e,a)} & \eqdef & R \cap [b,e)\; =
+                                                 \emptyset  \\
+    %
+    \notMapsToR{R}{\V{reg}} & \eqdef & \forall r \in \dom(\V{reg}) \ldotp\spac \notMapsToR{R}{\V{reg}(r)} \\
+    %
+    \notMapsToR{R}{\V{mem}} & \eqdef & \forall l \in \dom(\V{mem}) \ldotp\spac   \notMapsToR{R}{\V{mem}(l)}
+  \end{array}
+\]
+
+Then, one would use $\notMapsToR{\MMIO \cup [b,e)}$ in place of $\nonCap{}$.
+
+%% Biblio
+
 \bibliographystyle{alpha}
@@
 
-\appendix
-
 \end{document}

### 2020-04-03 capability-notes/driver-notes/main.tex
tweak

@@
 
-The capability $(\X{E}, g, b, e, a_i)$ then represents a closure for this same piece
-of code. Adversarial code only gets access to the closure (as an enter
-capability), in order to avoid them from executing or reading arbitrary lines of
-code within the trusted module.
-
-Notice the universal quantification over worlds in the precondition;
-this needs to be there, since we do not know in what world our code will be
-called.
+The capability $(\X{E}, g, b, e, a_i)$ then represents a closure for this same
+piece of code. Adversarial code should only get access to the closure (as an
+enter capability), in order to prevent it from executing or reading arbitrary
+lines of code within the trusted module.
+
+Notice that we also need to provide a world to instantiate the expression
+relation, here, we universally quantify over it...
 %
@@
 
-\medskip
-
-The theorem above is \emph{incomplete}. In other words, it is not provable as
+\bigskip
+
+The theorem above is \emph{incomplete}. In particular, it is not provable as
 is, because it includes no notion of adversarial context (similar to OCPL's

### 2020-04-03 capability-notes/driver-notes/main.tex

@@
         \bigast{m \in [b_{adv}, e_{adv})}{m \mapsto_{a} \X{inl}\; \_} \ast
-        %\bigast{m \in [b, e)}{m \mapsto_{a} \_} \ast
-        \driverC(b,e,\overline{\V{w}_l})\ast
+        \bigast{l \in [b, e)}{m \mapsto_{a}\V{w}_l} \ast
+        % \driverC(b,e,\overline{\V{w}_l})\ast
         \; \tracefrag{t} \ast \pure{P(t)})
@@
     \begin{array}{l}
-       (\exists b_{\X{adv}}\; e_{\X{adv}}\; t.\; \\[0.5em]
+       (\exists b_{\X{adv}}\; e_{\X{adv}}.\; \\[0.5em]
       \quad
@@
         \driverC(b,e,\overline{\V{w}_l})\ast
-        \; \tracefrag{t} \ast  \pure{P(t)}\ast \\[0.5em]
+        \; \trInv \; \ast \\[0.5em]
         \X{region}(\Winit) \ast  \X{sts\_full\_world}\; \X{sts\_std} \;
@@
 
+\subsection{Adapting WP and the FTLR}
+The $\I{Load}$ and $\I{Store}$ cases will require new WP rules, that do not mention
+points-to chunks but mention the trace resource instead (and specifically, we
+will need rules in the presence of invariants such as $\trInv$ above ), when interactions at
+\MMIO locations happen. The general WP spec will need to distinguish between
+the MMIO and non-MMIO cases (alternatively, if both cases are too distinct, the
+ftlr can make this distinction again).
+
+The relations $\VR,\ER,\RR$ need to be updated in terms of $\notMMIO$ and the FTLR will need to be re-stated in terms of these updated relations. This should again only
+require some changes for the $\I{Load}$ and $\I{Store}$ cases, depending on how
+we choose to formalize the WP cases above.
+
+\subsection{Proving the driver safe}
+Specifically, we need to prove the condition
+
+\[
+\forall a_{i} \in A_{\X{entry}} \ldotp\,\left\{
+    {\begin{array}{ll}
+    \trInv \ast
+    \driverC(b,e,\overline{\V{w}_l})\!
+         \vdash \\
+         \forall W \Wpriv \Winit \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
+        \end{array}} \right.
+\]
+
+mentioned above, for the set $A_{\X{entry}}$ of entry points that the boot code
+will actually hand closures out for. This will require reductive reasoning over the concrete code of
+the driver, but should not be too hard in the absence of local return pointers
+and local driver state.
+
+\subsection{Syntactic boot spec}
+
+Proving the syntactic boot specification given above, again through reasoning on
+the concrete boot code that we have written down in our example. This should be
+relatively easy, as our boot code erases adversary memory before handing over
+control. The proof includes verifying any IO that the boot code might need to
+do, using the new WP rules mentioned above.
+
+\subsection{Semantic boot spec}
+
+The semantic boot specification should be derivable from the syntactic one,
+combined with the proof of safety of the driver (so that we know that enter
+capabilities to the driver are safe). The different elements of the semantic bootSpec continuation can be proven as follows:
+\begin{itemize}
+\item First off, we can prove \[\exists \overline{\gamma}\ldotp \X{region}(\emptyset) \ast  \X{sts\_full\_world}\;
+  \X{sts\_std} \; \emptyset\] for \emph{some} ghost names $\overline{\gamma}$ (we
+  left this existential out for simplicity in our spec). This will
+  suffice, since all our other proofs are parametric in the concrete ghost names, and we
+  can use these specific ghost names to construct an instance of $\X{STSG}$. We
+  can then populate the world with a standard, permanent region for the
+  adversary's memory, using the chunks $\bigast{m \in [b_{adv}, e_{adv})}{m
+    \mapsto_{a} \X{inl}\; \_}$ from the syntactic boot spec. This allows us to
+  prove \[\exists \overline{\gamma}\ldotp \X{region}(\Winit) \ast  \X{sts\_full\_world}\;
+    \X{sts\_std} \; \Winit\] in the semantic bootspec.
+  We could generalize the semantic bootspec slightly, by only requiring $[b_{adv},
+  e_{adv})$ to be a subset of adversary memory, but it is not immediately clear
+  what we gain from that.
+\item $\trInv$ and $\driverC(b,e,\overline{\V{w}_l})$ are trivially proven by
+  allocating invariants for $\tracefrag{t} \ast \pure{P(t)}$ and $\bigast{l \in [b, e)}{m \mapsto_{a}\V{w}_l}$ respectively.
+\item In proving that
+  \[\bigast{r \in Reg\setminus \{\X{PC}\}}{(r \mapsto_{r} w_r \ast
+      \VR^{\notMMIO} (\Winit)(w_r))}\]
+  holds, we can distinguish two different cases:
+  \begin{itemize}
+  \item For registers $a_i$ containing closures to the driver, we use the chunks
+    $\bigast{a_{i} \in A_{\X{entry}}}{r_{a_{i}} \mapsto_{r} (\X{E},g,b,e,a_i)}$
+    from the syntactic bootspec, together with the proof of driver safety. To be
+    able to apply the latter, we have to use the $\driverC(b,e,\overline{\V{w}_l})$ and $\trInv$ invariants we allocated earlier.
+  \item All other registers will contain integers $\X{inl}\; z$ and will hence
+    trivially be in $\VR$. We can just use the chunks $\bigast{r \in
+      Reg\setminus \{\X{PC}, A_{\X{entry}}\}}{r \mapsto_{r} \X{inl}\; \_} $ to
+    satisfy the proof obligation.
+  \end{itemize}
+\item The proof of
+  \[\X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, b_{\X{adv}})
+        \ast \VR^{\notMMIO} (\Winit)({\color{red}\X{RW}}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, a_{\X{adv}}))
+  \]
+   follows directly from $ \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{adv}, e_{adv},
+   b_{adv}) $ in the syntactic boot spec, combined with the shape of $\Winit$
+   (i.e. the PC points to a standard, permanent region).
+\item The execution condition for the PC with respect to $\Winit$ can be proven
+  using the fundamental theorem, since we know from the previous bullet that
+  both the read- and write-conditions must be satisfied for the PC register.
+  This completes the proof of $\VR^{\notMMIO} (\Winit)({\color{red}\X{RWX}}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, a_{\X{adv}}))
+$.
+\end{itemize}
+
+\subsection{Proving the robust safety theorem}
+Given a proof of the semantic boot spec, it should be fairly simple to
+prove that the robust safety theorem holds.
+
+To prove that our entire execution is safe, we will first of all need to prove
+that $\X{bootCont_{sem}}(P, g, b, e, A_{\X{entry}},\Phi,\Winit)$ holds. This
+should be simple, since part of the precondition is that the PC is in the
+execution relation, which is essentially what we need to prove (under sensible
+-i.e.\ safe-
+memory and registers).
+
+When we know that our entire execution is safe according to the WP predicate, we
+can apply adequacy of the weakest precondition (and the fact that we have an
+invariant stating that our predicate $P$ must hold), to prove $P(t)$ over
+the entire I/O trace $t$.
+
 \appendix

### 2020-04-14 capability-notes/driver-notes/main.tex
driver notes: bootCont_sem: get rid of the InitW predicate
cosmetic tweak

@@
 \begin{align*}
-    \X{state\_interp} \; (r,m,t) :=& \X{gen\_heap\_ctx}\; r \ast\\
+    \X{state\_interp} \; (r,m,t) := {} & \X{gen\_heap\_ctx}\; r \ast {}\\
     &\X{gen\_heap\_ctx}\; m \ast \dom(m) \cap \MMIO = \emptyset \ast
@@
 
- \X{InitW}(\Winit,b,e) \eqdef\\
-\quad \text{Predicate that forces  $\Winit$ to be the world with one
-    permanent, standard region at $[b,e[$}\\[0.8em]
-
  \X{bootCont_{sem}}(P, g, b, e, A_{\X{entry}},\Phi,\Winit) \eqdef  \\[0.2em]
     \begin{array}{l}
-       (\exists b_{\X{adv}}\; e_{\X{adv}}.\; \\[0.5em]
+       (\exists b_{\X{adv}}\; e_{\X{adv}}\; a_{\X{adv}}\; w_r.\; \\[0.5em]
       \quad
       \begin{array}{l}
-        \X{InitW}(\Winit,b_{\X{adv}}, e_{\X{adv}})\ast \\[0.5em]
-        \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{\X{PC}}, e_{\X{PC}}, a_{\X{PC}})
-        \ast \VR^{\notMMIO} (\Winit)(\X{RW}, \X{G}, b_{\X{PC}}, e_{\X{PC}}, a_{\X{PC}})) \ast {} \\[1.2em]
+        \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, a_{\X{adv}})
+        \ast \VR^{\notMMIO} (\Winit)(\X{RW}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, a_{\X{adv}})) \ast {} \\[1.2em]
         \bigast{r \in Reg\setminus \{\X{PC}\}}{(r \mapsto_{r} w_r \ast

### 2020-04-16 capability-notes/driver-notes/main.tex
Few small tweaks + auctex master file

@@
 is also possible. It would contain the following elements, that should all
-follow directly from the specification above (TODO: write this out properly and
-do the same for the semantic bootspec):
+follow directly from the specification above:
 \begin{itemize}
@@
 the driver address space (TODO: we should change this though and add a
-requirement to $\initOKo$).
+requirement to $\initOKo$). (TODO 2: initOK\_sem is simply the precondition of bootcont\_sem below, we should inline it here or delete it)
 
@@
 
-  \X{bootSpec}(P, g, b, e, A_{\X{entry}},\V{bootCont}) \eqdef \\[0.8em]
+  \X{bootSpec}(P, g, b, e,\V{bootCont}) \eqdef \\[0.8em]
     \quad
@@
 
- \X{bootCont_{sem}}(P, g, b, e, A_{\X{entry}},\Phi,\Winit) \eqdef  \\[0.2em]
+ \X{bootCont_{sem}}(P, g, b, e, \Phi,\Winit) \eqdef  \\[0.2em]
     \begin{array}{l}
-       (\exists b_{\X{adv}}\; e_{\X{adv}}\; a_{\X{adv}}\; w_r.\; \\[0.5em]
+       (\exists b_{\X{adv}}\; e_{\X{adv}}\; a_{\X{adv}}\;\overline{w_r}.\; \\[0.5em]
       \quad
@@
 
-  \X{bootSpec_{sem}}(P, g, b, e, A_{\X{entry}},\Winit)
+  \X{bootSpec_{sem}}(P, g, b, e, \Winit)
     \eqdef\\
-     \quad \forall \Phi\ldotp  \X{bootSpec}(P, g, b, e, A_{\X{entry}},\X{bootCont_{sem}}(P, g, b, e, A_{\X{entry}},\Phi,\Winit))\\[0.8em]
+     \quad \forall \Phi\ldotp  \X{bootSpec}(P, g, b, e,\X{bootCont_{sem}}(P, g, b, e, \Phi,\Winit))\\[0.8em]
 
@@
          \forall W \Wpriv \Winit \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
-        \end{array}} \right. \\   \X{bootSpec}^{\X{sem}}(P,g,b,e,A_{\X{entry}}, \Winit) \\
+        \end{array}} \right. \\   \X{bootSpec}^{\X{sem}}(P,g,b,e, \Winit) \\
   (r_0, m_0, \emptyset) \longrightarrow^* (r, m, t)\\
@@
 
-One could also take as a parameter a predicate that restrict the set of traces
+One could also take as a parameter a predicate that restricts the set of traces
 that are allowed (that is what~\cite{sammler-2020} do).
@@
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2020-04-17 capability-notes/driver-notes/main.tex
First steps towards more modular bootspec

@@
 
-  \X{bootSpec}(P, g, b, e,\V{bootCont}) \eqdef \\[0.8em]
+  \X{bootSpec}(P,r_0,m_0,\V{bootCont},\Phi) \eqdef \\[0.8em]
     \quad
@@
       \ast \bigast{m \in \X{Mem} \setminus \MMIO}{m \mapsto_{a} m_{0}[m]}
-      \ast \; \tracefrag{\emptyset} \ast \\[1.2em]
+      \ast \; \trInv \; \ast \\[1.2em]
+      \later  \V{bootCont}
+    \end{array}
+    \right\}
+    \\[2.5em]
+    \quad \I{Seq}\; (\I{Instr}\; \I{Executable})
+    \\[0.2em]
+    \left\{ \Phi \right\}
+  \end{array}\\[4em]
+
+ \X{bootCont_{sem}}(P,\Phi,\Winit) \eqdef \forall b_{\X{adv}}\; e_{\X{adv}}\; a_{\X{adv}}\;\overline{w_r}\; (W' \Wpriv \Winit) \ldotp\;  \\[0.2em]
+    \begin{array}{l}
+      \begin{array}{l}
+        (\X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, a_{\X{adv}})
+        \ast \VR^{\notMMIO} (W')(\X{RW}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, a_{\X{adv}}) \ast {} \\[.6em]
+        \bigast{r \in Reg\setminus \{\X{PC}\}}{(r \mapsto_{r} w_r \ast
+        \VR^{\notMMIO} (W')(w_r))} \ast {} \\[1.2em]
+        \X{region}(W') \ast  \X{sts\_full\_world}\; \X{sts\_std} \;
+        W')\\[0.5em]       \end{array}
+    \end{array} \\[2em]
+    \quad\wand
+    \begin{array}{l}
+     \wpre{\I{Seq}\; (\I{Instr}\; \I{Executable})}{\Phi}
+    \end{array}\\[1.2em]
+
+  \X{bootSpec_{sem}}(P,r_0,m_0)
+    \eqdef\\
+     \quad \forall \Phi\ldotp  \X{bootSpec}(P,r_0,m_0,\X{bootCont_{sem}}(P,\Phi,\emptyset),\Phi)\\[0.8em]
+
+  \end{array}
+\]
+
+To prove that any closures the boot code might create in the above actually constitute valid register values, we would need to prove the following, with $A_{\X{entry}}$ the set of entry points to the driver:
+\[
+ \forall a_{i} \in A_{\X{entry}} \ldotp\,           \forall W \Wpriv \Winit \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
+\]
+
+To actually be able to construct this proof, we will need the following 2 elements (they both need to be persistent, so that we can use them within the execution condition):
+
+\[
+    \trInv \ast
+    \driverC(b,e,\overline{\V{w}_l})
+\]
+
+Note that the above specification does not allow for a scenario where we have multiple set-up programs in a row, i.e.\ multiple boot or set-up processes that each need some form of access to \MMIO, that each want to set up their own closures and that we want to verify modularly.
+The trust model of these set-up procedures goes bottom-up, i.e.
+any set-up procedure only trusts the set-up procedures that ran before it to ensure that its preconditions are satisfied when it assumes control.
+To allow for such a scenario, the execution of the bootcode has to be split up into different phases, where it holds that the postondition (precondition of the continuation) of boot program $i$ implies the precondition of boot program $i+1$.
+Additionally, the last bootloader has to make sure that the registers it delivers are safe with respect to $\VR^{\notMMIO}$, such that the ftlr (which will only hold for adversaries that have no access to \MMIO whatsoever) can be applied.
+
+In general, each set-up process $i$ will require the following as a precondition to safely execute.
+\begin{itemize}
+\item Safe register values, with respect to some argument World $W$ (whose regions are standard - this will mostly be used for proving the enter capabilities to e.g\ the driver secure, and I expect this world to be empty at the moment) and a set of \MMIO locations $\MMIO_{i}$, for which it holds that $\MMIO_i \subseteq \MMIO$.
+This allows the process to perform some $\MMIO$, but only to locations in $\MMIO \setminus \MMIO_i$.
+If needed, more strict functional requirements on registers can be made.
+\item Its code and data stored somewhere accessible in memory (using regular points-to chunks).
+Generally, also point-to chunks for memory that will be used in the future.
+This is the region, $R_{\X{adv}}$, currently represented as a contiguous range $[b_{\X{adv}},e_{\X{adv}}[$ for simplicity.
+Note that it is possible to write drivers that use relative addressing in their spec to not make their precondition unnecessarily strict.
+\item The $PC$ register points to (a subset of)  $R_{\X{adv}}$, so that it is actually possible for our process to access this non-standard memory.
+Optionally, part of the memory the $PC$ points to can be in $W$, but then the precondition needs to contain a proof that this part of the $PC$ is $RW$-valid, i.e.
+$\in \VR(\X{RW},\ldots)$, so that we can actually obtain the necessary points-to chunks from the region predicate.
+\item An invariant $\trInv$ on the trace, so that the process can perform IO itself or create more closures to do so.
+\end{itemize}
+
+Each set-up process will deliver the following postcondition (precondition of the continuation) to the next set-up process or to adversary code, starting from the precondition: (this also means that the proof of the next process has to go through, given this postcondition, i.e.
+it has to be sufficiently specific)
+\begin{itemize}
+\item Safe register values in any ($\forall$) private future world of $W$, with respect to the relation $\VR^{\MMIO_{i+1}}$, where it has to hold that $\MMIO_{i} \subseteq \MMIO_{i+1}$ i.e.
+the permissions to perform IO are monotonely decreasing as set-up progresses.
+This constitutes no problem for instances of $\VR$,$\ER$ and $\RR$ proven in earlier set-up programs, since they all respect the following contravariant monotonicity property: $S_1 \subseteq S_2 \Rightarrow \X{\mathfrak{X}}^{\overline{S_2}} \subseteq \X{\mathfrak{X}}^{\overline{S_1}}$.
+\item Some description of the layout of memory and concrete register values, that should be sufficiently detailed to be able to prove the precondition of the next set-up program.
+\item The trace invariant $\trInv$ (so that the next process can perform its own $\MMIO$).
+In the future, we can allow strengthening of $P$ throughout set-up by creating a monotone resource algebra for predicates, similarly to what we did for OCPL.
+\end{itemize}
+
+In the case of the last set-up process $n$, we will also need the following, to be able to prove the semantic bootspec's postondition and in the end apply the ftlr, thereby proving the rest of execution safe:
+\begin{itemize}
+\item We need that $\MMIO_n = \MMIO$, such that we can apply the ftlr (otherwise some valid registers might grant the adverary access to \MMIO memory, and we cannot consider it an adversarial context in the robust safety sense)
+\item For similar reasons, the $PC$ should now point to valid memory only, i.e. the condition $\VR^{\notMMIO} (W)(\X{RW}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, a_{\X{adv}})$ should hold.
+\item The trace invariant $\trInv$ is no longer required, since the untrusted adversary code will not be able to perform \MMIO anyway.
+Leaving it there is not a technical problem either, however, given that the registers and memory will be valid with respect to $\VR^{\MMIO}$.
+\end{itemize}
+
+Using the notation we introduced, we must prove the following for each set-up program (TODO: write an attempt at a more modular spec here):
+\[
+\begin{array}{l}
+
+  \X{bootSpec}(P, g, b, e,\V{bootCont},\Phi) \eqdef \\[0.8em]
+    \quad
+  \begin{array}{l}
+    \left\{
+    \begin{array}{l}
+      \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, 0, \X{MEM_{MAX}}, l_{\X{boot}}) \ast {} \\[0.5em]
+      \bigast{r \in \X{Reg} \setminus \X{PC}}{r \mapsto_{r} r_{0}[r]}
+      \ast \bigast{m \in \X{Mem} \setminus \MMIO}{m \mapsto_{a} m_{0}[m]}
+      \ast \; \trInv \; \ast \\[1.2em]
       \later  \V{bootCont}
@@
         % \driverC(b,e,\overline{\V{w}_l})\ast
-        \; \tracefrag{t} \ast \pure{P(t)})
+        \; \tracefrag{t} \ast \pure{P(t)}\\
+        \forall a_{i} \in A_{\X{entry}} \ldotp\,\left\{
+    {\begin{array}{ll}
+    \trInv \ast
+    \driverC(b,e,\overline{\V{w}_l})\!
+         \vdash \\
+         \forall W \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
+        \end{array}} \right. )
       \end{array}
@@
     \end{array}\\[1.2em]
-
-  \X{bootSpec_{synt}}(P, g, b, e, A_{\X{entry}}) \eqdef\\
-     \quad \forall \Phi\ldotp  \X{bootSpec}(P, g, b, e, A_{\X{entry}},\X{bootCont_{synt}}(P, g, b, e, A_{\X{entry}},\Phi))\\[0.8em]
-
- \X{bootCont_{sem}}(P, g, b, e, \Phi,\Winit) \eqdef  \\[0.2em]
-    \begin{array}{l}
-       (\exists b_{\X{adv}}\; e_{\X{adv}}\; a_{\X{adv}}\;\overline{w_r}.\; \\[0.5em]
-      \quad
-      \begin{array}{l}
-        \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, a_{\X{adv}})
-        \ast \VR^{\notMMIO} (\Winit)(\X{RW}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, a_{\X{adv}})) \ast {} \\[1.2em]
-        \bigast{r \in Reg\setminus \{\X{PC}\}}{(r \mapsto_{r} w_r \ast
-        \VR^{\notMMIO} (\Winit)(w_r))} \ast {} \\[1.2em]
-        \driverC(b,e,\overline{\V{w}_l})\ast
-        \; \trInv \; \ast \\[0.5em]
-        \X{region}(\Winit) \ast  \X{sts\_full\_world}\; \X{sts\_std} \;
-        \Winit)\\[0.5em]       \end{array}
-    \end{array} \\[4em]
-    \quad\wand
-    \begin{array}{l}
-     \wpre{\I{Seq}\; (\I{Instr}\; \I{Executable})}{\Phi}
-    \end{array}\\[1.2em]
-
-  \X{bootSpec_{sem}}(P, g, b, e, \Winit)
-    \eqdef\\
-     \quad \forall \Phi\ldotp  \X{bootSpec}(P, g, b, e,\X{bootCont_{sem}}(P, g, b, e, \Phi,\Winit))\\[0.8em]
-
 \end{array}
@@
 
-The resulting boot-code formulation of the robust safety theorem is now a
+The resulting boot-code formulation of the robust safety theorem is now
 
@@
   \inferrule
-  {  \forall a_{i} \in A_{\X{entry}} \ldotp\,\left\{
-    {\begin{array}{ll}
-    \trInv \ast
-    \driverC(b,e,\overline{\V{w}_l})\!
-         \vdash \\
-         \forall W \Wpriv \Winit \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
-        \end{array}} \right. \\   \X{bootSpec}^{\X{sem}}(P,g,b,e, \Winit) \\
+  { \X{bootSpec}^{\X{sem}}(P,r_0,m_0) \\
   (r_0, m_0, \emptyset) \longrightarrow^* (r, m, t)\\
@@
    \end{array}
+\]
+
+Actually, there are two options, where the second one is probably slightly more modular:
+\[
+  \begin{array}{lcl}
+
+   \trInvNp[\pure{|t| \le 1000} \ast
+  \exists n\ldotp l_n \mapsto n \ast n = |t| ]\\
+   \trInvNp[\pure{|t| \le 1000} \ast
+    \exists n\ldotp \ownGhost{\gamma}{\authfull{n}} \ast n = |t| ] \ast \knowInv{\iota}{\exists n \ldotp \ownGhost{\gamma}{\authfrag{n}} \ast  l_n \mapsto n}
+  \end{array}\\
 \]

### 2020-04-19 capability-notes/driver-notes/main.tex
Driver: threw out some outdated bootspec text

@@
 \newcommand{\app}{\mathbin{+\!\!+}}
+\newcommand{\nonCap}[1]{\ensuremath{\mathrm{nonCap}(#1)}}
 
@@
 
-We move to a theorem statement that correctly constrains the adversary in two
-steps:
-%
-\begin{enumerate}
-  \item We add a syntactic condition (on the state of memory and registers) to
-    the tentative statement of our robust safety theorem, ensuring that the
-    adversary code cannot gain access to any non-$\X{E}$ capabilities pointing into
-    the driver code or any capabilities pointing into $\MMIO$ directly.
-    Alternatively, we can formulate this condition in a semantic way, see below.
-    Additionally, we specify what exact code the driver contains, so that we can
-    use this in our proof that the driver is secure. This will involve adding an
-    invariant to the robust safety theorem.
-%
-  \item Then, we provide an end-to-end statement that assumes the existence of a
-    piece of boot code, that gets to run first when the capability machine
-    starts up, and makes sure the above condition indeed holds before passing
-    control to the adversary.
-
-    The boot code is part of the trusted code base, and must satisfy a given
-    specification, as a precondition of the toplevel theorem.
-
-    The particular implementation of the boot code depends on the model of the
-    initial state of the machine. If the memory can contain any capability at
-    startup, then the boot code has to erase the memory before passing control.
-    If we assume that the initial memory cannot contain any capabilities
-    whatsoever, then the boot code does not need to do any cleanup.
-\end{enumerate}
-
-In this setting, we can subdivide the memory $m$ into 3 relevant subsections;
-$m = m_{\MMIO} \uplus m_{\X{driver}} \uplus m_{\X{adv}}$ %\uplus m_{\X{frame}}
-with the following meaning:
-\begin{itemize}
-\item $m_{\MMIO}$ contains the memory-mapped locations
-\item $m_{\X{driver}}$ contains the driver code
-\item  $m_{\X{adv}}$ contains the adversary's code and data (this will also
-  include the boot code in the general setting, since it does not contain any
-  inherently dangerous capabilities itself).
-\end{itemize}
-
-\newcommand{\nonCap}[1]{\ensuremath{\mathrm{nonCap}(#1)}}
-
-We can then implement the syntactic check as requiring registers and memory to
-only contain non-capability values:
+We now try to develop a theorem statement that correctly constrains the adversary.
+
+A first thing that is lacking is a way to constrain the memory and the registers of the adversary once it gains control.
+We have to add a condition (on the state of the reachable memory and registers) to the tentative statement of our robust safety theorem, ensuring that the adversary code cannot gain access to any non-$\X{E}$ capabilities pointing into the driver code or any capabilities pointing into $\MMIO$ directly.
+A possible way of syntactically expressing this condition (and in the process overestimating it), is to define the predicate $\nonCap$ below.
+More specifically, we can implement the syntactic check as requiring registers and memory (except perhaps driver memory) to only contain non-capability values:
 
@@
 \end{mathpar}
-
-\newcommand{\notMapsToR}[2]{\rightarrow^{\overline{#1}}\!\!(#2)}
-\newcommand{\Winit}{\X{W_{init}}}
-\newcommand{\Wemp}{\emptyset}
-
-Given this definition, we can now formalize the initialization constraints on
+\rk{toplevel:less_restricted_syntactic_check_adverary}
+
+This condition can also easily be specified in terms of the value relation $\VR^{\notMMIO}$, by requiring the registers and part of the PC register to be valid, as will be demonstrated below.
+This more semantic specification is easily implied by the syntactic one above, and is additionally the one we will need to apply the ftlr to the adversary code, hence proving that it executes securely.
+
+Another change we add is that rather than having execution start off with the adversary, we model the entire boot process of a capability machine, and allow so-called \emph{boot code} to run before the adversary does.
+Because of this set-up, rather than having a theorem that assumes unknown adversary code to start executing immediately, we provide an end-to-end theorem that assumes the existence of this piece of boot code, that gets to run first when the capability machine starts up, and makes sure the necessary conditions for the ftlr indeed hold before passing control to the adversary.
+The boot code is part of the trusted code base, and must satisfy a given specification, as a precondition of the toplevel theorem.
+We call this bootcode specification BootSpec and will demonstrate it below.
+The particular implementation of the boot code depends on the model of the initial state of the machine.
+If the memory can contain any capability at startup, then the boot code has to erase the memory before passing control.
+If we assume that the initial memory cannot contain any capabilities whatsoever, then the boot code does not need to do any cleanup.
+
+In this setting, we can subdivide the memory $m$ into 3 relevant subsections;
+$m = m_{\MMIO} \uplus m_{\X{driver}} \uplus m_{\X{adv}}$ %\uplus m_{\X{frame}}
+with the following meaning:
+\begin{itemize}
+\item $m_{\MMIO}$ contains the memory-mapped locations
+\item $m_{\X{driver}}$ contains the driver code from address $b$ to $e$
+\item  $m_{\X{adv}}$ contains the adversary's code and data (this will also
+  include the boot code in the general setting, since it does not contain any
+  inherently dangerous capabilities itself). For simplicity, we will assume a contiguous region for the adversary, from $b_{\X{adv}}$ to $e_{\X{adv}}$.
+\end{itemize}
+
+Given this definition, we could now formalize the initialization constraints on
 registers $r$ and memory $m$, given the region \MMIO and
-the driver's address space $[b,e)$ as follows (for a resource-aware formulation;
-see the bootSpec postcondition below):
+the driver's address space $[b,e)$ on paper as follows:
 
@@
 
-\rk{toplevel:less_restricted_syntactic_check_adverary}
-
-It should be noted that a more general, semantic version of this specification
-is also possible. It would contain the following elements, that should all
-follow directly from the specification above:
+However, this predicate is not defined in terms of Iris resources, i.e.
+we cannot use it in our Coq development yet.
+We will redefine it below, in the process making it more semantic and hence general.
+The above predicate, however, demonstrates the state we would like our capability machine to be in when the adverary starts executing.
+Additionally, we will need to state that any registers pointing to the driver are secure; this is still lacking from the formalization.
+
+\newcommand{\notMapsToR}[2]{\rightarrow^{\overline{#1}}\!\!(#2)}
+\newcommand{\Winit}{\X{W_{init}}}
+\newcommand{\Wemp}{\emptyset}
+
+The general semantic precondition on adversary execution would then contain the following elements:
 \begin{itemize}
 \item All registers except \X{PC} are in the value relation with respect to
-  either the empty world $\Wemp$ (if we do not want to allow any non-empty, non-enter
-  capabilities in registers), or a world $\Winit$ containing a single standard region for
+  a world $W$, meant to contain a single standard region for
   all of the adversary's memory.
   The \X{PC} register satisfies both the read- and write conditions (note the
-  parallel with the fundamental theorem) with respect to this same world, if we
-  picked the non-empty world, or just points to the adversary's memory, which
-  does not contain capabilities (if we
-  picked the empty world, a capability pointing into memory can never be valid).
-  It feels slightly less artificial to pick the non-empty world option here.
+  parallel with the fundamental theorem) with respect to this same world.
+  If we ever want to allow more than just enter capabilities to point to the driver's memory, we have to make sure that some region governing the driver is part of the world $W$.
+It is not entirely clear yet how to do this in our current development.
 \item We have possession of the predicates $\I{region}$ and
@@
   need these to apply the fundamental theorem.
-\item A description of the exact driver code, still in terms of points-to chunks
-  (or, in this case equivalently, an invariant). We can write this invariant as
-      $\driverC(b,e,\overline{\V{w}_l}) = \knowInv{\iota}{\bigast{l \in [b,e)}{l \mapsto_{a} \mathit{\V{w}_l}}}$, with $\V{w}_l$
-      some hard-coded driver word (an instruction or a capability) at location $l$.
 \end{itemize}
 
-and subsequently formulate the robust safety theorem as follows (using the
-semantic specification for $\initOKo$, which references the initial world):
-
-\begin{mathpar}
-  \mprset{vskip=0.5em}
-  \inferrule
-  { \forall a_{i} \in A_{\X{entry}} \ldotp\,\left\{
-    {\begin{array}{ll}
-    \trInv \ast
-    \driverC(b,e,\overline{\V{w}_l})\!
-         \vdash \\
-         \forall W \Wpriv \Winit \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
-        \end{array}} \right. \\
-   \initOKS{[b,e)}[r][m][\Winit] \\
-  (r[\overline{r_{a_{i}} := (\X{E}, g, b, e, a_i)}], m, \emptyset) \longrightarrow^* (r', m', t)\\
-  }
-  {P(t)}
-\end{mathpar}
-
-where we have set the PC up to be a sensible capability (pointing to
-$b_{\X{adv}}$ without loss of generality).
-
-Notice how the syntactic check for the registers is only performed on the set of
-registers $r$, allowing the closures in $\overline{r_{a_i}}$ to point into to
-the driver address space (TODO: we should change this though and add a
-requirement to $\initOKo$). (TODO 2: initOK\_sem is simply the precondition of bootcont\_sem below, we should inline it here or delete it)
-
-Having this theorem in our toolbox, we can now take a closer look at scenario 2
-above, where a piece of boot code is used to set the memory up correctly,
+We can now take a closer look at the scenario where a piece of boot code is used to set the memory up correctly,
 obviating the need for any other assumptions than that the boot code behaves
-according to some spec. Concretely, we want the boot code to uphold a spec that
-allows us to prove the above theorem.
-%
-(Alternatively, we could also just implement one specific instance of boot code,
-and prove that after it finishes executing, all preconditions to apply the above
-theorem are met.)
+according to some spec.
 
@@
 We also assume the capability machine to initially start execution in the boot code  at address $l_{\X{boot}}$.
-%
-Then, the syntactic specification for the boot code that we have to prove is
-stated as follows (we give both a specific, syntactic specification
-\X{bootSpec_{synt}} and a more
-general, semantic one \X{bootSpec_{sem}} that should follow from the syntactic specification):
+
+%
+Then, the semantic specification for the boot code that we have to prove is
+stated as follows:
 
@@
 \]
+Here we defined $\driverC$ to be a description of the exact driver code, still in terms of points-to chunks. We can write this invariant as
+      $\driverC(b,e,\overline{\V{w}_l}) = \knowInv{\iota}{\bigast{l \in [b,e)}{l \mapsto_{a} \mathit{\V{w}_l}}}$, with $\V{w}_l$
+      some hard-coded driver word (an instruction or a capability) at location $l$.
+
+To prove the \X{PC} register valid, we will often be able to use the fact that the memory left by the boot code only contains integers $z$, i.e.\ that the $\mathrm{nonCap}$ predicate holds for memory.
 

### 2020-04-19 capability-notes/driver-notes/main.tex
Threw out some more outdated remarks

@@
   a different ghost name $\gamma$ than we usually do.
-  I do not know how exactly that would work on a technical level, but it should somehow.
+  I do not know how exactly that would work on a technical level, but it should somehow. We should probably have $\X{region}(\emptyset) \ast
+  \X{sts\_full\_world}\; \X{sts\_std} \; \emptyset$ in the precondition of the hoare triple, and then externally have a universal quantification over the concrete $\I{traceG}$ typeclass used, similar to how OCPL does this.
 \end{remark}
@@
 
-Note that to link scenario 2 to scenario 1, we need a slightly adapted version
-of the robust safety theorem stated previously, where the initial trace is
+Note that we will need a slightly more general adequacy theorem than is suggested by the above robust safety theorem, where the initial trace is
 allowed to not be empty, but rather just satisfy $P(t)$, but that should not
@@
 where $P$ is thus only true of the suffix of the trace for events that occur
-\emph{after} the boot code has run...
-
-\begin{remark}
-  On the level of the operational semantics, proving the boot specification
-  corresponds to proving that the following holds, in order to use our concrete
-  driver implementation with the robust safety theorem above (the driver still
-  needs to be proven safe separately):
-\[
-  \begin{array}{l}
-    (\textsc {BootCodeOK}) \\[0.5em]
-%
-    \exists n\; b\; e\ldotp (r_0, m_0, \emptyset) \longrightarrow^n (r'[\overline{r_{a_{i}}
-    := (\X{E}, g, b, e, a_i)}], m', t) \\[0.5em]
-    {} \land P(t) \land \initOK{[b,e)}[r'][m']
-  \end{array}
-\]
-\end{remark}
+\emph{after} the boot code has run.
+
 \begin{remark}
@@
 \end{remark}
+
 \subsubsection{Concrete boot and driver code}

### 2020-04-19 capability-notes/driver-notes/main.tex
Driver: small note on non-atomic invariants

@@
 
+Lastly, note how the trace invariant that appears in the boot specification is not an atomic invariant, even though atomic invariants might be involved in the verification of the driver if it has state (see e.g.\ the counter example below).
+We are obliged to have a regular invariant for the predicate over the trace, since otherwise adequacy breaks; even for infinite executions, we have to be able to show that any output values satisfy $P$ at \emph{any} point during execution.
+
 \begin{remark}
@@
 including what its boot code looks like. Our goal is to prove that this concrete
-example satisfies \X{bootSpec} (alternatively, \textsc{BootCodeOk}) and this
-last robust safety theorem.
+example satisfies \X{bootSpec} and hence the above robust safety theorem.
 
@@
 driver. Concretely, the invariant we will verify our driver under changes from
+
+\newcommandx{\NaInvT}[1][1 = \top]{[\X{NaInv}: \X{driver\_nais}.#1]}
+\newcommandx{\NaInvI}[2][1 = \iota]{\operatorname{NaInv}^{#1}(#2)}
 $\trInv[|t| \le 1000]$ to
-$\trInvNp[\pure{|t| \le 1000} \ast
-  \exists n\ldotp l_n \mapsto n \ast n = |t| ]$, where we defined
+$\trInvNp[\exists n \ldotp  \pure{|t| = n \mathrel{\land} n \le 1000} \mathrel{\ast} \ownGhost{\gamma_n}{\authfull n}] \mathrel{\ast} \NaInvI[\iota]{\exists n\ldotp l_n \mapsto n \ast \ownGhost{\gamma_n}{\authfrag n}]}  $, where we defined
 %
@@
 \]
-
-Actually, there are two options, where the second one is probably slightly more modular:
-\[
-  \begin{array}{lcl}
-
-   \trInvNp[\pure{|t| \le 1000} \ast
-  \exists n\ldotp l_n \mapsto n \ast n = |t| ]\\
-   \trInvNp[\pure{|t| \le 1000} \ast
-    \exists n\ldotp \ownGhost{\gamma}{\authfull{n}} \ast n = |t| ] \ast \knowInv{\iota}{\exists n \ldotp \ownGhost{\gamma}{\authfrag{n}} \ast  l_n \mapsto n}
-  \end{array}\\
-\]
-
 (i.e., the predicate over the trace does not need to be pure any longer).
@@
 other invariants relating to local state over the trace as well.
+
+Note that the following two formalizations do not work, since the trace and counter value are not updated at the same time in the code:
+\[
+  \begin{array}{lcl}
+
+   \trInvNp[\pure{|t| \le 1000} \ast
+  \exists n\ldotp l_n \mapsto n \ast n = |t| ]\\
+   \trInvNp[\pure{|t| \le 1000} \ast
+    \exists n\ldotp \ownGhost{\gamma}{\authfull{n}} \ast n = |t| ] \ast \knowInv{\iota}{\exists n \ldotp \ownGhost{\gamma}{\authfrag{n}} \ast  l_n \mapsto n}
+  \end{array}\\
+\]
+
+Using  a non-atomic invariant in the first of these two formalizations would not work either, as it would break the adequacy (as stated earlier); we would not know that the safety property inside $P_{\X{np}}$ was upheld at $any$ point during execution.
 

### 2020-04-20 capability-notes/driver-notes/main.tex
Driver: first attempt at modular driver spec

@@
 \newcommand{\pure}[1]{\tensor[^{\ulcorner}]{#1{}}{^{\urcorner}}} %Hacky {} to avoid double superscript
+\newcommand{\br}[1]{{\color{BrickRed} #1}}
 
@@
 
-  \X{bootSpec}(P,r_0,m_0,\V{bootCont},\Phi) \eqdef \\[0.8em]
-    \quad
-  \begin{array}{l}
-    \left\{
-    \begin{array}{l}
-      \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, 0, \X{MEM_{MAX}}, l_{\X{boot}}) \ast {} \\[0.5em]
-      \bigast{r \in \X{Reg} \setminus \X{PC}}{r \mapsto_{r} r_{0}[r]}
-      \ast \bigast{m \in \X{Mem} \setminus \MMIO}{m \mapsto_{a} m_{0}[m]}
-      \ast \; \trInv \; \ast \\[1.2em]
-      \later  \V{bootCont}
-    \end{array}
-    \right\}
-    \\[2.5em]
-    \quad \I{Seq}\; (\I{Instr}\; \I{Executable})
-    \\[0.2em]
-    \left\{ \Phi \right\}
-  \end{array}\\[4em]
-
- \X{bootCont_{sem}}(P,\Phi,\Winit) \eqdef \forall b_{\X{adv}}\; e_{\X{adv}}\; a_{\X{adv}}\;\overline{w_r}\; (W' \Wpriv \Winit) \ldotp\;  \\[0.2em]
+ \X{bootCont}(P,\Phi,\Winit) \eqdef \forall b_{\X{adv}}\; e_{\X{adv}}\; a_{\X{adv}}\;\overline{w_r}\; (W' \Wpriv \Winit) \ldotp\;  \\[0.2em]
     \begin{array}{l}
@@
 
-  \X{bootSpec_{sem}}(P,r_0,m_0)
-    \eqdef\\
-     \quad \forall \Phi\ldotp  \X{bootSpec}(P,r_0,m_0,\X{bootCont_{sem}}(P,\Phi,\emptyset),\Phi)\\[0.8em]
-
-  \end{array}
-\]
-
-To prove that any closures the boot code might create in the above actually constitute valid register values, we would need to prove the following, with $A_{\X{entry}}$ the set of entry points to the driver:
-\[
- \forall a_{i} \in A_{\X{entry}} \ldotp\,           \forall W \Wpriv \Winit \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
-\]
-
-To actually be able to construct this proof, we will need the following 2 elements (they both need to be persistent, so that we can use them within the execution condition):
-
-\[
-    \trInv \ast
-    \driverC(b,e,\overline{\V{w}_l})
-\]
-Here we defined $\driverC$ to be a description of the exact driver code, still in terms of points-to chunks. We can write this invariant as
-      $\driverC(b,e,\overline{\V{w}_l}) = \knowInv{\iota}{\bigast{l \in [b,e)}{l \mapsto_{a} \mathit{\V{w}_l}}}$, with $\V{w}_l$
-      some hard-coded driver word (an instruction or a capability) at location $l$.
-
-To prove the \X{PC} register valid, we will often be able to use the fact that the memory left by the boot code only contains integers $z$, i.e.\ that the $\mathrm{nonCap}$ predicate holds for memory.
-
-Note that the above specification does not allow for a scenario where we have multiple set-up programs in a row, i.e.\ multiple boot or set-up processes that each need some form of access to \MMIO, that each want to set up their own closures and that we want to verify modularly.
-The trust model of these set-up procedures goes bottom-up, i.e.
-any set-up procedure only trusts the set-up procedures that ran before it to ensure that its preconditions are satisfied when it assumes control.
-To allow for such a scenario, the execution of the bootcode has to be split up into different phases, where it holds that the postondition (precondition of the continuation) of boot program $i$ implies the precondition of boot program $i+1$.
-Additionally, the last bootloader has to make sure that the registers it delivers are safe with respect to $\VR^{\notMMIO}$, such that the ftlr (which will only hold for adversaries that have no access to \MMIO whatsoever) can be applied.
-
-In general, each set-up process $i$ will require the following as a precondition to safely execute.
-\begin{itemize}
-\item Safe register values, with respect to some argument World $W$ (whose regions are standard - this will mostly be used for proving the enter capabilities to e.g\ the driver secure, and I expect this world to be empty at the moment) and a set of \MMIO locations $\MMIO_{i}$, for which it holds that $\MMIO_i \subseteq \MMIO$.
-This allows the process to perform some $\MMIO$, but only to locations in $\MMIO \setminus \MMIO_i$.
-If needed, more strict functional requirements on registers can be made.
-\item Its code and data stored somewhere accessible in memory (using regular points-to chunks).
-Generally, also point-to chunks for memory that will be used in the future.
-This is the region, $R_{\X{adv}}$, currently represented as a contiguous range $[b_{\X{adv}},e_{\X{adv}}[$ for simplicity.
-Note that it is possible to write drivers that use relative addressing in their spec to not make their precondition unnecessarily strict.
-\item The $PC$ register points to (a subset of)  $R_{\X{adv}}$, so that it is actually possible for our process to access this non-standard memory.
-Optionally, part of the memory the $PC$ points to can be in $W$, but then the precondition needs to contain a proof that this part of the $PC$ is $RW$-valid, i.e.
-$\in \VR(\X{RW},\ldots)$, so that we can actually obtain the necessary points-to chunks from the region predicate.
-\item An invariant $\trInv$ on the trace, so that the process can perform IO itself or create more closures to do so.
-\end{itemize}
-
-Each set-up process will deliver the following postcondition (precondition of the continuation) to the next set-up process or to adversary code, starting from the precondition: (this also means that the proof of the next process has to go through, given this postcondition, i.e.
-it has to be sufficiently specific)
-\begin{itemize}
-\item Safe register values in any ($\forall$) private future world of $W$, with respect to the relation $\VR^{\MMIO_{i+1}}$, where it has to hold that $\MMIO_{i} \subseteq \MMIO_{i+1}$ i.e.
-the permissions to perform IO are monotonely decreasing as set-up progresses.
-This constitutes no problem for instances of $\VR$,$\ER$ and $\RR$ proven in earlier set-up programs, since they all respect the following contravariant monotonicity property: $S_1 \subseteq S_2 \Rightarrow \X{\mathfrak{X}}^{\overline{S_2}} \subseteq \X{\mathfrak{X}}^{\overline{S_1}}$.
-\item Some description of the layout of memory and concrete register values, that should be sufficiently detailed to be able to prove the precondition of the next set-up program.
-\item The trace invariant $\trInv$ (so that the next process can perform its own $\MMIO$).
-In the future, we can allow strengthening of $P$ throughout set-up by creating a monotone resource algebra for predicates, similarly to what we did for OCPL.
-\end{itemize}
-
-In the case of the last set-up process $n$, we will also need the following, to be able to prove the semantic bootspec's postondition and in the end apply the ftlr, thereby proving the rest of execution safe:
-\begin{itemize}
-\item We need that $\MMIO_n = \MMIO$, such that we can apply the ftlr (otherwise some valid registers might grant the adverary access to \MMIO memory, and we cannot consider it an adversarial context in the robust safety sense)
-\item For similar reasons, the $PC$ should now point to valid memory only, i.e. the condition $\VR^{\notMMIO} (W)(\X{RW}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, a_{\X{adv}})$ should hold.
-\item The trace invariant $\trInv$ is no longer required, since the untrusted adversary code will not be able to perform \MMIO anyway.
-Leaving it there is not a technical problem either, however, given that the registers and memory will be valid with respect to $\VR^{\MMIO}$.
-\end{itemize}
-
-Using the notation we introduced, we must prove the following for each set-up program (TODO: write an attempt at a more modular spec here):
-\[
-\begin{array}{l}
-
-  \X{bootSpec}(P, g, b, e,\V{bootCont},\Phi) \eqdef \\[0.8em]
+  \X{bootSpec}(P,r_0,m_0) \eqdef \forall \Phi \ldotp \\[0.8em]
     \quad
@@
       \bigast{r \in \X{Reg} \setminus \X{PC}}{r \mapsto_{r} r_{0}[r]}
-      \ast \bigast{m \in \X{Mem} \setminus \MMIO}{m \mapsto_{a} m_{0}[m]}
-      \ast \; \trInv \; \ast \\[1.2em]
-      \later  \V{bootCont}
+      \ast \bigast{m \in \X{Mem} \setminus \MMIO}{m \mapsto_{a} m_{0}[m]} \ast\\[.5em]
+      \X{region}(\emptyset) \ast  \X{sts\_full\_world}\; \X{sts\_std} \; \emptyset\ast \; \trInv \; \ast \\[.5em]
+      \later \X{bootCont}(P,\Phi,\emptyset)
     \end{array}
@@
 
-    \X{bootCont_{synt}}(P, g, b, e, A_{\X{entry}},\Phi) \eqdef  \\[0.2em]
+  \end{array}
+\]
+
+To prove that any closures the boot code might create in the above actually constitute valid register values, we would need to prove the following, with $A_{\X{entry}}$ the set of entry points to the driver:
+\[
+ \forall a_{i} \in A_{\X{entry}} \ldotp\,           \forall W \Wpriv \Winit \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
+\]
+
+To actually be able to construct this proof, we will need the following 2 elements (they both need to be persistent, so that we can use them within the execution condition):
+
+\[
+    \trInv \ast
+    \driverC(b,e,\overline{\V{w}_l})
+\]
+Here we defined $\driverC$ to be a description of the exact driver code, still in terms of points-to chunks. We can write this invariant as
+      $\driverC(b,e,\overline{\V{w}_l}) = \knowInv{\iota}{\bigast{l \in [b,e)}{l \mapsto_{a} \mathit{\V{w}_l}}}$, with $\V{w}_l$
+      some hard-coded driver word (an instruction or a capability) at location $l$.
+
+To prove the \X{PC} register valid, we will often be able to use the fact that the memory left by the boot code only contains integers $z$, i.e.\ that the $\mathrm{nonCap}$ predicate holds for memory.
+
+Note that the above specification does not allow for a scenario where we have multiple set-up programs in a row, i.e.\ multiple boot or set-up processes that each need some form of access to \MMIO, that each want to set up their own closures and that we want to verify modularly.
+The trust model of these set-up procedures goes bottom-up, i.e.
+any set-up procedure only trusts the set-up procedures that ran before it to ensure that its preconditions are satisfied when it assumes control.
+To allow for such a scenario, the execution of the bootcode has to be split up into different phases, where it holds that the postondition (precondition of the continuation) of boot program $i$ implies the precondition of boot program $i+1$.
+Additionally, the last bootloader has to make sure that the registers it delivers are safe with respect to $\VR^{\notMMIO}$, such that the ftlr (which will only hold for adversaries that have no access to \MMIO whatsoever) can be applied.
+
+In general, each set-up process $i$ will require the following as a precondition to safely execute.
+\begin{itemize}
+\item Safe register values, with respect to some argument World $W$ (whose regions are standard - this will mostly be used for proving the enter capabilities to e.g\ the driver secure, and I expect this world to be empty at the moment) and a set of \MMIO locations $\br{\MMIO_{i}}$, for which it holds that $\MMIO_i \subseteq \MMIO$.
+This allows the process to perform some $\MMIO$, but only to locations in $\MMIO \setminus \MMIO_i$.
+We could allow non-safe values in registers as well, but for simplicity's sake, we model the PC to access any non-safe values instead.
+Since the PC can point to unsafe values, we can currently set $\MMIO_i = \MMIO$ for simplicity's sake.
+In the more general case, more strict functional requirements $\br{\X{REG}_i^{\X{pre}}}$ on non-valid registers could be made.
+\item Its code and data stored somewhere accessible in memory (using regular points-to chunks), described by a set of constraints (containing points-to chunks) $\br{\X{MEM}_i^{\X{pre}}}$.
+Generally, also point-to chunks for memory that will be used in the future.
+This is the region $\br{R_{\X{adv},i}}$ Note that it is possible to write drivers that use relative addressing in their spec to not make their precondition unnecessarily strict.
+\item The $PC$ register points to (a subset of)  $R_{\X{adv},i}$.
+  The exact constraints are enforced in $\X{MEM}_i^{\X{pre}}$.
+  % currently represented as a contiguous range $\br{[b_{\X{adv},i},e_{\X{adv},i}[}$,
+  so that it is actually possible for our process to access this non-standard memory.
+  Optionally, part of the memory the $PC$ points to can be in $W$, but then the precondition needs to contain a proof that this part of the $PC$ is $RW$-valid, i.e.
+$\in \VR(\X{RW},\ldots)$, so that we can actually obtain the necessary points-to chunks from the region predicate.
+
+\item An invariant $\trInv$ on the trace, so that the process can perform IO itself or create more closures to do so.
+\end{itemize}
+
+Each set-up process will deliver the following postcondition (precondition of the continuation) to the next set-up process or to adversary code, starting from the precondition: (this also means that the proof of the next process has to go through, given this postcondition, i.e.
+it has to be sufficiently specific)
+\begin{itemize}
+\item Safe register values in any ($\forall$) private future world of $W$, with respect to the relation $\VR^{\notMMIO_{i+1}}$, where it has to hold that $\MMIO_{i} \subseteq \MMIO_{i+1}$ (which is currently trivially satisfied, since $\MMIO_i = \MMIO$) i.e.
+the permissions to perform IO are monotonely decreasing as set-up progresses.
+This constitutes no problem for instances of $\ER$ proven in earlier set-up programs, since the expression relation respects the following contravariant monotonicity property (since the value relation appears in the precondition of the observation relation): $S_1 \subseteq S_2 \Rightarrow \ER^{\overline{S_2}} \subseteq \ER^{\overline{S_1}}$.
+Note that for $\RR$ and $\VR$ the implication is covariant, so registers containing non-enter capabilities cannot trivially be ported to the next phase of boot.
+\item Some description of the layout of memory and concrete register values $\br{\X{REG}_i^{\X{post}}}$ and  $\br{\X{MEM}_i^{\X{post}}}$, that should be sufficiently detailed to be able to prove the precondition of the next set-up program.
+\item The trace invariant $\trInv$ (so that the next process can perform its own $\MMIO$).
+In the future, we can allow strengthening of $P$ throughout set-up by creating a monotone resource algebra for predicates, similarly to what we did for OCPL.
+We would then get $\knowInv{\iota}{\exists\; t, P \ldotp \tracefrag{t} \ast \ownGhost{\gamma_{\X{P}}}{\authfull{P}}\ast \pure{P(t)}}$ and could pass around a predicate $\ownGhost{\gamma_{\X{P}}}{\authfrag{P}}$ that we could strengthen before passing it to the next phase of set-up (we would need a notion of \emph{exact} and \emph{at least}, cfr.\ monotonic references).
+\end{itemize}
+
+In the case of the last set-up process $n$, we will also need the following, to be able to prove the semantic bootspec's postondition and in the end apply the ftlr, thereby proving the rest of execution safe:
+\begin{itemize}
+\item We need that $\MMIO_n = \MMIO$, such that we can apply the ftlr (otherwise some valid registers might grant the adverary access to \MMIO memory, and we cannot consider it an adversarial context in the robust safety sense)
+\item For similar reasons, the $PC$ should now point to valid memory only, i.e. the condition $\VR^{\notMMIO} (W)(\X{RW}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, a_{\X{adv}})$ should hold.
+\item The trace invariant $\trInv$ is no longer required, since the untrusted adversary code will not be able to perform \MMIO anyway.
+Leaving it there is not a technical problem either, however, given that the registers and memory will be valid with respect to $\VR^{\notMMIO}$.
+\end{itemize}
+
+Using the notation we introduced, we must prove the following for each set-up program (TODO: write an attempt at a more modular spec here):
+
+\[
 \begin{array}{l}
-       (\exists b_{adv}\; e_{adv}\; t.\; \\[0.5em]
-      \quad
+    \X{bootCont}_i(P,  b_{adv}, e_{adv}, a_{adv},W, \Phi) \eqdef  \\[0.2em]
     \begin{array}{l}
-        \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{adv}, e_{adv}, b_{adv}) \ast {} \\[0.7em]
-        \bigast{a_{i} \in A_{\X{entry}}}{r_{a_{i}} \mapsto_{r} (\X{E},g,b,e,a_i)} \ast
-        \bigast{r \in Reg\setminus \{\X{PC}, A_{\X{entry}}\}}{r \mapsto_{r} \X{inl}\; \_} \ast {} \\[1.4em]
-        \bigast{m \in [b_{adv}, e_{adv})}{m \mapsto_{a} \X{inl}\; \_} \ast
-        \bigast{l \in [b, e)}{m \mapsto_{a}\V{w}_l} \ast
-        % \driverC(b,e,\overline{\V{w}_l})\ast
-        \; \tracefrag{t} \ast \pure{P(t)}\\
-        \forall a_{i} \in A_{\X{entry}} \ldotp\,\left\{
-    {\begin{array}{ll}
-    \trInv \ast
-    \driverC(b,e,\overline{\V{w}_l})\!
-         \vdash \\
-         \forall W \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
-        \end{array}} \right. )
+      \forall  b'_{adv}, e'_{adv}, a'_{adv},W' \Wpriv W \ldotp\\
+       \quad\begin{array}{l}
+       (\X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b'_{adv}, e'_{adv}, a'_{adv}) \ast \br{\X{MEM}_i^{\X{post}}}(b'_{adv}, e'_{adv}, a'_{adv},W', b_{adv}, e_{adv}, a_{adv},W) \\[0.7em]
+        \bigast{r \in Reg\setminus \{\X{PC}\}}{(r \mapsto_{r} w_r \ast \VR^{\notMMIO} (W')(w_r))} \ast {} \\[1.2em]
+              \X{region}(W') \ast  \X{sts\_full\_world}\; \X{sts\_std} \; W'
+      \ast \; \trInv  )
       \end{array}
     \end{array} \\[4em]
-    \quad\wand
+    \qquad\wand
     \begin{array}{l}
@@
     \end{array}\\[1.2em]
-\end{array}
-\]
+
+  \X{bootSpec}_i(P) \eqdef \\[0.8em]
+    \quad
+  \begin{array}{l}
+    \forall  b_{adv}, e_{adv}, a_{adv},W,\Phi \ldotp\\
+    \left\{
+    \begin{array}{l}
+      \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{adv}, e_{adv}, a_{adv}) \ast \br{\X{MEM}_i^{\X{pre}}}(b_{adv}, e_{adv}, a_{adv},W) \\[0.7em]
+      \bigast{r \in Reg\setminus \{\X{PC}\}}{(r \mapsto_{r} w_r \ast \VR^{\notMMIO} (W)(w_r))} \ast {} \\[1.2em]
+      \X{region}(W) \ast  \X{sts\_full\_world}\; \X{sts\_std} \; W
+      \ast \; \trInv \; \ast \\[.8em]
+      \later \X{bootCont}_i(P,  b_{adv}, e_{adv}, a_{adv},W, \Phi)
+    \end{array}
+    \right\}
+    \\[2.5em]
+    \quad \I{Seq}\; (\I{Instr}\; \I{Executable})
+    \\[0.2em]
+    \left\{ \Phi \right\}
+  \end{array}\\[4em]
+
+\end{array}
+\]
+
+The overall $\X{bootSpec}$ is now proved by proving the individual $\X{bootSpec}_i$, and proving that they link together, i.e.
+$\br{\X{MEM}_{i+1}^{\X{pre}}}$ is implied by $\br{\X{MEM}_i^{\X{post}}} $, that the initial precondition of $\X{bootSpec}$ implies $\br{\X{MEM}_{0}^{\X{pre}}}$ and that the postcondition of $\X{bootSpec}$ is implied by $\br{\X{MEM}_n^{\X{post}}} $.
 
@@
 robust safety theorem slightly to incorporate these changes.
+Additionally, we probably want to ensure that the non-atomic invariants hold after each boot phase, so we have to add it to the postcondition of each boot spec.
 

### 2020-05-14 capability-notes/driver-notes/main.tex
Removed lingering 'sem'
Added few notes on how to implement drivers

@@
   \inferrule
-  { \X{bootSpec}^{\X{sem}}(P,r_0,m_0) \\
+  { \X{bootSpec}(P,r_0,m_0) \\
   (r_0, m_0, \emptyset) \longrightarrow^* (r, m, t)\\
@@
 stated above in some more detail.
+
+\subsection{Basics}
+
+\begin{itemize}
+\item Define types for traces
+\item Create a regular authoritative resource algebra over traces (we can refine this later on)
+\item Parameterize everything by the MMIO = Set(Loc)? locations?
+\item Formalize traces in the operational semantics $\rightarrow$ choose how we want to formalize MMIO with respect to memory
+\item Formalize traces in the state interpretation
+\end{itemize}
 

### 2020-05-27 notes/meeting-notes.tex

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}

\title{Capability Machine Meetings}

\begin{document}

\maketitle

\section{Week 6}

Status report of mechanisation:

\begin{itemize}
\item FTLR is done
\item Two examples are done (one that depends on LSE, another that depends on WBCF)
\end{itemize}

Current work on mechanisation (during week 6):

\begin{itemize}
\item Thomas: refactoring the FTLR Store case (one goal is to find out if there is a nicer way to prove each FTLR case, another is to consolidate certain equivalent definitions)
\item A\"{i}na: creating a spec for scall, and use that spec for much shorter proof of the examples. This involves also generalising the examples over some contiguous region of memory. 
\end{itemize}

Steps needed to mechanise uninitialised capabilities:

\begin{enumerate}
\item Update the operational semantics
\item Design the ghost state
\item Weakest precondition rules (either update existing ones or create new ones) 
\item Define the logical relation for the new U permission
\item FTLR (either update existing instructions or prove new cases) 
\item Make a new calling convention and prove that it satisfies the scall spec
\end{enumerate}

Alix will start out by updating the operational semantics. \\
During week 7, Thomas and A\"{i}na will determine the ghost state and the new LR definition. After that the weakest precondition/FTLR work (Alix/Thomas) can be done in parallel with the new calling convention (A\"{i}na)
\\[0.5em]
In week 8, we will discuss more concretely what refactoring is needed to get the work into a more presentable state. 
\\[1em]
Reading list: 
\begin{itemize}
\item OPCL paper (Arma\"{e}lle has offered to present the paper), Thomas can give a small presentation of the work he did in expanding the Iris implementation to handle effects (printing)
\item The high level benefits of low level sandboxing 
\item Benjamin Pierce ongoing work on property-based testing of a stack-protection micropolicy: 
\footnote{
\url{https://prosecco.gforge.inria.fr/personal/hritcu/talks/2017-12-18-Secure-Compilation-Infoiasi.pdf}} \footnote{\url{https://web.cs.wpi.edu/~rjwalls/nesd/slides/SullivanDraperNESDFall2016Dover.pdf}} \footnote{\url{https://semiwiki.com/ip/arm/7839-dover-microsystems-spins-new-approach-to-security/}}

\end{itemize}

SAIL: 
\begin{itemize}
\item Arma\"{e}lle will look into SAIL
\item One goal is to investigate how the current idealised machine can be extended to make it more realistic. To start out with this extension could be to add one of the following: 
	\begin{itemize}
		\item Virtual memory
    		\item Memory mapped I/O	
	\end{itemize}
\end{itemize}

In general, we should think about more security related properties (other than LSE and WBCF) that we can enforce with capabilities, and how to make the properties understandable for non PL people.

\end{document}

### 2020-06-29 Notes/uninitialized.tex
Merged in uninitialised_capabilities (pull request #3)
Merge uninitialized into master

* Squash commits

* Slight unfinished update of notes on uninitialized capabilities

* Test slack notifications

* Update notes on uninitialized cap

* Fix range of authority of an unintialized capablity in the notes

* Fix semantics of LoadU and StoreU, start wp rules for LoadU

* wp rules for LoadU

* wp rule for StoreU, admits stuff about LeastPermUpd and how that works with uninitialized capabilities

* Update definition of LeastPermUpd to take unintialized cap into accounts, finish wp rule for StoreU,

\documentclass[a4paper, 10pt]{article}
\topmargin-4.0cm

\usepackage[english]{babel}
\usepackage{csquotes}
\usepackage[a4paper, margin=2cm]{geometry}
\usepackage{fancyhdr}
\usepackage{color}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{url}
\usepackage{hyperref}
\usepackage{lastpage}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{calc}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}
\usepackage{tikz-cd}
\usepackage{interval}

\usepackage{newpxtext}
\usepackage{newpxmath}

\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\begin{document}
\lhead{}
\rhead{}
\renewcommand{\headrulewidth}{0pt} 
\renewcommand{\footrulewidth}{0pt}

\newcommand\code[1]{\textsf{#1}}

\begin{center}
  {\LARGE \bf Notes on uninitialized capabilities}\\
  {\small (to explore design possibilities and keep track of decisions)}
\end{center}

\section{Operational semantics}

We introduce new ``uninitialized'' counterparts to the standard RWLX, RWL, RWX and RW permissions as follows.

\begin{figure}[!h]
  \begin{center}
    \begin{tikzcd}[tips=false,column sep=1em,row sep=1.5em]
      & \text{RWLX} \ar[dashed,blue]{d} \ar{dl} \ar{dr} & & & \\
      \text{RWL} \ar[dashed,blue]{d} \ar{dr} & {\color{blue} \text{URWLX}} \ar[dashed,blue]{dl} \ar[dashed,blue]{dr} & \text{RWX} \ar{dl} \ar{dr} \ar[dashed,blue]{d} & &\\
           {\color{blue} \text{URWL}} \ar[dashed,blue]{dr} & \text{RW} \ar{dr} \ar[dashed,blue]{d} & {\color{blue} \text{URWX}} \ar[dashed,blue]{dl} & \text{RX} \ar{dl} \ar{d} & \\
           & {\color{blue} \text{URW}} \ar[dashed,blue]{dr} & \text{RO} \ar{d} & \text{E} \ar{dl} & \\
           & & \text{O} & & \\
    \end{tikzcd}
  \end{center}
  \caption{Permission hierarchy with new permissions in {\color{blue} blue}.}
\end{figure}

Uninitialized capabilities with authority over range $\interval[open right]{b}{e}$ and current address pointer $a$ have their counterpart's permission over range $\interval[open right]{b}{a}$ (actually $\interval[open right]{b}{\max(b, \min(a, e))}$ to be strictly correct), and need to write first to $a$ in order to increase the address pointer.

\subsection{Impact of current instructions}

We need to check the semantics of any instructions that can use capabilities in order to make sure that nothing strange can happen. The list of such instructions is the following: \code{Jmp}, \code{Jnz}, \code{Load}, \code{Store}, \code{Restrict}, \code{Subseg}, \code{Lea}, \code{IsPtr} and all \code{Get}.

We split the instructions in the following cases:
\begin{itemize}
  \itemsep0pt
\item It seems fairly clear that among those instructions, \code{IsPtr} and all \code{Get} instructions need not be changed.
\item For \code{Jmp} and \code{Jnz}, the current implementation is that the uninitialized capability would get copied into the PC register, and the machine would get stuck at the next step since the permission isn't one of RX, RWX or RWLX. This seems fine.
\item For \code{Load} and \code{Store}, we forbid the use of uninitialized capabilities with them and explain the reason in \ref{new}.
\item \code{Restrict} doesn't seem to need any change even with the new lattice.
\item For \code{Subseg}, the current implementation allows to go from a capability with range $\interval[open right]{b}{e}$ to $\interval[open right]{b'}{e'}$ if $\interval[open right]{b'}{e'} \subseteq \interval[open right]{b}{e}$. It assumes that the other instructions will have enough check to make sure that nothing bad happens even if the current address $a$ the capability points to is out of $\interval[open right]{b'}{e'}$.
\item For \code{Lea}, we need to make sure that we can only decrease the current address pointed to for uninitialized capabilities.
\end{itemize}

\subsection{\label{new}New instructions}
We cannot directly use uninitialized capabilities for reading (loading) in our setting as the read permission is informally given only on range $\interval[open right]{b}{a}$, and we do not have a way to provide an offset argument to the \code{Load} instruction lest we modify its semantics entirely (but possible).

We have thus implemented the following instruction: {\color{blue}\code{LoadU dst src offs}} checks that register \code{src} contains an uninitialized capability currently pointing to address $a$, and that $a' = a + \code{offs}$ is within $\interval[open right]{b}{a}$ (e.g., $b \leq a' < a$). Furthermore, we also need to check that $\interval[open right]{b}{a} \subseteq \interval[open right]{b}{e}$ (e.g., $b\leq a\leq e$). If the checks are passed, then the instruction loads the word contained at address $a'$ into register \code{dst}.

Similarly, we have implemented an instruction {\color{blue}\code{StoreU dst offs src}} which works in a similar fashion. It checks that \code{dst} contains an uninitialized capability with current address $a$, and that $a' = a + \code{offs}$ is such that $b \leq a' \leq a < e$. Furthermore, if $a' = a$ (e.g., $\code{offs} = 0$), the capability in \code{dst} is incremented. Obviously, regular checks concerning local words apply (e.g., if \code{src} contains a local word, the capability given by \code{dst} must have be allowed to write local words).

It would be nice to also modify the regular \code{Load} and \code{Store} operations to also take an offset argument if not only for symmetry's sake. Another design possibility is to remove the increment mechanism from \code{StoreU} and allows the regular \code{Store} instruction (without offset) to use uninitialized capabilities and increment them.

Finally, in order to be able to implement a calling convention, we also implemented a {\color{blue}\code{PromoteU dst}} instruction which semantics is to take a capability $(p, l, b, e, a)$ and transforms it into $(\hat{p}, l, b, \min{a}{e}, b)$ where $\hat{p}$ is X if $p = \text{UX}$ and is $p$ otherwise, it fails if $p = E$ though (similarly to \code{Restrict}, \code{Subseg} and \code{Lea}).

\begin{remark}
  We could get away with only implementing the \code{PromoteU} instruction and allows the regular \code{Store} instruction to use uninitialized capabilities and increment it. But this seems quite unpleasant to program with as one would need to write a bunch of data and promote the capability before having free read write permission on some chunk of memory.
\end{remark}

\subsection{Calling convention}

The calling convention should not change too much, except that the stack capability is now URWLX instead of RWLX. Consequently, in order to create the callback code on the stack, one needs write the code using the stack capability, copy it to another register, then use \code{PromoteU} to obtain a RWLX capability that can be restricted into an E capability.

\begin{remark}
  Allowing to directly restrict an URWLX or URWX capability to an E capability would be unsafe: one could use it to recover read-write permission on the uninitialized part of the capability.
\end{remark}

%% \section{Ghost state and resources}

%% Might be useful to have some notes on that whenever someones does it ?

\end{document}
